---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Background
```{r Backgroundsettings, echo = FALSE, warning=FALSE, message=FALSE}
source("./R/libraries.R"); source("./R/themes.R")
```

## Resistance-exercise prescription, a historical note and current challenges
Recommendations of systematic physical exercise with the purpose of improving health or physical performance has long been part of human culture, evident from records dating back to ancient Chinese, Indian and Greek civilizations
[@RN2640].<!-- Tipton Ancient China India and greek/roman written records on exercise precription -->
Today's exercise-training prescription still bear traces of ideas from these eras, further developed during the renaissance and formalized in systems like German Turnen and Ling gymnastics during the nineteenth century 
[@RN2663].<!-- Pfister 2003 describes differences and common ancestry of gymnastics and sports during 1700-1800 -->
German Turnen as a system of physical activities was established in a time when Germany developed from aristocracy to a unified nation.
Turnen not only served as a system of preparing men to fight for the developing nation but also as a way of establishing a national identity.
Ling gymnastics shared common origins with German Turnen and also served as a system of military preparation.
However, Ling also established systems for medical, pedagogical and aesthetic gymnastics.
Especially Ling's medical gymnastics was important for the development of modern exercise prescription as it was scientifically oriented, based on the physiological and medical understanding of that time [@RN2663].
The medical gymnastics of the nineteenth century is referenced in twentieth century texts on therapeutic exercise prescription
[@RN2634].<!-- Nicoll 1943 Dosage of exercise therapy "never exercised against more than 25% of their all out efficiency" with increased exercise periods and decreased rest periods as progression -->

With the introduction of "heavy resistance exercises" for development of muscle strength and mass after injury, DeLorme outlined a system on which modern resistance-exercise prescription is based [@RN2633].<!-- DeLorme 1945 case reports of systematic resistance training for restoration of muscle power and mass. First mention of RM -->
DeLorme published his system short after the Second World War [@RN2633]
during which he, as a newly graduated physician, had been working with war injury rehabilitation
[@RN2639]<!-- Todd 2012 DeLorme biography -->.
Inspired by practitioners of weight training [@RN2639],<!-- Todd 2012 DeLorme biography --> 
DeLorme specifically emphasized high-resistance, low-repetition exercises where progression was achieved with increased resistance [@RN2633] as opposed to previous recommendations of endurance-like exercise where progression was achieved through increased number of repetitions
[@RN2634].<!-- Nicoll 1943 -->
DeLorme originally used the term "heavy resistance exercises" in contrasts to low-resistance exercises [@RN2633], but as this could be perceived as exercises performed only with heavy weights, the system was renamed *progressive resistance exercise* to better reflect the method
[@RN2641]
^[In the current text, exercise is defined as an acute bout of physical activity designed to affect physical characteristics such as strength, speed or endurance. Training is defined as the systematic process of combining multiple exercise-sessions performed in sequence over time. DeLorme first used the adjective *heavy* to describe the resistance prescribed to overcome during exercises but later changed this adjective to *progressive*. In modern texts, the adjective is commonly omitted from the description and *resistance exercise/training* is used to describe strength-promoting exercises and training regimes requiring the neuromuscular system to exert force against resistance. Omitting the adjective has led to many heated debates among exercise physiologists as "endurance exercises are also performed against a[n] (external) resistance". With no ambition to resolve any conflict in the area, *resistance exercise/training* will be used synonymous with *progressive* or *heavy* *resistance exercise/training*.]
.<!-- DeLorme and Watkins 1948 -->
Indeed, central to the system was the concept of repetition maximum [@RN2633].
Repetition maximum refers to the external resistance that can be overcome with a given number of repetitions.
By adjusting external resistance to each individuals progression over the course of a training program, exercises are both individualized and progress is monitored [@RN2633].
DeLorme originally prescribed sessions of up to 100 repetitions performed in sets of 10 repetitions [@RN2633] but later revised this recommendation to three sets of 10 repetitions performed with increasing intensities
[@RN2641]. <!-- DeLorme 1948 Technics of progressive resistance exercise -->

Scientific inquiries into prescription of resistance training from the first part of the twentieth century concerned its therapeutic use
[e.g. @RN2633;<!-- DeLorme case report -->
@RN2646]<!-- DeLorme controlled trial of RT vs. traditional conservative (low-resistance) treatment -->
but was also evaluated in the context of improving strength and physical performance in healthy populations
[e.g. @RN2632;<!-- 1946 Houtz 16 untrained females increased strength over 4 weeks follwing DeLormes system -->
@RN2644; <!-- Chui 1950 Effect of strength training vs. control on physical performance in college stud -->
@RN2642]. <!-- Capen 1950 Effect of strength training in college stud.   -->
Scientific contributions soon moved from questions regarding the effectiveness of resistance training *per se* to comparing outcomes from different modes of resistance training 
[@RN2645;<!-- Hettinger 1953 comparing different loading patterns for isometric RT--> 
@RN1477;<!-- Capen 1956 Four different variations on the DeLorme system -->
@RN1472;<!-- Berger 1962 effect of varied weight training program on strength -->
@RN2659;<!-- Berger 1965 Increased %of 1RM increases strength gains -->
@RN2658;<!-- Berger 1967 Supra-maximal efforts (max of each rep) more efficient than 10RM loads   -->
@RN2656].<!-- O'Shea 1966 3 sets of 10, 5 or 3 reps on muscle strength and girth squat -->
A vocabulary for progressive resistance exercise-training developed through these investigations and parallel practice,
the introduction of repetition maximum by DeLorme being one example. These concepts established as modern definitions of exercise variables enabling precise prescription of training loads for a variety of populations and training goals
[@RN2537<!-- Fleck & Kraemer 2014 Textbook ch 5 : based on a needs analysis acute variables are selected   -->
@RN798].<!-- Bird 2005: Acute training variables leads to different training outcomes -->

Although this development started after the Second World War, resistance training was not part of general exercise guidelines until much later.
The American College of Sports Medicine (ACSM) position statement on exercise for healthy individuals from 1978 primarily concerned physical fitness in terms of cardio-respiratory fitness
[@RN2655]<!-- ACSM 1978 guidelines -->.
Since the updated 1990 ACSM statement, resistance training is recommended to be included as part of a sensible, general training program
[@RN2654].<!-- ACSM 1990 guidelines --> 
The introduction of resistance-training as part of the ACSM recommendation also coincides with specific recommendations on resistance training being part of other consensus statements
[@RN2666, Ch. 2].<!-- Physical Activity and Health: A Report of the Surgeon General provides an overview of recommendation guidelines evolution  -->
Consequently, informed by epidemiological data, the most recent general guidelines for physical activity do include resistance training [@RN2667].<!-- WHO guidelines -->

It could be argued that the above account reflects the fact that common understandings of *why* and *how to* exercise are influenced by societal norms and historic events such as the search of national identity in the nineteenth century or war injuries in the twentieth century
[@RN2663;
@RN2639].
In attempting to outline contemporary influences on exercise prescription, one could argue that the development of techniques to collect large amount of biological data is one such influence. The continuously decreasing cost of a sequenced human genome
[@RN2696] can serve as an example of this development.<!-- Wetterstrand webpage genome sequence costs -->
Such molecular techniques has enabled efforts to describe mechanisms by which exercise training induce favorable adaptations. The newly established Molecular Transducers of Physical Activity Consortium is an example of a large scale effort, explicitly initiated to develop personalized exercise recommendations and identify molecular targets through which effects of exercised may be mimicked
[@RN2678].<!-- MToPAC explicitly stating diagnostics is a goal --> 
Advances in bio-medical technologies are enablers of this enterprise and the quest to *individualize* exercise based on molecular diagnostics can be seen as a motivation for modern exercise science
[@RN2678;<!-- MToPAC explicitly stating diagnostics is a goal -->
@RN2677].<!-- Sparks 2017, exercise response heterogenity, goal to test appropriate exercise dosa based on e.g. DNA -->
Scientific research into contemporary exercise prescription can thus be understood as a part of the era of *personalized medicine*. 

A challenge facing this program is to accurately describe etiologies of response heterogeneity associated with physical training. A wide variation of individual responses are commonly observed after standardized resistance-training programs where measures of muscle strength-changes varies from -32 to +250\% and muscle size-changes varies from negative to (-11\%) to impressively large (+59\%)
[@RN764;<!-- Hubal wide range of responses to standardized exercise  -->
@RN1741].<!-- Ahtiainen 2016 7% (strength) and 29% (muscle size) categorized as non responders (accounting for controls)  -->
By relating such variations to individuals genome (DNA)
[@RN1263],<!-- Pescatello review of the famuss data set -->
and messenger RNA (mRNA) profiles
[@RN826;<!-- Thalacker-Mercer, Cluster analysis, non-responders pro-inflammatory mRNA profile at baseline -->
@RN2684]<!-- Stokes co-relation between mRNA profiles and LMM changes  -->
we are beginning to gain knowledge about the genetic influence on training responses.
In such studies, a common strategy has been to dichotomize responses into "responders" and "non-responders" to exercise training. From a public health perspective this is probably fruitful when *non-response* is defined as the absence of meaningful health-related adaptations, or even adverse effects in response to a given training regime
[@RN2677;<!-- Sparks 2017 response heterogeneity -->
@RN2698].<!-- Roth viewpoint 3 scenarios for genomic screening -->
The existence of non-responders would have large implications regarding exercise prescription on the population scale
[@RN758],<!-- Timmons 2011  -->
and for any given individual as it, if properly diagnosed, would guide clinical decision-making.

<!-- Although "molecular exercise prescription" as a mean of eliminating adverse responses to a given training regime shows potential for future applications
[@RN2698; <!-- Roth viewpoint 3 scenarios for genomic screening 
@RN2547],
the rise of data intensive techniques has also highlighted its complexity.

In simplistic terms, genetic information is transferred from DNA to RNA through transcription and from RNA to protein through translation.
This flow of information, indicates a deterministic role of DNA in controlling the characteristics of an organism. It is however apparent that the transcription of specific genes are under tight control of several mechanisms including a group of non-genetic modifications to the DNA organization of each cell, known as epigenetic modifications
[@RN2690;<!-- Seaborn sharples review 2020 exercise metabolism as a determinant for epigenetic changes 
@RN2691].<!-- McGee Hargreaves 2019 review, epigentics/metabolism/exercise and mention intergenerational trandmission thorugh epigentic modification 
Such control determines gene expression in response to progressive resistance training with modifications persisting after a period of de-training
[@RN2683;<!-- Transcriptome is associated with epigentic changes in response to RE 
@RN2012].<!-- Seaborn 2018 epigenetic changes persist after de-training, epigenetic memory of training 
Regulation of gene expression is thus related to modifiable variables such as exercise and diet
[@RN2690;<!-- Seaborn sharples review 2020 exercise metabolism as a determinant for epigenetic changes 
@RN2691;<!-- McGee Hargreaves 2019 review, epigentics/metabolism/exercise and mention intergenerational trandmission thorugh epigentic modification 
@RN2694;<!-- Yoshie 2020 13 mo old mice have epigenetic changes in skm due to childhood diet 
@RN2695]<!-- Godfrey 2016 review childhood environmental conditions affects epigenetics 
linking environmental factors to the complexity of gene expression.
-->
A key aspect of successful exercise diagnostics would be to take advantage of the relationship between exercise variables (i.e. modality, intensity, volume etc.) and exercise response for a given individual.
By adapting an individuals training program based on some prior knowledge about the individual, it is possible that the response could be positively affected. 
Observations supporting such notion exists as an individual classified as non-responsive to a specific exercise modality (e.g. endurance training) may be classified as a responder to another (e.g. resistance training)
[@RN2681].<!-- Changing from end to res training removed non responsiveness --> 
Even changing training variables within a specific modality have been shown to convert non-responders to responders as endurance training volume was increased
[@RN2699].<!-- Lundby increasing training volume improved non-responders -->

Although apparent reversal of non-response to exercise training has been observed by manipulating training variables, decisive indications for such manipulations are still lacking.

## Adaptations to resistance training

### Muscle hypertrophy and strength
Systematic resistance training typically increases muscle mass and strength, adaptations through which many beneficial effects on health (e.g. increased amino acid storage physical independence) and athletic performance are conveyed. 
Muscle growth is a well characterized response to resistance training.
On the whole muscle level, healthy untrained individuals can be expected to increases their muscle mass by $\sim$ 5-20\% when training is conducted over two weeks to 6 months
[@RN1741; <!-- Ahtianen -->
@RN346;<!-- Wernbom  -->
@RN764].<!-- Hubal 2005 elbow flexor 18.9% increase -->
Over this time span, muscle growth is approximately linear with time
[@RN1596;<!-- Defreitas 2011 approx Linear increase in muscle mass and strength in response to 8 wk str training -->
@RN2113;<!-- Stock 2017 Concentric training only for arms muscle thickness and lean mass gains after 6 and 7 sessions -->
@RN2736]<!-- Narici 1989 Increase in CSA (MRI, linearly ~ 0.14%/day total 8.5%) and MVC (total 20.8%) no change in untrained control in CSA, numerical increases occured in untrained in EMG/MVC   -->
<!-- Although some of the early increase muscle mass can be attributed muscle damage-induced edema
[@RN2746],Damas 2016 increased echo intensity indicating muscle edema at 3 weeks -->
and has been detected as early as 3-4 weeks after training initiation, without apparent muscle edema 
[@RN1596;<!-- Defreitas 2011 approx Linear increase in muscle mass and strength in response to 8 wk str training -->
@RN2113].<!-- Stock 2017 Concentric training only for arms muscle thickness and lean mass gains after 6 and 7 sessions -->

Relative muscle growth can be expected to be more pronounced in upper- compared to lower-body muscles when loading patterns are similar
[@RN346;<!-- Wernbom  -->
@RN2739].<!-- Welle 1996 larger incr in csa in upper body -->
This possibly relates to greater every day activity levels of lower-body muscles requiring a larger stimuli for adaptation
[@RN767].<!-- Folland review -->
Small, but detectable differences in muscle growth is typically seen between sexes for training induced muscle growth in the upper-body
[@RN764]<!-- Hubal CSA men: 20.4% vs 17.9% -->
but not for lower-body muscles
[@RN2740].<!-- Roberts 2020, meta analysis when all muscles combined no diff between sexes in within sex effects sizes -->
Furthermore can hypertrophic responses be expected to be reduced with increasing age 
[@RN2739; <!-- Welle 1996 larger incr in csa in upper body --> 
@RN752]<!-- Peterson 2011 Meta analysis, age (negatively) and training volume (positively) affects rt outcomes  -->
but increased with sufficient addition of dietary protein
[@RN2199].<!-- Morton meta Dietary protein supplementation significantly enhanced changes in muscle strength and size during prolonged RET in healthy adults. Increasing age reduces and training experience increases the efficacy of protein supplementation during RET. -->
Additionally, training variables such as intensity, volume, frequency (reviewed below) together with other training aids, e.g. manipulation of blood flow through pressure cuffs
[@RN2745]<!-- Lixandrao BFR meta analysis hypertrophy similar between low load BFR and high load -->
can effectively modulate resistance-training induced hypertrophy.
Together, this underlines that both non-modifiable (e.g. sexual dimorphism and age) and modifiable factors (e.g. training variables and protein supplementation) affects training-induced muscle hypertrophy.

Whole muscles growth in response to short term resistance training occurs primarily through growth of individual muscle fibers (muscle cells).
This can be assumed as training-induced splitting of existing fibers and/or myogenesis, formation of new muscle cells, are likely slow processes and any increase in the number of fibers by such mechanisms would only represent a small addition to the whole muscle mass 
[@RN2741;<!-- Murach 2019 review on hyperplasia  -->
@RN2742; <!-- Sjöström 1991 Increase in left/right tibialis anterior -->
@RN2754;<!-- MacDougall no diff between body builders and untrained in number of fibers, equal relative amounts of collagen -->
@RN767].<!-- Folland 2007 review on mechanism for hypertrophy and strength gains -->
Growth of muscle fibers transfers to greater muscle strength trough an expansion of the fibers contractile elements.
The muscle cell is to a large degree occupied by myofibrils
(about 80% of the cell volume 
[@RN2731]<!-- Lüthi 1986 Myofibrils occupie more space in the mucsle fiber after resistance training mitochondria occupie less. No sign increase in fiber CSA with strength training no control group -->),
which in turn contain sarcomeres, arranged in series.
Upon neural activation and subsequent Ca\textsuperscript{2+} release, the sarcomere shortens through interaction between actin and myosin resulting in force generation
[@RN2756].<!-- Sweenney review on muscle contraction -->
With resistance training, the number of parallel myofibrils increases with the growth of individual fibers
[@RN2669]<!-- Straight 2020 muscle fiber size increases with RT, in midlle age old, age inversely related to training effects -->
leading to a greater force generating capacity of the whole muscle
[@RN2731].<!-- Lüthi 1986 Myofibrils occupie more space in the mucsle fiber after resistance training mitochondria occupie less. No sign increase in fiber CSA with strength training no control group -->

Measures of whole muscle size corresponds well with strength, particularly when a measure of muscle size reflects the muscles cross-sectional area
[@RN2758;<!-- Bamman 2000 Anatomical and estimates physiological cross section better prediction for strength on whole muscle level in vivo in humans -->
@RN1142].<!-- Erskin 2013 muscle volume and CSA to strength r = 0.812 and 0.806  -->
However, in response to resistance training, increases in strength are typically greater in magnitude than muscle growth
[@RN2629;<!-- Ikai 1970 Arm flexor training within-subject protocol, trained arm 23% increase in CSA and 90 % strength 100 days of 6 days/week training. Control arm 30% str incr, 5% CSA incr -->
@RN2737;<!-- Young 1983 diproportional higher strength gains than CSA, increase in CSA with ultrasound in response to RT -->
@RN2736;<!-- Narici 1989 Increase in CSA (MRI, linearly ~ 0.14%/day total 8.5%) and MVC (total 20.8%) no change in untrained control in CSA, numerical increases occured in untrained in EMG/MVC   -->
@RN2735].<!-- Narici 1996 Local changes in whole muscle CSA after 6 months of RT, no stat sig increase in fiber CSA at a site with low MR based increase in CSA -->
When relationships between resistance-training induced change in muscle size and strength in previously untrained individuals are analyzed, only a portion of the variation in muscle strength can be accounted for by changes in muscle size ($\sim$ 2.5-28\%)
[@RN1142;<!-- Erskine 2013 23% and 28% muscle volume and mvf/1rm  -->
@RN1741;<!-- Ahtiainen r = 0.157, r2 = 0.025 -->
@RN2158]<!-- Vigotsky 2018 within subject model 12-18% accounted for, smaller when between subjects -->
depending on type of measurements and statistical model used
[@RN2158]<!-- Vigotsky 2018 within subject model 12-18% accounted for, smaller when between subjects -->
This underlines that muscle hypertrophy is one contributing factor to muscle strength gains, but not the only one.

Different experimental models have shown that strength can increase without concomitant muscle hypertrophy.
Getting acquainted to the actual strength test through repeated training of maximal performance produces similar gains in strength without concomitant hypertrophy 
[@RN2760].<!-- Mattocks 2017 Practicing 1RM leads to similar strength increases as RT  -->
If resistance training is performed unilaterally and the contralateral limb act as a control, strength gains are typically also seen here
[@RN2219;<!-- Carroll 2006 contralateral effects review -->
@RN2766].<!-- Green 2018 training has contralateral effects both for strength and skill -->
Additionally, systematic imagery training, without muscle activation produces greater strength gains than control and low-intensity training conditions 
[@RN2767].<!-- Plantar flexion imaginary more effective than control in strength gains -->
Together these observations indirectly points to the central nervous system and motor learning as important factors for strength gains.
In addition to effects that mainly can be attributed to motor learning, resistance training leads to changed behavior of motor units, estimated from surface electromyograms 
[@RN2763].<!-- Del Vechchio 2019 rt changes fiering rates in motor units   -->
Such changes could be attributed to morphological and functional changes of motoroneurons
[@RN2764].<!-- Gardiner aplha motorneuron changes with RT -->


### Changes in muscle fiber contractile and metabolic characteristics with resistance training 
In adult human skeletal muscles, muscle fiber types can be identified based on their myosin heavy chain isoform composition. Pure fibers express a single mysoin heavy chain isoform whereas hybrid fibers co-express isoforms.
Primarily three myosin heavy chain protein isoforms are expressed in adult human skeletal muscle, determined transcriptionally through expression of the genes *MYH7*, *MYH2* and *MYH1* corresponding to myosin heavy chain I, IIA and IIX
[@RN819]<!-- -->

Single fibers of different fiber types have specific contractile properties regardless of muscular origin
[@RN846],<!-- Harridge 1996 specific contraction characteristics in different fiber types determined by myosin heavy chain-->
with type II fibers displaying greater force-generating capacity and shortening velocities than type I fibers when normalized to fiber cross sectional area
[@RN2169;<!-- Widrick IIx > IIa > I in force/power unloaded shortening, training alters these charecteristics in proportion to hypertrophy -->
@RN846]<!-- Harridge 1996 specific contraction characteristics in different fiber types determined by myosin heavy chain-->
These differences directly relates to the myosin heavy chain proteins displaying different characteristics when interacting with actin
[@RN849].<!-- Larsson diffrenet motility speeds of different myosin isoforms -->
Importantly, *in vitro* assays performed at physiological temperatures shows that myosin heavy chain isoforms extracted from type II fibers are two-fold faster compared to type I fibers with no difference between type IIX and IIA
[@RN849].<!-- Larsson diffrenet motility speeds of different myosin isoforms -->

In addition to contractile characteristics, fiber types identified on the basis of their myosin heavy chain content also differs in metabolic profiles.
Type I fibers are characterized as having lower glycolytic but higher oxidative potential compared to type II fibers
@RN1885].<!-- Essen 1975 Type II fibers lower oxidative potential compared to type I assumed from activity of oxidative enzymes -->
Differences in metabolic profiles relates to fatigue resistance were type I fibers can maintain power output as well as ATP levels during intense exercise but type II and especially type IIX fibers fail to do so 
[@RN819;<!-- -->
@RN2801].<!-- Karatzaferi 2001 Drop in ATP most pronounced in type 2x fibers in response to acute exercise -->

Fiber type characteristics effectively modulate the muscle's ability to perform specific activities. Differences in fiber type composition within individuals, between different muscles, reflects this as anti-gravity muscle of the lower body typically express more type I fibers compared to upper-body muscles
[@RN819;<!-- -->
@RN846].<!-- Harridge 1996 specific contraction characteristics in different fiber types determined by myosin heavy chain-->
Differences in fiber type composition between individuals and sexes are to some degree genetically determined
[@RN2798;<!-- Simoneau 1995 40% variation in type I fibers can be attributed to genetic factors, 15% to sampling error and 45 to environment -->
@RN2795;<!-- Differences between men and women more type i in women -->
@RN285].<!-- Fiber types in yopung men and women, high percentage of 2X (b) fibers ~ 20%-->
Resistance training on the other hand is among non-genetic factors influencing fiber-type composition. Short term resistance training, designed for muscle hypertrophy and strength gains specifically converts type IIX fibers to more fatigue resistant type IIA fibers with unaltered type I fiber proportions
[@RN2220;<!-- Adams Down regulation of MyHc 2x content in training group both myhc isoform sds page and atpase staining -->
@RN2799;<!-- Williamsson less IIx dominant hybrid fibers with training conversion towards IIa -->
@RN2056;<!-- Andersen 2018 type IIx to 2a most prominent switch-->
@RN1489].<!-- Ellefsen 2014, training induced switch from 2A to 2X -->
Such conversion is apparent both when measured on the protein and mRNA level
[@RN1489;<!-- Ellefsen  -->
@RN2056].<!-- 2x gene swithoff Anderssen --> 
In contrast, reduced activity or inactivity readily increases the proportion of type IIX expressing fibers 
@RN2057<!-- Andersen 2x overshoot after de training -->
@RN2108].<!-- Bed rest increases type IIx -->

Concomitant with resistance training-induced muscle hypertrophy, and fiber type switch, the relative mitochondrial density of the muscle decrease as myofibrilar protein fractions increases, evident from electron microscopy
[@RN2731].<!-- Lüthi  -->
However, a single session of resistance training, albeit with low resistance (30\% of 1RM), was shown to increase mitochondrial, as well as myofibrilar and sarcoplasmic protein synthesis rates. When exercise was performed with slower movement speeds (longer time under tension), the increase in mitochondrial protein synthesis was shown to be greater
[@RN1505].<!-- Burd 2012 A single session of resistance exercise (30%1RM) increases mitochondrial protein synthesis, more so with higher metabolic stress (6s repetitions compared to 1s reps). -->
This indicates that the magnitude of metabolic stress induced by resistance training affects the subsequent mitochondrial remodeling.
Such remodeling could explain improved mitochondrial function, measured as mitochondrial respiration in response to 12-weeks of resistance training with less pronounced changes seen in mitochondrial proteins
[@RN2608].<!-- Higher mitochondrial respiration after 12 wk RT --> 
An improved mitochondrial efficiency could also be linked to fiber type transitions.
Mitochondria have the ability to form dynamic networks within cells by fusion (and fission) of individual mitochondria, a characteristic important for normal function
[@RN2817].
Such behavior have been shown to be fiber type specific in muscle, with oxidative fibers (type I and IIA) compared to glycolytic fibers (type IIX and IIB in mice) displaying greater, elongated mitochondrial networks
[@RN2816].<!-- Mishra 2015 elongated   -->
In response to endurance training, fiber type switch from glycolytic to oxidative, coincided with switch to less fragmented mitochondria
[@RN2816].
Such coordinated remodeling can be linked to common molecular mechanisms regulating both fiber type and mitochondrial biogenesis
[@RN2818;<!-- Lin 2002 PGC1a expression drives type II to type I characteristics-->
@RN2819].<!-- PGC1a knock out shift from type I to type IIx and reduced endurance capacity -->


### Connective tissue
In addition to adaptive changes to the contracting apparatus of single muscle fibers and neural mechanisms regulating their activity, resistance training modulates bone, tendon and connective tissue.
From a general perspective, tissues enabling e.g. locomotion by conveying forces produced by contracting muscles and stabilizing body segments adapts in an activity-specific manner
[@RN2775;<!-- greendale 2003 activity patterns affect bone mineral density, sporting activities more effective than active living home activities -->
@RN2771]<!-- Magnusson 2008 review tendon CSA increase with endurance training -->
Specifically, short term resistance training leads to changes in mechanical properties of bone without increases in bone mineral content or density suggesting qualitative changes
[@RN2774]<!-- Miller 2009 rt increases tibial bending stiffness without changes in bobne mineral content 20 weeks, 3 sessions/week con and ecc, females 18-26 years-->
Similarly, tendons respond to short term resistance training by increasing stiffness at high mechanical stress-levels without increased cross sectional area
[@RN2772].<!-- Massey 2017 short term (12 weeks) rt increase patellar tendon stiffness at high stress to a similar degree as long (4 years) compared to controls. Long term training led to increase stiffness at low stress . No increase in tendon CSA in any training group -->
Interestingly, these adaptations seem to reach a plateau as no additional change in this characteristic is seen in individuals who have exercised over four years. However increased stiffness at lower mechanical stress levels have been documented
@RN2771].<!-- Magnusson 2008 review tendon CSA increase with endurance training -->
Changes in tendon properties in response to resistance training may thus primarily be associated with qualitative changes
[@RN2771]<!-- Magnusson 2008 review tendon CSA increase with endurance training -->
as consequence of exercise and training-induced increased turnover of collagen in tendons, indicating remodelling
[@RN2771;<!-- Magnusson review -->
@RN2770].<!-- Kjaer review, focus on collagen metabolism in  -->

Muscle fibers are embedded in connective tissue, observed as surrounding the whole muscle (epimysium), muscle fasicles (perimysium) and muscle fibers (endomysium)
[@RN2783].<!-- Borg Rat skeletal muscle electron micropscopy  -->
Connective tissue structures constituting the endomysium connects muscle fibers to adjacent fibers, capillaries and nerves which together with higher order structures make up the extracellular matrix enabeling mechanical and biochemical interaction between cell types
[@RN2783;<!-- Rat skeletal muscle  -->
@RN2421].<!-- Kjaer review -->
Together with the myotendinous junction, intramuscular structures (primarily perimysium) transmits forces originating from contracting muscle fibers to tendon and bone and act as an elastic energy storage during e.g. locomotion
[@RN2421].<!-- Kjaer review -->
The mechanical properties of the extracellular matrix also allows for mechanical stimuli to be converted to biochemical signaling initiating e.g. responses to exercise.

There is a general coordination between connective tissue and muscle-cell remodeling in response to loading, evident coordinated responses from different cell types in response to exercise
[@RN2421].<!-- Kjaer review -->
The major constituent of the extracellular matrix is collagen, produced in fibroblasts.
In response to acute endurance-type exercise, collagen synthesis and muscle cell specific protein synthesis (myofibrillar and sarcoplasmic fractions) rise in a coordinated fashion
[@RN2788].<!-- Miller coordinated prot synthesis response to exercise (not RT) -->
Also, in response to short- and long-term resistance training with subsequent muscle hypertrophy, relative collagen content of muscle tissue remains stable
[@RN2754].<!-- Macdougall Similal relative collagen content and non contracting tissue in untrained and trained human skm  -->
However, fine tuning of such coordination could exist as contraction mode has shown to differentially affect myofibrillar protein but not collagen synthesis after acute exercise
[@RN2454].<!-- Holm 2010 Contraction mode.. No diff between fasted/fed nor heavy/light load -->

Increased remodeling of components of the extracellular matrix is a typical response to resistance training evident from gene expression studies
[@RN2298;<!-- Damas long term training ECM up gene -->
@RN774],<!-- Raue, upregulation of ECM after long term exercise -->
and studies of acute protein synthesis
[@RN2457;<!-- Moore 2005 Collagen synthesis up irrespective of ecc or conc 6 * 10 maximal contractions -->
@RN2454].<!-- Holm 2010 Contraction mode.. No diff between fasted/fed nor heavy/light load -->
Such remodeling may contribute to increases in specific force (force generated per muscle cross section) seen after resistance training through improved lateral force transfer
[@RN1504].<!-- Erskine 2011 Hypothesized increased lateral force transfer increase specific tension in vivo in response to strength training in humans as single fiber specific tension nor mhc composition could explain -->


## Effects of exercise prescription on muscle mass and strength
Precise exercise-training prescription gives information on exercises, their sequential order, intensity and volume, rest periods between efforts or sessions and the frequency at which exercise sessions are to be performed
[@RN789].<!-- Bird designing resistance training programs -->
By manipulating these variables, resistance training programs can be tailored to better fit goals and starting points of any individual. 
The relative importance of exercise-training variables for training outcomes has been examined in numerous studies including (but not limited to) the overall organization of exercise sessions,
[@RN2575;<!-- Evans mini review on periodization --> 
@RN2572]<!-- Grgic 2017 No effect of linear and daily unduluting periodization on muscle hypertrophy -->
training frequency 
[@RN2571]<!-- Schoenfeld 2016 training frequency twice per week better in volume equated studies for hypertrophy -->,
<!-- Higher frequencies gives higher gains in strength but not when volume equated -->
and intensity
[@RN2569].<!-- Schoenfeld, no difference between high vs low load strength training in hypertrophy but 1rm favors high loads -->
It could be argued that training volume is of particular importance for muscle growth as when this variable is held constant, manipulation of other variables has little or no effect hypertrophy
[@RN1612;<!-- Schoenfeld 2014 7*3Rm vs 3*10RM equal muscle hypertrophy but dissimilar strength -->
@RN2569].<!-- Meta analysis high vs low resistance has little effect on muscle mass gains but strength is affected positively with higher intensities. --> 
For development of strength, factors such as intensity and within session organization of exercises is of importance 
[@RN2570;<!-- Frequency has little effect on strength when volume is equated -->
@RN2591],<!-- Exercise order influences strength but not hypertrophy -->
however, when other factors are held constant, increased training volume generally leads to increased strength
[@RN2570;
@RN2492;<!-- Ralston 2017 meta analysis medium to high volume more effective strength gains -->
@RN793],
similarly to effects of training volume on muscle growth
[@RN1767;
@RN789].

### Effects of resistance exercise volume on muscle strength and mass
Exercise volume can be prescribed as the within session number of sets performed per muscle group. This unit is practical as it comparable between individuals and muscle groups [@RN2130].<!-- Baz-Valle 2018 sets per exercise better than mechanical load times rep or total work -->
Berger conducted an early study concerning effects of resistance exercise volume with the goal to determine what method most efficiently produced strength gains (in healthy young males) [@RN1476]. Berger compared one, two and three sets performed with two, six or ten repetition maximum (RM) in the bench press, three times per week, over twelve weeks. As the combined effect of three sets per session was superior regardless of the number of repetitions performed Berger concluded in favor of three sets. This conclusion was later challenged on the basis of data interpretation
[@RN794;<!-- Carpenelli and Otto questions the analysis-->
@RN2538].<!-- Feigenbaum -->
Reveiwing the study by Berger and others, Carpinelli and Otto arrived to the conclusion that there was "insufficient evidence to support the prevalent belief that a greater volume of exercise (through multiple sets) will elicit superior muscular strength or hypertrophy" [@RN794]. This stand has since been repeatedly put forward as a criticism of higher volume training programs 
[@RN2568; 
@RN2201] and sparked considerable scientific activity. The main argument against the recommendation of additional volume in strength training programs has been the lack of statistically significant results in single studies [@RN2568; @RN2538]. 
Indeed, individual studies do not generally agree on dose-dependent effects of training volume on muscle mass and strength gains
[@RN2465; <!-- Ribiero 2015 Strength but not body composition change with dose dep older women -->
@RN2464;<!-- Correa 2015 older women, increase muscle thickness and strength with volume  -->
@RN2463;<!-- Bottaro 2011 untrained young male similar results in muscle thickness and strength -->
@RN1570; <!-- Radaelli 2014 1 vs 3vs 5 sets, 5 sets most effective in muscle thickness and some strength exercise -->
@RN1518;<!-- Radaelli 2014 older women no diff between volume groups in muscle thickness or strength gains -->
@RN1474;<!-- McBride 1474 Men and women DXA no diff but strength increase more in multiple sets -->
@RN1472; <!-- Galvao 2005 Older adults, no diff in body comp DXA, but increase in strength  -->
@RN1456;<!-- Starkey 1996 young male and female no diff in strength or muscle thickenss gains -->
@RN1454;<!-- Ostrowski 1997 trained young med, low volume decrease from previous levels no difference between groups -->
@RN1384; <!-- Rhea 2002 young men bod pod no diff in body comp changes but diff in strength gains in favour of multi set -->
@RN1382; <!-- Cannon 2010, Older women and younger women, muscle volume and strength increased similarly irrespective of age or training volume and without interaction -->
@RN776],<!-- Rønnestad, three set more effectiv in increaseing strength and muscle mass in lower body compared to upper body -->
including studies performed within participants, where different training volumes are allocated to either extremity 
[@RN834;<!-- Mitchell 2012 young men no sig difference in strength or mass gains between 1 and 3 sets knee extension -->
@RN1607].<!-- Sooneste 2013 young men p=0.076 on strength gains in favor of three sets, CSA increased more in 3 sets leg exercise -->
For example, differences in strength are between volume conditions are found in older individuals 
[@RN2465; <!-- Ribiero 2015 Strength but not body composition change with dose dep older women -->
@RN2464;<!-- Correa 2015 older women, increase muscle thickness and strength with volume  -->
@RN1472] <!-- Galvao 2005 Older adults, no diff in body comp DXA, but increase in strength  -->
but not confirmed in another study
@RN1518].<!-- Radaelli 2014 older women no diff between volume groups in muscle thickness or strength gains -->
Studies shows that more volume does not lead to increased muscle gains in young individuals
[@RN1384; <!-- Rhea 2002 young men bod pod no diff in body comp changes but diff in strength gains in favour of multi set -->
@RN1456;<!-- Starkey 1996 young male and female no diff in strength or muscle thickenss gains -->
@RN2463]<!-- Bottaro 2011 untrained young male similar results in muscle thickness and strength -->
a conclusion challenged by others 
[@RN776;<!-- Rønnestad, three set more effectiv in increaseing strength and muscle mass in lower body compared to upper body -->
@RN1570]. <!-- Radaelli 2014 1 vs 3vs 5 sets, 5 sets most effective in muscle thickness and some strength exercise -->

As previously noted, combining the above results and additional studies, meta-analyses concluded that training volume dose-dependency exists for the development of muscle mass and strength
[@RN2570;
@RN2492;<!-- Ralston 2017 meta analysis medium to high volume more effective strength gains -->
@RN793;
[@RN1767;
@RN789].
As a second argument against additional volume in resistance training recommendation has been the cost/benefit relationship of adding training volume without meaningful or substantial additional gains
[@RN2538;<!-- Feigenbaum 1999 questions the analysis and the practical utility of volume -->
@RN2568],<!-- Carpinelli Berger in retrospect questions the practical significance of 1.8 kg diff in maximal strength -->
a subsequent question is, whom would benefit from greater volumes and whom would not?
Schoenfeld *et al.* combined data from published studies to explore if participant characteristics of the above mentioned studies interacted with training volume in explaining study outcomes. Neither sex, muscle groups nor age interacted with volume prescription indicating that no such factor would be able refine training prescription guidelines
[@RN1767].<!-- Schoenfeld 2016 meta-->
As the number of studies used to synthesis the meta-analysis was relatively low (*n* = 15) and the studies were heterogeneous in terms of e.g. outcome measurements, it may have lacked in power to detect any meaningful interactions. Additionally, included studies may not have been reporting relevant characteristics for such analysis.

Collectively, the available evidence suggest that there is overlap between training outcomes in studies were different volume has been utilized.
The overlap cannot, with available data, be explained by general population characteristics such as age or sex.
Studying the effect of different training volumes within participants could potentially help to define determinants of training outcomes in response to different volume conditions.
Two within-participant studies have investigated the effects of training volume on strength and hypertrophy outcomes.
Sooneste *et al.* compared strength outcomes in response to three- and one-set elbow flexor training for 12 weeks in young males using a whitin-participant protocol (arms allocated to either volume condition).
The results showed general benefit of three- over one-set training for muscle hypertrophy and tended to do so also for strength gains [@RN1607].
No attempts were made to relate baseline characteristics to the magnitude of differences between volume conditions, presumably due to the small sample size (*n* = 8).
Mitchell *et al.* compared muscle hypertrophy and strength gains in response to three- and one-set of knee-extension exercise performed three times per week for ten weeks.
The study contained an additional training condition (low intensity, 30% of 1RM performed with three sets) with participants legs assigned to either of the three conditions in a random fashion.
No significant differences were reported between volume conditions for muscle mass or strength gains [@RN834].
However, the analyses were performed without taking the correlation between individuals into account due to the mixed design [@RN834].
No attempts were made to relate any measured characteristic to differences in responses.

## Molecular determinants of training-induced muscle hypertrophy
Muscle mass fluctuates as a consequence of the balance between muscle protein synthesis and breakdown. When a net-positive balance is achieved, muscle protein accumulates and muscle mass increases.
Following a single bout of resistance exercise, muscle protein synthesis increases
over resting levels up to 48-h post-exercise 
[@RN2710;<!-- Dreyer 2006, blunting of protein synthesis during ex, increase post ex -->
@RN2711;<!-- McDougall 1995 Highest values at 24h after ex, at 36 not sig higher than ctrl -->
@RN786;<!-- Phillips 1997 fasted state, FSR and FBR both increased but FBR to a lesser extent giving increase in net balance. Fasted state -->
@RN2712;<!-- Biolo 1995, fasted state ex increase both breakdown and synthesis FSR/FBR at 3h net balance increase but not positive -->
@RN2717;<!-- Increase in FSR over control at 4 and 24 h post ex in biceps, no change in RNA -->
@RN1521]<!-- Kim, increase after exercise in mixed and myofibrillar prot synth in trained and untrained state, lower mixed after training, higher basal protein synt rate after training period 8 weeks 2-4set 8RM 8 weeks biopsy 12 h after ex Same absolute intensity -->
after being blunted during exercise
[@RN2710].<!-- Dreyer 2006, blunting of protein synthesis during ex, increase post ex -->
Muscle protein synthesis and breakdown rates are highly correlated
[@RN2713;<!-- Phillips 1999, training alters both breakdown and synthesis --> 
@RN786]<!-- Phillips 1997, correlation between fbr and fsr-->
indicating that these processes are mechanistically coupled and fluctuates together.
While acute resistance exercise thus also stimulates to breakdown of muscle protein, it does so to a lesser extent, leading to an increase in the net protein balance from baseline under favorable conditions
[@RN2713;<!-- Phillips 1999, training alters both breakdown and synthesis --> 
@RN786;<!-- Phillips 1997, correlation between fbr and fsr-->
@RN2714;<!-- Børsheim 2002 Post ex prot synthesis increased with ingestion of essential amiono acids -->
@RN2715].
<!--[@RN1908] Millward 1979 Muscle protein mass is regulated primarily through alterations in protein synthesis in all except emergency conditions. --> 
When resistance training is performed under such favorable conditions, in the fed state with availability of dietary protein, a net positive protein balance can be expected after exercise
[@RN2714;<!-- Børsheim 2002 Post ex prot synthesis increased with ingestion of essential amiono acids -->
@RN2715].<!-- Tipton 1999, post-exercise net balance is increased when amino acids are administrated -->

Indispensable for the process of protein synthesis is the ribosome, a cellular machine capable of translating genetic information in the form of messenger RNA (mRNA) to proteins.
A synthesizing ribosome consists of four ribosomal RNA species (rRNA 18S, 5.8S, 28S and 5S) and about 80 proteins constructing two ribosomal subunits. Translation of mRNA occurs at the ribosomal core as ribosomal RNA catalyzes binding of amino acids to form a polypeptide chain corresponding to the ribosome-bound mRNA sequence.
The rate of translation per ribosome (i.e. translational efficiency) can be modified by specific stimulus, of which mechanical stress and nutrients are of importance to this discussion. 
Mechanical stress (such as resistance exercise) or amino acid availability (as increased after ingestion of dietary protein) stimulates to protein synthesis through increased translational efficiency.
This can be observed as the formation of polysomes as functional ribosomes bind to mRNA in response to e.g. mechanical stress
[@RN866;<!-- Baar/esser increased formation of polysomes after resistance exercise in rats  -->
@RN860].<!-- Kubica formation of polysomes in response to resistance exercise are abolished with rapamycin in rats -->
The available data from acute studies on protein synthesis in humans suggests that resistance training leads to muscle hypertrophy through the accumulative effect of repeated bouts of an anabolic stimuli. In recent years, this view has been supplemented by evidence suggesting that chronic resistance training also leads to increased rates protein synthesis at rest
[@RN1521;<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --> 
@RN1866;<!-- Wilkingson 2008 -->
@RN1897],<!-- Reidy 2017 Increased rested state protein synthesis -->
which has been postulated to be associated with an accumulation of ribosomes, i.e. an increased translational capacity
[@RN1897;<!-- Reidy 2017 -->
@RN1912]<!-- Figueiredo 2017 Ribosome biogenesis review -->.
This notion is supported by training-induced increases in muscle RNA abundance.
As the RNA pool to a large degree consists of ribosomal RNA
[@RN2054;<!-- Zak 1967 85% of RNA is rRNA in heart muscle -->
@RN2223]<!-- Young Ch 40 ~ 85% of RNA is ribosomal in rat skeletal muscle -->,
total RNA can be used as a surrogate measure of ribosomal abundance. 

Given the above discussion, the concepts of translational efficiency and capacity can be illustrated by data from Millward
[@RN2145],
wherein RNA concentrations and its association with protein synthesis rates were measured in rats starved or fed a diet containing protein.
Protein feeding increased the rate of protein synthesis, but the relationship to RNA (ribosomal) abundance was largely maintained (Figure \@ref(fig:Millward1973)). This underlines the fundamental importance of ribosomal abundance and activity in determining protein synthesis. 


(ref:millward1973) [@RN2145]

```{r Millward1973, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2.5, fig.width = 2.5, fig.cap="Relationship between RNA abundance and protein synthesis in rat skeletal muscle. Rats were either starved or fed a protein rich diet stimulating protein synthesis. Data from (ref:millward1973).", fig.scap="Relationship between RNA content and protein synthesis in rat skeletal muscle, data from (ref:millward1973)", fig.align='center'}

millward <- read_excel("./data/litrev.xlsx", sheet = "millward1973") 

source("./R/themes.R")

millward %>%
        mutate(group = if_else(group == "A", "Protein fed", "Protein starved")) %>%
        ggplot(aes(RNA, proteinSynthesis, fill = group, color = group)) +
        geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) + 
        geom_point(color = "black", shape = 21, size = 2.5, alpha = 0.6) +
        scale_x_continuous(name = TeX("$RNA\\;(\\mu g \\; mg^{-1} \\, protein)$"), 
                           breaks = c(0, 4, 8, 12, 16),
                           minor_breaks = c(2, 6, 10, 14),
                           limits = c(0, 16), 
                           expand = c(0, 0)) +
        scale_y_continuous(name = TeX("$Protein\\,synthesis\\,rate\\;(days^{-1})"), 
                           breaks = c(0, 0.05, 0.1, 0.15, 0.20, 0.25),
                           minor_breaks = c(0.02, 0.06, 0.10, 0.14, 0.18, 0.22),
                           limits = c(0, 0.25), 
                           expand = c(0, 0)) + 
        scale_color_manual(values = c(group.study.color[2], group.study.color[5])) +
        scale_fill_manual(values = c(group.study.color[2], group.study.color[5])) +
        

        dissertation_theme() +
                theme(legend.position = c(0.3, 0.9), 
                      legend.title = element_blank(), 
                      legend.text = element_text(size = 7)) 



```


### mTORC1 a multifaceted coordinator of cell growth

The discovery of an organic compound called rapamycin in the 1960’s led to the characterization of a rapamycin sensitive protein involved in cell growth. The protein was later named mechanistic target for rapamycin (mTOR)
[@RN1049].<!-- Goodman review -->
mTOR is found in two protein complexes (mTOR complex 1, mTORC1; mTOR complex 2, mTORC2) where primarily mTORC1 has been found to be sensitive to rapamycin treatment
[@RN1049].<!-- Goodman review -->
Bodine *et al.* performed a comprehensive characterization of mTORC1-mediated skeletal muscle hypertrophy using rodent models, showing that mTORC1 activation was essential for load-induced hypertrophy. Additionally, using transfection techniques, they showed that constitutively activated signaling upstream of mTORC1 (Akt) led to hypertrophy in an mTORC1-dependent manner, confirmed with concurrent administration of rapamycin
[@RN782].<!-- Bodine 2001 -->
Further utilizing rapamycin in genetically modified mice where mTOR was made rapamycin-resistant, specifically in skeletal muscle cells, confirmed that muscle fiber-specific rapamycin-sensitive mTORC1 signaling was needed to induce muscle hypertrophy in response to mechanical loading
[@RN1072].<!-- Goodman transgenic muscle-specific rapamycin insensitive mice confirm that skeletal muscle mTORC1 is responsible for the effect of rapamycin   -->
These mechanistic studies supports previous observational evidence connecting mTORC1 signaling to muscle growth in rats
[@RN866],<!-- Baar Esser -->
and more recently, in humans
[@RN791;<!-- Burd Correlation in single session between p70 and  -->
@RN785].<!-- Terzis correlation between p70S6K and hypertrophy-->

Administration of rapamycin in humans has also confirmed that mTORC1 signaling is important for protein synthesis in the acute phase after resistance training (2 hours) and in response to protein ingestion
[@RN780;<!-- Drummond administration of rapamycin blocks protein synthesis induced by res ex -->
@RN781].
However, extending the time-frame (up to 24 hours), differences in responses to resistance exercise between rapamycin treatment and control conditions were less pronounced
[@RN2826].
This could in part be explained by the lower dosage of rapamycin administrated in human compared to animal trials but also indicate rapamycin-insensitive mechanisms controlling translational efficiency
[@RN2839;<!-- Choo 2008 mTOR/raptor is needed but 4E-BP1 phosphorylation is recovered in spite of rapamycin treatment -->
@RN2836].<!-- Ogasawara 2018 rapamycin is mTORC1/S6K1 specific in inhibition, additional inhibiton of mTOR kinase (both mTORC1 and 2) also leads to complete downregulation of 4E-BP1 -->

mTORC1 functions as a signaling hub by integrating multiple environmental cues to regulate cellular growth. Among such cues is mechanical stimulation which leads to accumulation of phosphatidic acid in muscle cells
[@RN2844].<!-- O Neil 2009 Increase in phosphatidic acid and mTORC1 activity, not affected by PI3K/Akt inhibiton -->
Such accumulation was shown to be independent of mechanisms upstream of mTORC1 (phosphoinositide 3-kinase (PI3K)/AKT and ERK signaling)
[@RN2844;<!-- O Neil 2009 Increase in phosphatidic acid and mTORC1 activity, not affected by PI3K/Akt inhibiton -->
@RN2119]<!-- ERK inhibition not inhibiting PA induced mTORC1 activation -->
but still readily led to mTORC1 activation
[@RN2844;<!-- O Neil 2009 Increase in phosphatidic acid and mTORC1 activity, not affected by PI3K/Akt inhibiton -->
@RN2119].<!-- ERK inhibition not inhibiting PA induced mTORC1 activation -->
In cellular models, phosphatidic acid has been shown to interact with mTORC1 on the same site that is targeted by rapamycin 
[@RN2126]<!-- Fang PA leads to activation of mTOR, PA binds to FRB domain (same as rapamycin) -->
indicating a more direct link between mechanical stimulation and mTORC1 activity.
The enzyme primarily responsible for mechanically induced increases of phosphatidic acid has been shown to be diacylglycerol kinase $\zeta$
[@RN1728].<!-- You 2014,   -->


In the context of muscle growth in response to resistance training, adequate supplementation of dietary protein augments responses
[@RN2199].<!-- Morton 2018 meta, ~ 1.6g/kg/day maximizes fat free mass gains in response to RT -->
Mechanistically, dietary protein intake increase the availability of amino acids in muscle cells and these in turn stimulate protein synthesis through mTORC1 by multiple mechanisms
[@RN2848].<!-- Goberdhan cell metabolism perspective, amino acids are sensed through multiple mechanisms by mTORC1 -->
mTORC1 capabilities to fine tune its response based on cellular status can be exemplified from studies examining responses to different amino acid compositions. 
Providing a mixture of essential amino acids potentiated mTORC1 signaling in response to resistance exercise more than provision of essential amino acids without leucin or leucin alone
[@RN1641].<!-- Apro 2015 leucin + EAA vs placebo,  leucin and EAA-leucin. Full mixture leads to the most potent mTORC1 activity (S6K1) -->

In addition to mechanical stimuli and amino acids, mTORC1 integrate several environmental cues related to growth factors, energy and oxygen status with downstream signaling differing depending on upstream signaling and cellular characteristics
[@RN2139].<!-- Laplante mTOR signaling different between tissues and input -->
Two downstream targets of mTORC1 rely much of the information related to translational control one of which is the well characterized eIF-4E (eukaryotic translation initiation factor 4E)-bining protein 1 (4E-BP1). Upon activation, 4E-BP1 releases eIF-4E
[@RN2837]<!-- Akt dependent phosphorylation of 4e-bp1, upon activation eIF-4E no longer associated with 4E bp1 -->
which enables formation of an preinitiation complex and subsequent recruitment of the small ribosomal subunit to mRNA
[@RN2838].<!-- Gingras review 43S recruitment to mRNA is the rate limiting step in translation initiation mediated by eIF4E and thus 4E-BP1 -->
eEIF-4E-dependent initiation of translation is believed to be rate limiting and thus a control point for protein synthesis.
Interestingly, formation of the preinitiation complex, induced by mTORC1-4E-BP1-mediated release of eIF-4E results in enhanced translation of special classes of mRNAs, containing a 5' structures that does not permit efficient translation
[@RN2838;
@RN2840]
Among the resulting gene products from such mRNA are growth factors cell cycle regulators such as cyclin D1 and c-Myc as well as ribosomal proteins
[@RN2838;
@RN1049].<!-- Goodman review --->

Parallell to 4E-BP1, is another well described downstream target of mTORC1, S6 kinase 1 (S6K1).
S6K1 was named after its ability to phosphorylate ribosomal protein S6, but has since been shown to have multiple roles related to both translational efficiency and indirectly to translational capacity
@RN1049].<!-- Goodman review --->
The importance of S6K1 in control of muscle mass is apparent from S6K1 depletion in mice that results in reduces muscle growth and experiments showing that constitutevly active S6K1 resulting in increased myotube growth in cell cultures
[@RN2828;<!-- Ohanna  -->
@RN783].<!-- Rommel 2001 Cell culture, constituvely active s6k1 leads to increased  -->
S6K1 deficient mice did however not show a reduced translation of 5'TOP mRNA, an effect still sensitive to rapamycin
[@RN2827].<!-- Pende 2004, S6K1/S6K2 null mice,  -->
Instead, S6K1 deficient mice have been shown to be unable to induce transcription of genes related to ribosomal biogenesis
[@RN2321].<!-- Chauvin S6K1/S6K2 null mice does not respond with ribosomal biogenesis transcription after feeding -->
Upon activation of Akt, such mice fail to respond with increased ribosomal biogenesis, estimated through accumulation of total RNA and confirmed by targeted rRNA analysis
[@RN2849].<!-- Marabita 2016 S6K1 null mice compensatory 4EBP1 activity leads to hypertrophy but abolished ribosomal biogenesis and less strength -->
S6K1 activity also leads to phosphorylation of downstream targets that enables translation initiation and elongation in addition to its most well known substrate ribosomal protein S6 (rpS6)
[@RN2320].<!-- Goodman review 2019 -->
Although a target of mTORC1, rpS6 have been shown to have conterintuitive role in protein synthesis.
Despite its location within the ribosome close to its core, mice genetically modified to be unable to phosphorylate rpS6 upon stimulation still form polysomes indicating that rpS6 phsophorylation is not needed for translational initiation
[@RN2824].<!-- Ruvinsky 2006 review of rpS6 -->
Protein synthesis rates in the same mice are also higher compared to wild type mice suggesting an inhibitory role of rpS6 phosphorylation in protein synthesis
[@RN2824].<!-- Ruvinsky 2006 review of rpS6 -->
Interestingly, mice depleted of S6K1 showed reduced specific force compared to wild-type mice, coinciding with forming of protein aggregates
[@RN2849]<!-- Marabita 2016 S6K1 null mice compensatory 4EBP1 activity leads to hypertrophy but abolished ribosomal biogenesis and less strength -->
Together these observations points to fine-tuning mechanisms in the S6K1-rpS6-axis, balancing protein synthesis, protein quality and energy wastage
[@RN2849;<!-- Marabita 2016 S6K1 null mice compensatory 4EBP1 activity leads to hypertrophy but abolished ribosomal biogenesis and less strength -->
@RN2824].<!-- Ruvinsky 2006 review of rpS6 -->
Fine-tuning may also exist within the mTORC1-S6K1 axis as S6K1 stimulates to mTOR phsophorylation at Ser\textsuperscript{S2448}
[@RN1902],
a commonly used read-out for mTORC1 activity
[@RN1949].

By unknown mechanisms, mTORC1 signaling is sensitive to training status evident from changes in acute signaling in response to resistance training depending on the acute training status
[@RN2320;<!-- Goodman unknown mechanism in training-induced changes in exercise induced mTORC activity changes-->
@RN1809].<!-- Brook abolished mTORC1 induction after 3 and 6 weeks of training -->







### Ribosomal biogenesis and resistance training induced muscle hypertrophy

The overall role of ribosomal abundance in determining protein synthesis and subsequent cellular and tissue size was briefly mentioned above.
In addition to correlations between RNA abundance and protein synthesis in mice 
[@RN2145]<!-- Millward -->
and cell culture
[@RN1754] <!-- West correlation in cell culture -->
inhibition of ribosomal RNA (rRNA) transcription or inhibition of up-stream transcription factors act to diminish muscle cell growth upon stimulation
[@RN1755<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty -->;
@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
@RN1810].<!-- von Walden 2016 Inhibition of Pol 1 rDNA tracsription and myotube growth -->
In the context of resistance training-induced muscle growth, observational evidence from human studies further supports a determining role of ribosomal biogenesis to achieve increased translational capacity and enable hypertrophy.
Figueiredo *et al.* observed a correlation between the changes in RNA abundance and magnitude of muscle growth over eight weeks of resistance training
[@RN1644].<!-- Figueriedo 2015 correlation between delta csa and fold change of total RNA -->
Stec and collegues observed increased ribosomal RNA and total RNA abundance only in participants that were classified as modest or extreme responders in terms muscle growth but not low responders after four weeks of resistance training
[@RN1755].<!-- Stec 2016 only high responders have increased total RNA and rRNA abundance after short training -->
Similarly, Mobley *et al.* found larger increases in total RNA in participants classified as high- vs. low-responders to 12 week resistance training (34% vs. 8% increase in total RNA) together with a correlation between total RNA increases and muscle growth over the same time period
[@RN2055].<!-- Mobley weak correlation r = 0.28 p = 0.026  and no cluster * time interaction effect but increases in total RNA 34% in high responders vs 8% in low -->
Together these studies underlines the importance of ribosomal biogenesis and translational capacity in resistance training-induced muscle hypertrophy.
It should be noted however, that protein synthesis and cellular growth may occur in the absence of ribosomal biogenesis.
In cultured myotubes stimulated with IGF-1, inhibition of ribosomal RNA transcription led to reduced RNA content but not myotube size compared to non-inhibited controls
[@RN1929].<!-- Crossland inhibition of Pol 1 upon stimulation of IGF-1 does not inhibit growth -->
In aged male participants, three sessions of resistance training did not lead to increased levels of RNA but a 30% increase in protein synthesis rates
[@RN2155].<!-- Welle three sessions over a seven day period no change in RNA but prot synth -->
Furthermore, in response to a greater number of training sessions, absence or reduced ribosomal biogenesis are observed in selected individuals but muscle growth may still be detected
[@RN1809].<!-- Brook -->
However, when e.g. comparing aged and young skeletal muscle, aged muscle typically respond with reduced hypertrophy after mechanical stimuli, coinciding with reduced ribosomal biogenesis
[@RN1809;
@RN1631],
indicating that potent ribosomal biogenesis is needed to support greater hypertrophy.
Similarly, when comparing cell culture experiments, a "broader" stimuli induced by serum compared to a single growth factor could induce greater cellular growth that subsequently require support of an increased translational capacity
[@RN1929;<!-- Crossland -->
@RN1810;<!-- von Walden 2016 Inhibition of Pol 1 rDNA tracsription and myotube growth -->
@RN1755].<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty -->

Synthesis of ribosomes is indeed a hallmark of the early response to resistance training as a single resistance-exercise session leads to increases in precursor rRNA (pre-rRNA 45S)
[@RN1656<!-- A single session upregulates 45 S 24 h post ex -->;
@RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->]
and repeated bouts lead to accumulation of rRNA and thus presumably functional ribosomes
[@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
@RN1897<!-- 11% 36 sessions young -->].
However, a time course of ribosomal transcription and accumulation in response to resistance training in humans remains largely unstudied.
Only a few studies have investigated exercise- or training-induced changes in markers of ribosomal abundance over multiple time-points.
For example, two consecutive bouts of electrically evoked muscle contractions were associated with increased levels of total RNA, with peak values being observed 72 hrs after the second bout
[@RN1520]<!-- Total RNA increase 72 h after a second bout -->. 
Using voluntary contractions, peak values in total RNA were reported after nine sessions, followed by a slight decrease to after 18 sessions [@RN1809].
These data suggest that ribosomes accumulates with some delay from initiation of training, reaches a plateau in the early phase of resistance training (three weeks) and slightly decreases as muscle mass further increases
[@RN1809;
@RN1520].

Synthesis of new ribosomes is a complex and energy demanding process. It involves the synthesis of both ribosomal proteins and four mature rRNA species as well as their assembly into functional ribosomal subunits
[@RN1920;<!-- Warner review yeast -->
@RN1940;
@RN1820].<!-- Moss review -->
All mature rRNA (except rRNA 5S) are derieved from a single pre-rRNA transcript (45S pre-rRNA). After being transcribed from ribosomal DNA (rDNA), the 45S pre-rRNA transcript goes through several "splitting" events ultimately leading to the formation of three rRNA species, 18S, 5.8S and 28S. Coinciding with step-wise splitting of the pre-cursor transcript, modifications to the rRNA structure and assembly with ribosomal proteins into precursor ribosomal subunits occur in the nucleolar compartment
[@RN1940].<!-- Ribosome biogenesis at a glance -->
After export to the cytoplasm, additional maturation steps are required in both subunits before they can form functional ribosomes  
[@RN1940].<!-- Ribosome biogenesis at a glance -->

Ribosomal biogenesis is believed to be determined by the rates of pre-rRNA transcription by RNA polymerase I (Pol I) which in turn is regulated by coordinated assembly of a complex of transcription factors at the rDNA promoter
[@RN1820].<!-- Moss review -->.
This process is co-regulated with translation as initiation of rDNA transcription is controlled by e.g. mTORC1 through multiple mechanisms.
Activation of the of the upstream binding factor (UBF) through phosphorylation is needed to initiate transcription
[@RN2582; <!-- Tuan UBF phosphorylation important for SL1-UBF interaction -->
@RN2563],<!-- Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
and such activation is partly controlled by mTORC1 activity, with its inhibition being associated with blocked UBF phosphorylation and reduced UBF availability trough de-phosphorylation of retinablastoma leading to reduced ability of UBF to recruit secondary factors to the rDNA promotor and stimulate rRNA transcription
[@RN1632; <!-- Nader rapamycin block UBF availability  -->
@RN2564].<!-- Hannan mTOR block in postmitotic cardimyocytes -->
mTORC1 also controls one of these secondary factors, TIF-1A, as rapamycin leads to specific phosphorylation and translocation away from the nuclei
[@RN2851].<!-- Mayer rapamycin translocates TIF1-A away from rDNA transcription in HeLa cells -->

Parallell to mTORC1 control, specific inhibition of MEK showed that MEK/ERK signaling is important for UBF binding to rDNA
[@RN2604].<!-- Stefanovsky 2001 MEK specific inhibition under mitogen activation inhibits rRNA transcriptionand UBF phosphorylation at 117 and 201 -->
Furthermore, the transcription factor c-Myc has also been implicated in ribosomal biogenesis as its inhibition coincides with less UBF activity and rRNA transcription irrespective of mTORC1 signaling
[@RN1754].<!-- West -->
c-Myc is also found at the rDNA promotor and is required for rDNA transcription
[@RN1834;<!--  -->
@RN1828],<!-- von Walden c-myc is enriched at the cdna promotor  -->
in addition to its role in transcription of genes imoportant for rRNA transcription, e.g. UBF
[@RN2565]<!-- c-myc activate transcription of the ubf promoter -->
Interestingly, the availability of UBF *per se* has also been shown to be a determinant of rRNA transcription
[@RN2566;<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->
@RN2564]<!-- mTOR controls UBF availability, rapamycin block serum induced increase in UBF -->
through control of rDNA gene activity
[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->

In summary, ribosomal biogensis as well as the regulation of translation is under coordinated control of several pathways integrating multiple stressors and environmental cues to regulate cellular protein synthesis. 

### Myonuclear addition and transcriptional regulation of training-induced muscle remodeling 

Skeletal muscle fibers are polynuclear cells. Each fiber contains multiple nuclei which supports its transcriptional requirements. Muscle fiber-specific nuclei lack the ability to proliferate and are therefore relying on a dormant cell population located in muscle tissue to activate and supply the fiber with transcriptional capacity
[@RN2080]<!-- Blaauw 2014 review satellite cells and hypertrophy -->
These cells, called satellite cells can activate fuse with the fiber upon specific stimuli such as hormonal
[@RN912]<!-- Egner 2013 steroid treatment leads to increased number of myonuclei -->
mechanical stress
[@RN2104].<!-- Eftestøl 2016 activation of satellite cells is dependent on mechanical strain isometric force or 50-60% of maximal force -->
Satellite cells readily activates in humans in response to resistance-exercise, as assessed by markers of satellite cell state
[@RN1048;<!-- Snijders 2012 using DLK -->
@RN2081].<!-- Mackey using Pax7, Ki67 and CD56 -->





[@RN1532]<!-- Addition of myonuclei preceds hypertrophy, gained myonuclei are retained during de-training, cellular memory?  -->




McCarthy JJ, Mula J, Miyazaki M, Erfani R, Garrison K, Farooqui AB, Srikuea
R, Lawson BA, Grimes B, Keller C, et al. (2011). Effective fiber hypertrophy
in satellite cell-depleted skeletal muscle. Development 138, 3657–3666.

During muscle hypertrophy 


where fiber-specific nuclei contains the genetic material and thus supplies each fiber with transcriptional 






[@RN2618]<!-- Newlands 1998 Transcriptional activity of myonuclei are not maximised, increase in transcription factors increase the chance of a loci being transcribed... -->

[@RN2616]<!-- Kirby 2016 nascent RNA is detected in myonuclei after SA (mice) 47S is detected in EU labeled RNA pool -->

[@RN2847]<!--  -->







Previous studies have investigated transcriptome responses to acute resistance exercise [@RN1825;<!-- Gordon 2012 Diff expr after acute exercise in trained vs. untrained (identifies ECM as higher in trained)-->
@RN2298;<!-- Damas reghulation in rested state pre post 10 w training and pre post ex before and after training n = 9--> @RN2408]<!-- Hyldahl 2011 acute activation in reponse to 100 eccentric contractions-->
and chronic resistance training [@RN2400;<!-- Robinson 2017 differences on common regulation between exercise types in you and old--> 
@RN2397;<!-- Melow 2007 Age was associated with DE of 596 genes, training in aged led to RNA profiles more like young. --> 
@RN2298;<!-- Damas reghulation in rested state pre post 10 w training and pre post ex before and after training n = 9--> @RN2299;<!-- Murton Exercise induced genes with no response in control group-->
@RN1825;<!-- Gordon 2012 Diff expr after acute exercise in trained vs. untrained (identifies ECM as higher in trained)-->
@RN774]<!-- Raue 2012. Differential expression  -->,
as well as described associations between transcriptome characteristics and degrees of muscle growth [@RN774;<!--Raue 2012. 661 genes regulated by exercise and correlated with muscle size and strength--> 
@RN753],<!-- Phillips 2013 association with  -->
and function [@RN2402;<!-- Hangelbroek 2018 gene cluster predicts 1RM, training induces expression of ECM/collagen -->
@RN2398].<!-- Stepto 2009 Difference between trained (strength and endurance) and untrained controls. A gene clsuter correlated with strength -->


Whereas these studies have merited interesting findings, they lack clear coherences in terms of differential expression events, even for classical exercise-inducible genes such as PGC1$\alpha$ [@RN2360].<!-- Pillon et al -->
This lack of clear coherence is potentially due to a combination of issues such as differences in study design and methods for synthesis and analysis of transcriptome data. First, biologically founded variability can be attributed to differences in exercise protocols (e.g., differences in exercise-volume or intensity). This makes it difficult to discern a general transcriptome exercise response, as training variables are not standardized between studies. Biological heterogeneity is also caused by differences between research participants, affecting signal-to-noise ratios and making it difficult to discern the effects of single independent factors such as training variables. 

[@RN2403]<!-- Tarnopolsky no diff (only 0.6% differentially expressed) between legs in rested state (small n=8) --> are likely to reduce this variation and to provide transcriptome data with increased biological meaningfulness. Second, technical variability can be attributed to decisions made during the bioinformatical treatment of data. 
As described by Concea *et al.*, there is no optimal pipeline for sequencing technology as new tools keep evolving and emerging, different tools should be explored to an optimum pipeline for the specific type of data [@RN2426].<!-- Conessa 2016 best practices fom RNA seq-->
To exploit the potential of any study design, there is a need for identifying an appropriate pipeline for transcriptome analyses to ensure a biologically valid interpretation of data. This entails identifying potential violations of common assumptions caused by the experimental model at hand, relating to, for example, data normalization [@RN2414;<!-- Robinson 2010 TMM method normalization assumes most genes not affected by treatment-->
@RN2363]<!-- Dillies et al -->. 

For transcriptome data to provide adequate biological information about a given experimental set-up, numerous bioinformatic steps need to be adopted in a customized manner [@RN2426;<!-- Conessa 2016 best practices fom RNA seq-->
@RN2757].<!-- Arora 2020 differences between pipelines in gene abundance estimates --> 
Of these steps, data normalization is particularly decisive [@RN2363]<!-- Dillies et al -->,
as it aims to transform naïve transcript counts into biologically meaningful results. This essentially means expressing them as *per*-cell abundances [@RN2359]<!-- Loven 2012 et al  -->.
For most experimental models, this is equivalent to providing transcript-to-total RNA ratios, given the fulfillment of the assumption that total RNA levels remain stable between conditions on a per-unit-cell or per-unit-tissue basis [@RN2359]<!-- Loven 2012 et al  -->.
In cell models that exhibit high degrees of plasticity, gene expression events result in increased amounts of total RNA and mRNA transcripts per cell [@RN2430]<!-- Lin 2012 c-Myc induced gene amplification through -->, specifically violating the assumption that most genes are not differentially expressed [@RN2359;<!-- Loven 2012 et al  -->
@RN2414].<!-- Robinson 2010 TMM method normalization assumes most genes not affected by treatment-->

We are not aware of any study that has addressed the need to account for such perspectives during transcriptome analyses of skeletal muscle subjected to mechanical stress, such as resistance training. Indeed, this assumption can be expected to be violated, as total RNA content increases markedly on a per-unit-weight basis [@RN2358]<!-- Hammarstrom 2020 -->, with potential global changes also occurring for the mRNA pool, though this remains unknown. The extent to which total RNA, and therefore ribosomal RNA, increases, coincides with the increase in muscle mass [@RN2358<!-- Hammarstrom 2020 -->;
@RN1755]<!-- Stec -->, underlining its importance for cellular growth but also its inevitable presence as a potential confounding factor in RNA sequencing experiments.













## Effects of exercise volume on molecular determinants of muscle growth
Given that exercise-training variables can modify responses to resistance training, it is reasonable to assume that these effects are mediated through determinants of muscle growth.
Exercise intensity has been evaluated with respect to protein synthesis and activation of targets downstream of mTORC1 (4E-BP1 and S6K1).
Kumar *et al.* showed that maximal stimulation of fractional synthetic rate was achieved with intensities greater than 60\% of 1RM, coinciding with signaling events
[@RN2720].<!-- Kumar 2009 increasing exercise intensities maximises post ex prot synthesis at 75% in young (plateau at 60-90% of 1RM)  -->
The same group subsequently investigated the effect of training volume at an intensity presumably leading to maximal stimulation of protein synthesis (75% of 1RM).
This analysis revealed a volume-dependent dose-response as six-sets of leg extension led to greater protein synthesis one hour after exercise and sustained S6K1 phsophorylation over up to four hours after exercise
[@RN2716].<!-- Kumar 2012 higher protein synthesis after 6 sets compared to 3 sets coinciding with sustained S6K1 phosphorylation -->
Further extending the time frame, Burd and colleagues evaluated a single session consisting of either one or three sets with biopsies sampled 5, 24 and 29 hour after exercise. Both conditions led to increased myofibrillar protein synthesis five and 29 hours after exercise, but to a larger extent in response to three sets.
However, volume-dependent regulation of S6K1 was only seen at 29 hours after exercise with earlier events (< 5 hours) possibly missed due to the timing of biopsy sampling.
No clear volume-dependency was seen in p90RSK1 (downstream of ERK) or rpS6, however eukaryotic initiation factor 2B (eIF2B$\epsilon$) phosphorylation was reduced only in the three set condition at five hours post exercise [@RN791], presumably mediating translation initiation, although its exact role is still unclear
[@RN2853].<!-- Pavit eIF2B review -->
Volume dependent regulation of S6K1 at Thr389 and rpS6 at Ser235/236 were reported 30 minutes after exercise by Terzis *et al.* as six sets of 6RM bilateral legpress resulted in greater phosphorylation compared to three and one set
[@RN784].
No clear differences between volume conditions were seen in Akt at Ser473, mTOR at Ser2448, ERK 1/2 at Thr202/Tyr204, p38 ($\alpha$,$\alpha$ and $\delta$) at Thr180/Tyr182, p38$\gamma$ Thr180/Tyr182 or AMPK at Thr172
[@RN784].
Corroborating previous observations regarding exercise volume-dependence of the S6K1-rpS6-axis, Ahtiainen and colleagues also found greater phosphorylation of S6K1 at Thr389, rpS6 at Ser235/236 and Ser240/244 30 minutes after exercise with ten compared to five sets of 10RM legpress.
Furthermore, ERK1/2 at Thr202/Tyr204 and p38 at Thr180/Tyr182 phosphorylations were not found to be volume dependent
[@RN1837].
However, in contrast to Terzis *et al.*, they reported volume-dependence in exercise-induced phsophorylation of AMPK$\alpha$ at Thr172, this together with AS160 at Thr642, IRS-I Ser636/639 and Akt at Ser473 together with a tendency in LKB1 at Ser428, all related to cellular energy status, glucose uptake and metabolism
[@RN1837;
@RN2856]<!-- Sano 2003 when unable to phosphorylate AS160 glut4 translocation to membrane is inhibited in  
with relatively high exercise volumes.-->

Collectively, although limited in precision due to small sample sizes (n = 8-19) and temporal resolution and 
<!-- Mitchell n = 18, Burd n = 8, kumar n = 12, Ahtiainen n = 19, Terzis = 8 --> these studies points to volume dependency in exercise-induced S6K1\textsuperscript{Thr389} phosphorylation, potentially further augmented when increased number of sets are examined in the acute phase 
(> 3 sets, < 5 hours; [@RN834; @RN791] vs. [@RN1837; @RN784; @RN2716]).
Increased S6K1 phosophorylation coincides with rpS6\textsuperscript{Ser235/236} and \textsuperscript{Ser240/244} in the acute phase (< 5 hours) after exercise
[@RN784; @RN1837] but not in the late phase (29 hours) [@RN791].
Although rpS6\textsuperscript{Ser235/236} is a known substrate for both S6K1 and ERK1/2 and p90RSK1
[@RN2311],
ERK1/2\textsuperscript{Thr202/Tyr204} was not been found to be volume-sensitive in the acute phase
[@RN784; @RN1837],
nor was P90RSK1\textsuperscript{Thr573} in the late phase [@RN791]
suggesting that signaling along the mTORC1-S6K1-rpS6-axis is more sensitive to variable volume than the parallell MEK/ERK-axis.
The above signaling events suggest exercise volume-dependent activation of the translational machinery, and increased volume indeed leads to increased muscle protein synthesis
[@RN791; @RN2716].
However, relatively high exercise volumes may lead to signaling events inhibiting protein synthesis as AMPK\textsuperscript{Thr172} showed volume dependency, presumably relating to cellular energy stress and glucose metabolism
[@RN1837;
@RN2856;<!-- Sano 2003 when unable to phosphorylate AS160 glut4 translocation to membrane is inhibited in  
with relatively high exercise volumes.-->
@RN2858].<!-- Cork review of cross talk between AMPK and mTORC in nutrient signaling, AMPK can inhibit mTOR -->
Interestingly, an increased AMPK-related signaling may be related to an accumulated effect due to short inter-set rest periods, coinciding with reduced proteins synthesis, at least in the acute phase after exercise
[@RN2857].<!--McKendry 2016 short interset rest leads to increased AMPK and coincides with reduced protein synthesis 0-5 h post exercise -->

In addition to signaling events and , training-induced lead to increased satellite cells 
[@RN796]






