
# Background
```{r Backgroundsettings, echo = FALSE, warning=FALSE, message=FALSE}
source("./R/libraries.R"); source("./R/themes.R")
```

## Resistance-exercise prescription, influences and challenges
Recommendations of systematic physical exercise for the purpose of improving health or physical performance has long been part of human culture, evident from records dating back to ancient Chinese, Indian and Greek civilizations
[@RN2640].<!-- Tipton Ancient China India and greek/roman written records on exercise precription -->
Today's exercise-training prescription still bear traces of ideas from these eras, further developed during the renaissance and formalized in systems like German Turnen and Ling gymnastics during the nineteenth century 
[@RN2663].<!-- Pfister 2003 describes differences and common ancestry of gymnastics and sports during 1700-1800 -->
German Turnen as a system of physical activities was established in a time when Germany developed from aristocracy to a unified nation.
Turnen not only served as a system of preparing men to fight for the developing nation but also as a way of establishing a national identity.
Ling gymnastics shared common origins with German Turnen and also served as a system of military preparation.
However, Ling also established systems for medical, pedagogical and aesthetic gymnastics.
Especially Lings medical gymnastics was important for the development of modern exercise prescription as it was scientifically oriented, based on the physiological and medical understanding of that time [@RN2663].
The medical gymnastics of the nineteenth century is referenced in twentieth century texts on therapeutic exercise prescription
[@RN2634].<!-- Nicoll 1943 Dosage of exercise therapy "never exercised against more than 25% of their all out efficiency" with increased exercise periods and decreased rest periods as progression -->

With the introduction of "heavy resistance exercises" for the development of muscle strength and mass after injury, DeLorme outlined a system on which modern resistance-exercise prescription is based [@RN2633].<!-- DeLorme 1945 case reports of systematic resistance training for restoration of muscle power and mass. First mention of RM -->
DeLorme published his system short after the Second World War [@RN2633]
during which he, as a newly graduated physician, had been working with war injury rehabilitation
[@RN2639]<!-- Todd 2012 DeLorme biography -->.
Inspired by practitioners of weight training [@RN2639],<!-- Todd 2012 DeLorme biography --> 
DeLorme specifically emphasized high-resistance, low-repetition exercise where progression was achieved with increased resistance [@RN2633] as opposed to previous recommendations of endurance-like exercise where progression was achieved through increased number of repetitions
[@RN2634].<!-- Nicoll 1943 -->
DeLorme originally used the term "heavy resistance exercises" in contrasts to low-resistance exercises [@RN2633], but as this could be perceived as exercises performed only with heavy weights, the system was renamed *progressive resistance exercise* to better reflect the method
[@RN2641].<!-- DeLorme and Watkins 1948 -->
Indeed, central to the system was the concept of repetition maximum [@RN2633].
Repetition maximum refers to the external resistance that can be overcome with a given number of repetitions.
By adjusting external resistance to each individuals progression over the course of a training program, exercises are both individualized and progress is monitored [@RN2633].
DeLorme originally prescribed sessions of up to 100 repetitions performed in sets of 10 repetitions [@RN2633] but later revised this recommendation to three sets of 10 repetitions performed with increasing intensities
[@RN2641]. <!-- DeLorme 1948 Technics of progressive resistance exercise -->

Scientific inquiries into prescription of resistance training from the first part of the twentieth century concerned its therapeutic use
[e.g. @RN2633;<!-- DeLorme case report -->
@RN2646]<!-- DeLorme controlled trial of RT vs. traditional conservative (low-resistance) treatment -->
but was also evaluated in the context of improving strength and physical performance in healthy populations
[e.g. @RN2632;<!-- 1946 Houtz 16 untrained females increased strength over 4 weeks follwing DeLormes system -->
@RN2644; <!-- Chui 1950 Effect of strength training vs. control on physical performance in college stud -->
@RN2642]. <!-- Capen 1950 Effect of strength training in college stud.   -->
Scientific contributions soon moved from questions regarding the effectiveness of resistance training *per se* to comparing outcomes from different modes of resistance training 
[@RN2645;<!-- Hettinger 1953 comparing different loading patterns for isometric RT--> 
@RN1477;<!-- Capen 1956 Four different variations on the DeLorme system -->
@RN1472;<!-- Berger 1962 effect of varied weight training program on strength -->
@RN2659;<!-- Berger 1965 Increased %of 1RM increases strength gains -->
@RN2658;<!-- Berger 1967 Supra-maximal efforts (max of each rep) more efficient than 10RM loads   -->
@RN2656].<!-- O'Shea 1966 3 sets of 10, 5 or 3 reps on muscle strength and girth squat -->
A vocabulary for progressive resistance exercise-training developed through these investigations and parallel practice,
the introduction of repetition maximum by DeLorme being one example. These concepts established as modern definitions of exercise variables enabling precise prescription of training loads for a variety of populations and training goals
[@RN2537<!-- Fleck & Kraemer 2014 Textbook ch 5 : based on a needs analysis acute variables are selected   -->
@RN798].<!-- Bird 2005: Acute training variables leads to different training outcomes -->

Although this development started after the Second World War, resistance training was not part of general exercise guidelines until much later.
The American College of Sports Medicine (ACSM) position statement on exercise for healthy individuals from 1978 primarily concerned physical fitness in terms of cardio-respiratory fitness
[@RN2655]<!-- ACSM 1978 guidelines -->.
Since the updated 1990 ACSM statement, resistance training is recommended to be included as part of a sensible, general training program
[@RN2654].<!-- ACSM 1990 guidelines --> 
The introduction of resistance-training as part of the ACSM recommendation also coincides with specific recommendations on resistance training being part of other consensus statements
[@RN2666, Ch. 2].<!-- Physical Activity and Health: A Report of the Surgeon General provides an overview of recommendation guidelines evolution  -->
Consequently, informed by epidemiological data, the most recent general guidelines for physical activity do include resistance training [@RN2667].<!-- WHO guidelines -->

It could be argued that the above account reflects the fact that common understandings of *why* and *how to* exercise are influenced by societal norms and historic events such as the search of national identity in the nineteenth century or war injuries in the twentieth century
[@RN2663;
@RN2639].
In attempting to outline contemporary influences on exercise prescription, one could argue that the development of techniques to collect large amount of biological data is one such influence. The continuously decreasing cost of a sequenced human genome
[@RN2696] can serve as an example of this development.<!-- Wetterstrand webpage genome sequence costs -->
Such molecular techniques has enabled efforts to describe mechanisms by which exercise training induce favorable adaptations. The newly established Molecular Transducers of Physical Activity Consortium is an example of a large scale effort, explicitly initiated to develop personalized exercise recommendations and identify molecular targets through which effects of exercised may be mimicked
[@RN2678].<!-- MToPAC explicitly stating diagnostics is a goal --> 
Advances in bio-medical technologies are enablers of this enterprise and the quest to *individualize* exercise based on molecular diagnostics can be seen as a motivation for modern exercise science
[@RN2678;<!-- MToPAC explicitly stating diagnostics is a goal -->
@RN2677].<!-- Sparks 2017, exercise response heterogenity, goal to test appropriate exercise dosa based on e.g. DNA -->
Scientific research into contemporary exercise prescription can thus be understood as a part of the era of *personalized medicine*. 

A challenge facing this program is to accurately describe etiologies of response heterogeneity associated with physical training. A wide variation of individual responses are commonly observed after standardized resistance-training programs where measures of muscle strength-changes varies from -32 to +250% and muscle size-changes varies from negative to (-11%) to impressively large (+59%)
[@RN764;<!-- Hubal wide range of responses to standardized exercise  -->
@RN1741].<!-- Ahtiainen 2016 7% (strength) and 29% (muscle size) categorized as non responders (accounting for controls)  -->
By relating such variations to individuals genome (DNA)
[@RN1263],<!-- Pescatello review of the famuss data set -->
and messenger RNA (mRNA) profiles
[@RN826;<!-- Thalacker-Mercer, Cluster analysis, non-responders pro-inflammatory mRNA profile at baseline -->
@RN2684]<!-- Stokes co-relation between mRNA profiles and LMM changes  -->
we are beginning to gain knowledge about the genetic influence on training responses.
In such studies, a common strategy has been to dichotomize responses into "responders" and "non-responders" to exercise training. From a public health perspective this is probably fruitful when *non-response* is defined as the absence of meaningful health-related adaptations, or even adverse effects in response to a given training regime
[@RN2677;<!-- Sparks 2017 response heterogeneity -->
@RN2698].<!-- Roth viewpoint 3 scenarios for genomic screening -->
The existence of non-responders would have large implications regarding exercise prescription on the population scale
[@RN758],<!-- Timmons 2011  -->
and for any given individual as it, if properly diagnosed, would guide clinical decision-making.

<!-- Although "molecular exercise prescription" as a mean of eliminating adverse responses to a given training regime shows potential for future applications
[@RN2698; <!-- Roth viewpoint 3 scenarios for genomic screening 
@RN2547],
the rise of data intensive techniques has also highlighted its complexity.

In simplistic terms, genetic information is transferred from DNA to RNA through transcription and from RNA to protein through translation.
This flow of information, indicates a deterministic role of DNA in controlling the characteristics of an organism. It is however apparent that the transcription of specific genes are under tight control of several mechanisms including a group of non-genetic modifications to the DNA organization of each cell, known as epigenetic modifications
[@RN2690;<!-- Seaborn sharples review 2020 exercise metabolism as a determinant for epigenetic changes 
@RN2691].<!-- McGee Hargreaves 2019 review, epigentics/metabolism/exercise and mention intergenerational trandmission thorugh epigentic modification 
Such control determines gene expression in response to progressive resistance training with modifications persisting after a period of de-training
[@RN2683;<!-- Transcriptome is associated with epigentic changes in response to RE 
@RN2012].<!-- Seaborn 2018 epigenetic changes persist after de-training, epigenetic memory of training 
Regulation of gene expression is thus related to modifiable variables such as exercise and diet
[@RN2690;<!-- Seaborn sharples review 2020 exercise metabolism as a determinant for epigenetic changes 
@RN2691;<!-- McGee Hargreaves 2019 review, epigentics/metabolism/exercise and mention intergenerational trandmission thorugh epigentic modification -->
@RN2694;<!-- Yoshie 2020 13 mo old mice have epigenetic changes in skm due to childhood diet 
@RN2695]<!-- Godfrey 2016 review childhood environmental conditions affects epigenetics 
linking environmental factors to the complexity of gene expression.
-->
A key aspect of successful exercise diagnostics would be to take advantage of the relationship between exercise variables (i.e. modality, intensity, volume etc.) and exercise response for a given individual.
By adapting an individuals training program based on some prior knowledge about the individual, it is possible that the response could be positively affected. 
Observations supporting such notion exists as an individual classified as non-responsive to a specific exercise modality (e.g. endurance training) may be classified as a responder to another (e.g. resistance training)
[@RN2681].<!-- Changing from end to res training removed non responsiveness --> 
Even changing training variables within a specific modality have been shown to convert non-responders to responders as endurance training volume was increased
[@RN2699].<!-- Lundby increasing training volume improved non-responders -->

Although apparent reversal of non-response to exercise training has been observed by manipulating training variables, decisive indications for such manipulations are still lacking.



## Adaptations to resistance training

### Muscle hypertrophy and strength
Systematic resistance training typically increases muscle mass and strength, adaptations through which many beneficial effects on health (increased amino acid storage physical independence) and athletic performance are conveyed. 
Muscle growth is a well characterized response to resistance training.
On the whole muscle level, healthy untrained individuals can be expected to increases their muscle mass by $\sim$ 5-20% when training is conducted over two weeks to 6 months
[@RN1741; <!-- Ahtianen -->
@RN346;<!-- Wernbom  -->
@RN764].<!-- Hubal 2005 elbow flexor 18.9% increase -->
Over this time span, muscle growth is approximately linear with time
[@RN1596;<!-- Defreitas 2011 approx Linear increase in muscle mass and strength in response to 8 wk str training -->
@RN2113;<!-- Stock 2017 Concentric training only for arms muscle thickness and lean mass gains after 6 and 7 sessions -->
@RN2736]<!-- Narici 1989 Increase in CSA (MRI, linearly ~ 0.14%/day total 8.5%) and MVC (total 20.8%) no change in untrained control in CSA, numerical increases occured in untrained in EMG/MVC   -->
<!-- Although some of the early increase muscle mass can be attributed muscle damage-induced edema
[@RN2746],Damas 2016 increased echo intensity indicating muscle edema at 3 weeks -->
and has been detected as early as 3-4 weeks after training initiation, without apparent muscle edema 
[@RN1596;<!-- Defreitas 2011 approx Linear increase in muscle mass and strength in response to 8 wk str training -->
@RN2113].<!-- Stock 2017 Concentric training only for arms muscle thickness and lean mass gains after 6 and 7 sessions -->

Relative muscle growth can be expected to be more pronounced in upper- compared to lower-body muscles when loading patterns are similar
[@RN346;<!-- Wernbom  -->
@RN2739].<!-- Welle 1996 larger incr in csa in upper body -->
This possibly relates to greater every day activity levels of lower-body muscles requiring a larger stimuli for adaptation
[@RN767].<!-- Folland review -->
Small, but detectable differences in muscle growth can be expected between sexes for training induced muscle growth in the upper-body
[@RN764]<!-- Hubal CSA men: 20.4% vs 17.9% -->
but not for lower-body muscles
[@RN2740].<!-- Roberts 2020, meta analysis when all muscles combined no diff between sexes in within sex effects sizes -->
Hypertrophic responses can be expected to be reduced with increasing age 
[@RN2739; <!-- Welle 1996 larger incr in csa in upper body --> 
@RN752]<!-- Peterson 2011 Meta analysis, age (negatively) and training volume (positively) affects rt outcomes  -->
but increased with sufficient addition of dietary protein
[@RN2199]<!-- Morton meta Dietary protein supplementation significantly enhanced changes in muscle strength and size during prolonged RET in healthy adults. Increasing age reduces and training experience increases the efficacy of protein supplementation during RET. -->
Additionally, training variables such as intensity, volume, frequency (reviewed below) together with other training aids, e.g. manipulation of blood flow through pressure cuffs
[@RN2745]<!-- Lixandrao BFR meta analysis hypertrophy similar between low load BFR and high load -->
can effectively modulate resistance training induced hypertrophy.
Together, this underlines that both non-modifiable (e.g. sexual dimorphism and age) and modifiable factors (e.g. training variables and protein supplementation) affects training induced muscle hypertrophy.

An increased muscle mass transfers to muscle strength trough an increase of contractile elements in parallel.
The muscle cell is to a large degree (about 80% of the cell volume 
[@RN2731]<!-- Lüthi 1986 Myofibrils occupie more space in the mucsle fiber after resistance training mitochondria occupie less. No sign increase in fiber CSA with strength training no control group -->)
occupied by these contractile elements, called myofibrils, where the interaction between actin and myosin determines force production
[@RN2756].
With resistance training, the number of myofibrils and thus the muscles force generating capacity increases
[@RN2731].<!-- Lüthi 1986 Myofibrils occupie more space in the mucsle fiber after resistance training mitochondria occupie less. No sign increase in fiber CSA with strength training no control group -->
<!-- Although the number of fibers per muscle changes as a consequence of age
[@RN2193] (decrease with age) Lexell 1988 number of fibers reduced with age 
and activity
[@RN2742] (increase with activity), Sjöström 1991 Increase in left/right tibialis anterior -->
Whole muscles growth in response to short term resistance training occurs primarily through addition of myofibrils in a set number of muscle fibers.
This can be assumed as training-induced fiber splitting or myogenesis is likely a slow process and any increase in the number of fibers would only represent a small addition to the whole muscle  
[@RN2741;<!-- Murach 2019 review on hyperplasia  -->
@RN2742; <!-- Sjöström 1991 Increase in left/right tibialis anterior -->
@RN2754;<!-- MacDougall no diff between body builders and untrained in number of fibers, equal relative amounts of collagen -->
@RN767].<!-- Folland 2007 review on mechanism for hypertrophy and strength gains -->


[@RN2669].<!-- Straight 2020 muscle fiber size increases with RT, in midlle age old, age inversely related to training effects -->



Muscle fiber hypertrophy is difficult to study in humans as such studies are restricted to small biopsy samples.
Muscle fiber characteristics varies across the muscle biopsy specimens
[@RN2742] <!-- Sjöström 1991 Increase in left/right tibialis anterior -->



[@RN2629]<!-- Ikai 1970 Arm flexor training within-subject protocol, trained arm 23% increase in CSA and 90 % strength 100 days of 6 days/week training. Control arm 30% str incr, 5% CSA incr -->
[@RN2737]<!-- Young 1983 diproportional higher strength gains than CSA, increase in CSA with ultrasound in response to RT -->

[@RN2736]<!-- Narici 1989 Increase in CSA (MRI, linearly ~ 0.14%/day total 8.5%) and MVC (total 20.8%) no change in untrained control in CSA, numerical increases occured in untrained in EMG/MVC   -->
[@RN2735]<!-- Narici 1996 Local changes in whole muscle CSA after 6 months of RT, no stat sig increase in fiber CSA at a site with low MR based increase in CSA -->  


[@RN787]<!-- MacDougall 1980 increases in fiber are in response to training and immobilization fiber areas affected pos and neg in resp condition (small n one-tailed tests) -->

[@RN2734]<!-- Alway 1988 larger fiber area but not stat sig myofibrill larger in strength trained (n=6) than endurance trained (n=6), active (n=6) and sedentary controls (n=6). -->









[@RN1710]<!-- Straight 2015 resistance training increases muscle power in middle age and older adults, training volume influences the magnitude of change -->


### Muscle fiber-type transitions
Skeletal muscle tissue is heterogeneous in terms 


Muscle fibre composition is another potential determinant of muscular responses to resistance training. Type II fibres have greater growth potential compared to type I fibres
[@RN1755;<!-- Stec 2016 8 (type I) vs 21% (type II) and 44 vs 83% CSA change in mo and xtr responders --> 
@RN2208]<!-- Jespersen Type II fibers larger fluctuations in CSA higher hypertrophy and atrophy rates-->, and readily switch from IIX to IIA phenotypes in response to mechanical loading
[@RN2056;<!-- Andersen 2018 type IIx to 2a most prominent switch-->
@RN2169;<!-- Widrick 2002 2a to 2x switch; quantitative changes related to contractile proteins-->
@RN1489]<!-- Ellefsen 2014, training induced switch from 2A to 2X -->,
suggesting that these fibres display greater plasticity in response to resistance training.



### Mitochondrial function

Increased mitochondrial respiration after 12 weeks of RT in young men [@RN2608]

Fiber type distributions do not predict muscle mitochondrial density in endurance trained individuals [@RN2615]

PMID:158694 Reduced mitochondrial density per fiber area in reponse to RT

## Effects of exercise prescription on muscle mass and strength
Precise exercise-training^[Exercise is herein defined as an acute bout of physical activity designed to affect physical characteristics such as strength, speed or endurance. Training is defined as the systematic process of combining multiple exercise-sessions performed in sequence over time. Resistance-exercise is defined as an acute strength-promoting program requiring the neuromuscular system to exert force against resistance. Resistance training is defined as a long-term process of multiple resistance exercise-sessions performed over a defined period of time.]
prescription gives information on exercises, their sequential order, intensity and volume, rest periods between efforts or sessions and the frequency at which exercise sessions are to be performed
[@RN789].<!-- Bird designing resistance training programs -->
By manipulating these variables, resistance training programs can be tailored to better fit goals and starting points of any individual. 
The relative importance of exercise-training variables for training outcomes has been examined in numerous studies including (but not limited to) the overall organization of exercise sessions,
[@RN2575;<!-- Evans mini review on periodization --> 
@RN2572]<!-- Grgic 2017 No effect of linear and daily unduluting periodization on muscle hypertrophy -->
training frequency 
[@RN2571]<!-- Schoenfeld 2016 training frequency twice per week better in volume equated studies for hypertrophy -->,
<!-- Higher frequencies gives higher gains in strength but not when volume equated -->
and intensity
[@RN2569].<!-- Schoenfeld, no difference between high vs low load strength training in hypertrophy but 1rm favors high loads -->
It could be argued that training volume is of particular importance for muscle growth as when this variable is held constant, manipulation of other variables has little or no effect hypertrophy
[@RN1612;<!-- Schoenfeld 2014 7*3Rm vs 3*10RM equal muscle hypertrophy but dissimilar strength -->
@RN2569].<!-- Meta analysis high vs low resistance has little effect on muscle mass gains but strength is affected positively with higher intensities. --> 
For development of strength, factors such as intensity and within session organization of exercises is of importance 
[@RN2570;<!-- Frequency has little effect on strength when volume is equated -->
@RN2591],<!-- Exercise order influences strength but not hypertrophy -->
however, when other factors are held constant, increased training volume generally leads to increased strength
[@RN2570;
@RN2492;<!-- Ralston 2017 meta analysis medium to high volume more effective strength gains -->
@RN793],
similarly to effects of training volume on muscle growth
[@RN1767;
@RN789].

### Effects of resistance exercise volume on muscle strength and mass
Exercise volume can be prescribed as the within session number of sets performed per muscle group. This unit is practical as it comparable between individuals and muscle groups [@RN2130].<!-- Baz-Valle 2018 sets per exercise better than mechanical load times rep or total work -->
Berger conducted an early study concerning effects of resistance exercise volume with the goal to determine what method most efficiently produced strength gains (in healthy young males) [@RN1476]. Berger compared one, two and three sets performed with two, six or ten repetition maximum (RM) in the bench press, three times per week, over twelve weeks. As the combined effect of three sets per session was superior regardless of the number of repetitions performed Berger concluded in favor of three sets. This conclusion was later challenged on the basis of data interpretation
[@RN794;<!-- Carpenelli and Otto questions the analysis-->
@RN2538].<!-- Feigenbaum -->
Reveiwing the study by Berger and others, Carpinelli and Otto arrived to the conclusion that there was "insufficient evidence to support the prevalent belief that a greater volume of exercise (through multiple sets) will elicit superior muscular strength or hypertrophy" [@RN794]. This stand has since been repeatedly put forward as a criticism of higher volume training programs 
[@RN2568; 
@RN2201] and sparked considerable scientific activity. The main argument against the recommendation of additional volume in strength training programs has been the lack of statistically significant results in single studies [@RN2568; @RN2538]. 
Indeed, individual studies do not generally agree on dose-dependent effects of training volume on muscle mass and strength gains
[@RN2465; <!-- Ribiero 2015 Strength but not body composition change with dose dep older women -->
@RN2464;<!-- Correa 2015 older women, increase muscle thickness and strength with volume  -->
@RN2463;<!-- Bottaro 2011 untrained young male similar results in muscle thickness and strength -->
@RN1570; <!-- Radaelli 2014 1 vs 3vs 5 sets, 5 sets most effective in muscle thickness and some strength exercise -->
@RN1518;<!-- Radaelli 2014 older women no diff between volume groups in muscle thickness or strength gains -->
@RN1474;<!-- McBride 1474 Men and women DXA no diff but strength increase more in multiple sets -->
@RN1472; <!-- Galvao 2005 Older adults, no diff in body comp DXA, but increase in strength  -->
@RN1456;<!-- Starkey 1996 young male and female no diff in strength or muscle thickenss gains -->
@RN1454;<!-- Ostrowski 1997 trained young med, low volume decrease from previous levels no difference between groups -->
@RN1384; <!-- Rhea 2002 young men bod pod no diff in body comp changes but diff in strength gains in favour of multi set -->
@RN1382; <!-- Cannon 2010, Older women and younger women, muscle volume and strength increased similarly irrespective of age or training volume and without interaction -->
@RN776],<!-- Rønnestad, three set more effectiv in increaseing strength and muscle mass in lower body compared to upper body -->
including studies performed within participants, where different training volumes are allocated to either extremity 
[@RN834;<!-- Mitchell 2012 young men no sig difference in strength or mass gains between 1 and 3 sets knee extension -->
@RN1607].<!-- Sooneste 2013 young men p=0.076 on strength gains in favor of three sets, CSA increased more in 3 sets leg exercise -->
For example, differences in strength are between volume conditions are found in older individuals 
[@RN2465; <!-- Ribiero 2015 Strength but not body composition change with dose dep older women -->
@RN2464;<!-- Correa 2015 older women, increase muscle thickness and strength with volume  -->
@RN1472] <!-- Galvao 2005 Older adults, no diff in body comp DXA, but increase in strength  -->
but not confirmed in another study
@RN1518].<!-- Radaelli 2014 older women no diff between volume groups in muscle thickness or strength gains -->
Studies shows that more volume does not lead to increased muscle gains in young individuals
[@RN1384; <!-- Rhea 2002 young men bod pod no diff in body comp changes but diff in strength gains in favour of multi set -->
@RN1456;<!-- Starkey 1996 young male and female no diff in strength or muscle thickenss gains -->
@RN2463]<!-- Bottaro 2011 untrained young male similar results in muscle thickness and strength -->
a conclusion challenged by others 
[@RN776;<!-- Rønnestad, three set more effectiv in increaseing strength and muscle mass in lower body compared to upper body -->
@RN1570]. <!-- Radaelli 2014 1 vs 3vs 5 sets, 5 sets most effective in muscle thickness and some strength exercise -->

As previously noted, combining the above results and additional studies, meta-analyses concluded that training volume dose-dependency exists for the development of muscle mass and strength
[@RN2570;
@RN2492;<!-- Ralston 2017 meta analysis medium to high volume more effective strength gains -->
@RN793;
[@RN1767;
@RN789].
As a second argument against additional volume in resistance training recommendation has been the cost/benefit relationship of adding training volume without meaningful or substantial additional gains
[@RN2538;<!-- Feigenbaum 1999 questions the analysis and the practical utility of volume -->
@RN2568],<!-- Carpinelli Berger in retrospect questions the practical significance of 1.8 kg diff in maximal strength -->
a subsequent question is, whom would benefit from greater volumes and whom would not?
Schoenfeld *et al.* combined data from published studies to explore if participant characteristics of the above mentioned studies interacted with training volume in explaining study outcomes. Neither sex, muscle groups nor age interacted with volume prescription indicating that no such factor would be able refine training prescription guidelines
[@RN1767].<!-- Schoenfeld 2016 meta-->
As the number of studies used to synthesis the meta-analysis was relatively low (*n* = 15) and the studies were heterogeneous in terms of e.g. outcome measurements, it may have lacked in power to detect any meaningful interactions. Additionally, included studies may not have been reporting relevant characteristics for such analysis.

Collectively, the available evidence suggest that there is overlap between training outcomes in studies were different volume has been utilized.
The overlap cannot, with available data, be explained by general population characteristics such as age or sex.
Studying the effect of different training volumes within participants could potentially help to define determinants of training outcomes in response to different volume conditions.
Two within-participant studies have investigated the effects of training volume on strength and hypertrophy outcomes.
Sooneste *et al.* compared strength outcomes in response to three- and one-set elbow flexor training for 12 weeks in young males using a whitin-participant protocol (arms allocated to either volume condition).
The results showed general benefit of three- over one-set training for muscle hypertrophy and tended to do so also for strength gains [@RN1607].
No attempts were made to relate baseline characteristics to the magnitude of differences between volume conditions, presumably due to the small sample size (*n* = 8).
Mitchell *et al.* compared muscle hypertrophy and strength gains in response to three- and one-set of knee-extension exercise performed three times per week for ten weeks.
The study contained an additional training condition (low intensity, 30% of 1RM performed with three sets) with participants legs assigned to either of the three conditions in a random fashion.
No significant differences were reported between volume conditions for muscle mass or strength gains [@RN834].
However, the analyses were performed without taking the correlation between individuals into account due to the mixed design [@RN834].
No attempts were made to relate any measured characteristic to differences in responses.

## Molecular determinants of training-induced muscle hypertrophy
Muscle mass fluctuates as a consequence of the balance between muscle protein synthesis and breakdown. When a net-positive balance is achieved, muscle protein accumulates and muscle mass increases.
This is a scenario observed in response increased mechanical stress.
Following a single bout of resistance exercise, muscle protein synthesis increases
over resting levels up to 48-h post-exercise 
[@RN2710;<!-- Dreyer 2006, blunting of protein synthesis during ex, increase post ex -->
@RN2711;<!-- McDougall 1995 Highest values at 24h after ex, at 36 not sig higher than ctrl -->
@RN786;<!-- Phillips 1997 fasted state, FSR and FBR both increased but FBR to a lesser extent giving increase in net balance. Fasted state -->
@RN2712;<!-- Biolo 1995, fasted state ex increase both breakdown and synthesis FSR/FBR at 3h net balance increase but not positive -->
@RN2717;<!-- Increase in FSR over control at 4 and 24 h post ex in biceps, no change in RNA -->
@RN1521]<!-- Kim, increase after exercise in mixed and myofibrillar prot synth in trained and untrained state, lower mixed after training, higher basal protein synt rate after training period 8 weeks 2-4set 8RM 8 weeks biopsy 12 h after ex Same absolute intensity -->
after being blunted during exercise
[@RN2710].<!-- Dreyer 2006, blunting of protein synthesis during ex, increase post ex -->
Mixed muscle protein synthesis and breakdown are highly correlated
[@RN2713]<!-- Phillips 1999, training alters both breakdown and synthesis --> 
[@RN786]<!-- Phillips 1997, correlation between fbr and fsr-->
indicating that these processes are mechanistically coupled.
And similarly to muscle protein synthesis, acute resistance exercise also stimulates protein breakdown but to a lesser extent than protein synthesis leading to an increase in the net protein balance from baseline
[@RN2713]<!-- Phillips 1999, training alters both breakdown and synthesis --> 
[@RN786].<!-- Phillips 1997, correlation between fbr and fsr-->
When resistance exercise is performed in the fasted state, post exercise net protein balance can be expected to be negative
[@RN2713]<!-- Phillips 1999, training alters both breakdown and synthesis --> 
[@RN786],<!-- Phillips 1997, correlation between fbr and fsr-->
in contrast when performed in the fed state where net balances are positive
[@RN2714]<!-- Børsheim 2002 Post ex prot synthesis increased with ingestion of essential amiono acids -->
[@RN2715].<!-- Tipton 1999, post-exercise net balance is increased when amino acids are administrated -->
This indicates an additive effect of 

Measurement of protein synthesis (and breakdown) rates in response to a single resistance exercise session has been given a lot of attention


In recent years, this view has been supplemented<!-- supported? --> by evidence suggesting that chronic RT leads to increased resting synthetic rate of muscle protein
[@RN1521;<!-- Kim 2005 Fasted state skeletal muscle PS altered with training  --> 
@RN1866;<!-- Wilkingson 2008 -->
@RN1897],<!-- Reidy 2017 Increased rested state protein synthesis -->
which has been postulated to be associated with increased translational capacity, i.e. accumulation of ribosomes
[@RN1897;<!-- Reidy 2017 -->
@RN1912]<!-- Figueiredo 2017 Ribosome biogenesis review -->.
This notion is supported by exercise-induced increases in total RNA, a proxy marker of ribosome abundance, which is closely connected to protein synthesis
[@RN1754;<!-- West 2016 correlation between total RNA and protein synthesis in C2C12-->
@RN2145]<!-- Millward 1973 relationship between RNA and protein synthesis in rat in rats in vivo constant infusion -->
and muscle hypertrophy
[@RN1644;<!-- Figueiredo 2015 Correlation between total RNA fold change and CSA hyp -->
@RN1755; <!-- Stec Ribosome bio higher in high responders -->
@RN2358].<!-- Volume dependent RNA accum relates to volume depend hyper -->
Conversely, inhibition of ribosomal RNA (rRNA) transcription and inhibition of its up-stream transcription factors act to diminish muscle cell growth
[@RN1755<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty -->;
@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
@RN1810]<!-- von Walden 2016 Inhibition of Pol 1 rDNA tracsription and myotube growth -->.  

Biosynthesis of novel ribosomes is a complex, highly coordinated and energy demanding process that involves synthesis of both ribosomal proteins and the four mature rRNA transcripts
[@RN1920;<!-- Warner review yeast -->
@RN1820;<!-- Moss review -->
@RN1810]<!-- Inhibition of Pol II affects trasncription at Pol I -->.
Ribosomal accumulation is believed to be determined by the rates of pre-rRNA transcription by RNA polymerase I (Pol I), which in turn is regulated by coordinated assembly of a complex of transcription factors at the rDNA promoter
[@RN1820]<!-- Moss review -->.
Specifically, activation of the of the upstream binding factor (UBF) through phosphorylation is needed to initiate transcription
[@RN2582; <!-- Tuan UBF phosphorylation important for SL1-UBF interaction -->
@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->
Such activation is at least partly controlled by the mechanosensitive mTOR pathway, with its inhibition being associated with blocked UBF phosphorylation and subsequent rRNA transcription
[@RN1632; <!-- Nader rapamycin block UBF availability  -->
@RN2564].<!-- Hannan mTOR  -->
Interestingly the availability of UBF *per se* has been shown to be a determinant of rRNA transcription
[@RN2566]<!-- Hannan 1996 over expression of UBF leads to rRNA transcription in cardiomyocytes -->
through control of rDNA gene activity
[@RN2556].<!-- Sanij, UBF depleting leads to inactivation of rDNA genes -->




[@RN2724]<!-- Phillips 2002 fed state same absolute intensity, training attenuated ex induced FSR and FBR, FSR/FBR incr at rest after training   -->
[@RN2723]<!-- Tang 2008 -->

[@RN2144]<!-- Damas 2016 Correlation between integrated MyoPS and CSA increase after muscle damage accounted for. Deuterium used for MyoPS as opposed to infused tracer -->

### mTOR

The discovery of an immunosuppressant organic compound called rapamycin in the 1960’s led to the characterization of a rapamycin sensitive protein involved in cell growth. The protein was later named mechanistic target for rapamycin (mTOR)
[@RN1049]. This protein has since been shown play a key role in skeletal muscle hypertrophy in relation to mechanical loading. Bodine *et al.* made a comprehensive characterization of mTOR-mediated skeletal muscle hypertrophy using rodent models in 2001, showing that mTOR activation was essential for load-induced hypertrophy. Additionally, using transfection techniques, they showed that constitutively activated Akt signaling led to hypertrophy in an mTOR-dependent manner, confirmed with concurrent administration of rapamycin [@RN782]<!-- Bodine 2001 -->.


[@RN860]<!-- Kubica 2004, rapamycin prevents increases in ex induced protein synthesis trough eukaryotic initiation factor 2B mRNA translation (rats) -->



In this context, molecular signatures conveyed by the mechanistic target of rapamycin complex 1 (mTORC1) has been in particular focus. Inhibition of mTORC1 impairs protein synthesis in humans [@RN780] and activation of its associated downstream target S6 kinase 1 (S6K1) correlates with increases in muscle protein synthesis and subsequent muscle growth</span>
[@RN791;<!-- Burd Correlation in single session between p70 and  -->
@RN785].<!-- Terzis correlation between p70S6K and hypertrophy-->
In line with this, surplus <span custom-style="CHANGE">exercise</span> volume leads to greater phosphorylation of S6K1 
[@RN791;<!-- Burd in single session single and multiple sets phosphorylation p70   -->
@RN784;<!-- Terzis volume affect activation -->
@RN1837]<!-- Ahtiainen 5 vs 10 sets p70 and s6 different --> and <span custom-style="CHANGE">is accompanied by increases in </span>myofibrillar protein synthesis [@RN791]<!-- Burd -->,
fitting the notion that increased training volume provides more pronounced adaptations<span custom-style="CHANGE"> through repeated episodes of increased protein synthesis</span>. 
<!-- However, also from a cell biological perspective, present findings on effects of different training volumes are heterogeneous. For example, Mitchell *et al.* [-@RN834] failed to show differences in S6K1 phosphorylation between volume protocols, corroborating with similar effects of different volumes on muscle strength and mass. -->


Individual response patterns to resistance training, including muscle strength and mass, correlate closely with muscle cell characteristics, measured in both rested-state and acute training-phase conditions
[@RN826; <!-- Thalacker-Mercer 2013 differences between response cluster in gene expression -->
@RN774; <!-- Raue 2012 Gene expression signatures correlate with strength and CSA gains -->
@RN1755; <!-- Stec 2016 Differential ribosome biogenesis between responders and non responders -->
@RN785]<!-- Terzis correlation between P70 and hypertrophy -->. <span custom-style="CHANGE">



### Ribosomal biogenesis 

Conversely, inhibition of ribosomal RNA (rRNA) transcription and inhibition of its up-stream transcription factors act to diminish muscle cell growth
[@RN1755<!-- Stec 2016 Pol I-specific inhibition leads to GF induced myotube hypertrophty -->;
@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
@RN1810]<!-- von Walden 2016 Inhibition of Pol 1 rDNA tracsription and myotube growth -->.  

Biosynthesis of novel ribosomes is a complex, highly coordinated and energy demanding process that involves synthesis of both ribosomal proteins and the four mature rRNA transcripts
[@RN1920;<!-- Warner review yeast -->
@RN1820;<!-- Moss review -->
@RN1810]<!-- Inhibition of Pol II affects trasncription at Pol I -->.
Ribosomal accumulation is believed to be determined by the rates of pre-rRNA transcription by RNA polymerase I (Pol I), which in turn is regulated by coordinated assembly of a complex of transcription factors at the rDNA promoter
[@RN1820]<!-- Moss review -->.
Specifically, activation of the of the upstream binding factor (UBF) through phosphorylation is needed to initiate transcription
[@RN2582; <!-- Tuan UBF phosphorylation important for SL1-UBF interaction -->
@RN2563].<!--Lin 2007 phospho of UBF shows it has a role in recruioting SL1 to the rDNA promoter -->





[@RN2563] <!-- Lin 2007 Phosphorylation at Ser389 is needed for interaction with SL1, but ser584 do not, non of these phosphorylation sites leads to location to nucleolar   -->
[@RN2602]<!-- UBF is a target for cyclin d1 and cyclin E NIH 3T3 fibroblasts,Mutation of Ser484 impairs rDNA transcription  -->

[@RN2604]<!-- Stefanovsky 2001 ERK inhibition through MEK specifik inhibitor blocks EGF induced rRNA transcription in vitro. Direct Raf activation -> ERK activity stimulated rRNA transcription. UBF is a target of ERK signalling  -->

[@RN2717]<!-- Chesley 1992 increase in FSR in response to re-ex i biceps, no change in RNA content, increase in "RNA activity" prot synth/RNA -->

[@RN2155]<!-- Welle 1999 No increase in mRNA or total RNA after 7 days in older subjects but 30% increase in myofibrillar protein synthesis -->

### Transcriptional regulation of muscle mass

[@RN2618]<!-- Newlands 1998 Transcriptional activity of myonuclei are not maximised, increase in transcription factors increase the chance of a loci being transcribed... -->

[@RN2616]<!-- Kirby 2016 nascent RNA is detected in myonuclei after SA (mice) 47S is detected in EU labeled RNA pool -->


#### Ribosomal biogensis as a determinant of RT-induced hypertrophy

Resistance exercise is a potent stimuli for rRNA transcription as a single session leads to increases in pre-rRNA
[@RN1656<!-- A single session upregulates 45 S 24 h post ex -->;
@RN1037<!-- Nader 2014 45S up in response to 4 h after acute exercise -->]
and repeated bouts lead to accumulation of mature rRNA thus also total RNA and presumably functional ribosomes
[@RN1656;<!-- Stec2015 A single session ns ~ 9% increase in young, -4% in old -->
@RN1520;<!-- Bickel 2005 Two bouts electric stimulation, young ~ 30% incr-->
@RN2358; <!-- Four sessions ~29 and 18 % in mod vs. low volume|31 sessions 18/12% young-->
@RN1755<!-- Stec 2016 12 sessions 26, 9 and 6% increase in high, mod and low responders OLD-->;
@RN1644;<!-- Figuieredo 2015 16 sessions 49% increase young -->
@RN1809; <!-- Brook 18 sessions 8 and 40% (9 sessions) and 15,22% (18 sessions) old and young-->
@RN1897<!-- 11% 36 sessions young -->].
However, the true time course of ribosomal transcription and accumulation in response to RT remains largely unstudied, with a mere few studies having investigated exercise-induced changes in rRNA over multiple time-points, all of which are either limited to a selected few time-points or a limited time frame.
For example, two consecutive bouts of electrically evoked muscle contractions were associated with increased levels of total RNA, with peak values being observed 72 hrs after the second bout
[@RN1520]<!-- Total RNA increase 72 h after a second bout -->. 
Using voluntary contractions, peak values were reported after nine sessions, followed by a slight decrease to after 18 sessions [@RN1809],
resembling data from our lab where five sessions of RT led to marked increase in total RNA levels (per-unit muscle tissue), whereby a numerical lowering occurred to after the last training session of the 12 wk interventions (31 sessions).
[@RN2358].
Interestingly, during the initial phase of RT, total RNA accumulation seems to be volume-dependent, as three sets per exercise in leg exercises led to augmented total RNA and rRNA levels compared to one set per exercise, coinciding with the differences in muscle hypertrophy seen after 12 weeks of RT.
[@RN2358].
These data suggest that ribosome accumulation reaches a plateau in the early phase of RT and that increases are sensitive to training volume in constant volume protocols. 


Recent observations in humans are challenging this view by indicating that translational capacity is a limiting factor for training-induced muscle hypertrophy. First, increased abundances of <span custom-style="CHANGE">ribosomal RNA (rRNA)</span> in response to resistance training, measured as total RNA per-weight-unit muscle tissue, correlate with muscle hypertrophy [@RN1644]<!-- Figueiredo et al., 2015)-->.
In accordance with this, training-induced increases in rRNA are larger in muscle hypertrophy high-responders than in low-responders
[@RN1755;<!--Stec 2016 4 weeks rRNA and total RNA abundance larger in high responders -->
@RN2055].<!-- Mobely 2018 Post-hoc comparisons at week 12 indicate weaker association -->
Secondly, elderly participants typically show blunted ribosome biogenesis, coinciding with attenuated hypertrophic responses
[@RN1656;<!-- Stec 2015 acute bout, blunted 45S induction --> 
@RN1809]<!-- Brook 2016 blunted responses in elderly together with blunted rna induction-->.
Collectively, these observations suggest that muscle growth depends at least in part on increased translational capacity, making it a prime candidate for explaining the diverse response patterns seen in resistance training with different volume in different individuals. To date, no study has investigated the association between training volume, ribosome biogenesis and regulation, and gross training adaptations.


<!-- OUTLINE 






<!-- 
Also in humans, administration of rapamycin, hindering the activity of mTOR leads to an abolished exercise-induced increase in protein synthesis
[@RN780]<!-- Drummond rapamycin blocks protein synthesis in humans -->. 

<!-- 
Furthermore, observational evidence linking mTOR to load-induced hypertrophy comes from human and rodent studies correlating acute phosphorylation the downstream target of mTOR, ribosomal protein S6 kinase (S6K) in response to acute mechanical loading and hypertrophic responses following a subsequent training period
[@RN785;<!-- Terzis 2008 mTOR activity correlate with subsequent adaptation 
@RN866]
--> 

<!-- 
### Protein synthesis


Positive net protein balance in responce to exercise <!-- Tipton 10.1152/ajpendo.00234.2002.-->

<!-- Coordinated mTOR 10.1038/nature13492. -->

<!-- Inhibition of RNA synthesis restrict protein synthesis Goldspink 1977 -->

<!--Indicated in Goldspink 1977 and 1976 RNA reflects ribosomal availability  Goldspink 1977-->

<!--Protein synthesis is proportional to RNA content  Millward 1973 -->


<!-- Increase loading leads to increased RNA Goldberg 1975 Mechanism of work-induced hypertrophy of skeletal muscle 
```{r Millward1973, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2.5, fig.width = 2.5, fig.cap="Data from Millward et al. 1973. Group A were fed a diet containing protein, group B were starved or fed a diet not containing protein."}

millward <- read_excel("./data/litrev.xlsx", sheet = "millward1973") 

millward %>%
        mutate(group = if_else(group == "A", "Protein fed", "Protein starved")) %>%
        ggplot(aes(RNA, proteinSynthesis, fill = group, color = group)) +
        geom_smooth(method = "lm", se = FALSE) + 
        geom_point(color = "black", shape = 21) +
        scale_x_continuous(name = TeX("$RNA\\;(\\mu g \\; mg^{-1} \\, protein)$"), 
                           breaks = c(0, 4, 8, 12, 16),
                           minor_breaks = c(2, 6, 10, 14),
                           limits = c(0, 16), 
                           expand = c(0, 0)) +
        scale_y_continuous(name = TeX("$Protein\\,synthesis\\,rate\\;(days^{-1})"), 
                           breaks = c(0, 0.05, 0.1, 0.15, 0.20, 0.25),
                           minor_breaks = c(0.02, 0.06, 0.10, 0.14, 0.18, 0.22),
                           limits = c(0, 0.25), 
                           expand = c(0, 0)) + 
        

        dissertation_theme() +
                theme(legend.position = c(0.2, 0.8), 
                      legend.title = element_blank()) 



```

### The mammalian target of rapamycin (mTOR) and translational efficiency
The mammalian target of rapamycin (mTOR) is a large serine-threonine protein kinase wich in complex with oyher regulatory proteins forms a signaling hub responsible for responses to environmental cues such as nutrients and mechanical stress.

mTOR has several phosphorylation sites 

Phosphorylation of Ser2448 is mediated by S6K1 to reduce mTOR activity in a negative feedback loop <!--[@RN1902]-->. 

<!--Ser2448 is phosphorylated by S6K1, changes in nutrient avaliability modifies S6K1 and Ser2448, Ser2448 phosphorylation is abolished when S6K1 is depleted [@RN1902]-->

<!--
When the C-terminal is deleted, mTOR gets constitutively active




## Ribsome biogenesisand muscle growth

<!-- Stec 2016 growth factor induced hypertrophy is abolished when ribo biog is inhibited with the specific CX-5461  -->
<!-- Crossland 2017 in contrasts, IGF-1 induced growth was not impaired when ribo biog was inhibited by CX-5461 -->
<!--Crossland et al. noted that using IGF-1 induces specific growth related effects in muscle cells. These may be different than the effects induced by sreum as in Stec et al. 

 Studies showing correlation between growth and ribo biogenesis -->


## Effects of exercise volume on molecular determinants of muscle growth


[@RN2716]<!-- Kumar 2012  -->

[@RN2720]<!-- Kumar 2009 increasing exercise intensities maximises post ex prot synthesis at 75% in young (plateau at 60-90% of 1RM)  -->




### From Training response

In humans, the biological adaptation to resistance training varies with exercise-training variables such as volume, intensity, rest between repetitions and sets, selection and order of exercises, repetition velocity and frequency of training sessions [@RN1].
In addition, genetic and epigenetic disposition and environmental factors play a role in variations in adaptations
[@RN758;<!-- Timmons 2011 reports gene polymorphism studies -->
@RN2012;<!-- Seaborn 2018 methylation leads to increased gene expression relates to lean mass increase with exercise-->
@RN2199]<!-- Morton 2018 Systematic review of protein supplementation i.e. protein availability as environmental factors -->.
As time constraints often hinder participation in exercise training-programs [@RN2063], numerous studies have searched for the minimal required exercise dose to promote beneficial adaptations. Within-session volume has received particular attention, and although a handful of studies have shown that low-volume training provides similar gains in strength and muscular mass as moderate-volume training
[@RN1382;<!-- Cannon et al 2010 No difference between volume conditions in strength or hypert in women  -->
@RN1454;<!-- Ostrowski no diff between volumes in strength and hypert. in men -->
@RN834]<!-- Mitchell no stat different between volume conditions -->, meta-analyses conclude in favor of moderate volume protocols [@RN793; @RN789; @RN1486; @RN1767].
This apparent discrepancy of specific studies to demonstrate benefits of increased training volume is likely due to a combination of small sample sizes and substantial variation in training responses between individuals and experimental groups. In theory, within-participant designs should alleviate these limitations.


## From RNA-seq

Skeletal muscle is a highly adaptable tissue that responds to environmental stress by altering growth rates and differentiation processes. During resistance training, signaling cascades that stimulate muscle plasticity are triggered. Upon repeated exposures, this facilitates growth and a phenotypic shift in a metabolically active direction [@RN2364],<!-- Egan et al -->
with the opposite happening during inactivity [@RN2365]<!-- Dirks 2016, one week of bed rest muscle loss reduxed ox-capacity and whole body insuline resistance -->.
Despite this generalized view, muscle responsiveness and plasticity vary, both in response to different resistance-training protocols [@RN2358]<!-- Hammarstrom 2020 --> and, perhaps more importantly, between individuals [@RN764; ><!-- Hubal 2005 n=585, large variability reported -->
@RN1741]<!-- Ahtiainen 2016 Heterogenity --->. 
Selected individuals show a near-complete absence of muscle growth after prolonged resistance training, which markedly reduces the beneficial outcomes of such interventions for muscle function and overall health [@RN1741; <!-- Ahtiainen 2016 Heterogenity --->
@RN764]<!-- Hubal 2005 n=585, large variability reported -->.
Currently, little is known about the etiology of this variation. However, it is usually associated with phenotypic traits of skeletal muscle [@RN765;<!-- Bamann cluster analysis showes different expression profiles of e.g. MGF between resp groups -->
@RN1755;  <!-- Stec et al, Ribosome biogenesis different between responders and nonresponders -->
@RN755 <!-- Davidsen 2011 miRNA in response to RT different between responder groups --> 
], which implies interactions with environmental factors, genetics, epigenetics, and composites of the intra physiological milieu [@RN2199;<!-- Morton systematic review showing extra benefit of protein on muscle mass--> 
@RN1966].<!-- Brook 2016 rev. Anabolic resistance in age, nutritional and training variations to potimize outcomes -->
This multifaceted origin makes the training-response-spectrum difficult to study directly, with each of the underlying factors offering limited explanatory value alone [@RN758].<!-- Timmons 2011 Review critique of "single gene master regulator" -->
Instead, a more indirect approach is necessary, whereby the combined effects of the factors are targeted by studying global patterns of mRNA, protein expression, and skeletal muscle biology.

Previous studies have investigated transcriptome responses to acute resistance exercise [@RN1825;<!-- Gordon 2012 Diff expr after acute exercise in trained vs. untrained (identifies ECM as higher in trained)-->
@RN2298;<!-- Damas reghulation in rested state pre post 10 w training and pre post ex before and after training n = 9--> @RN2408]<!-- Hyldahl 2011 acute activation in reponse to 100 eccentric contractions-->
and chronic resistance training [@RN2400;<!-- Robinson 2017 differences on common regulation between exercise types in you and old--> 
@RN2397;<!-- Melow 2007 Age was associated with DE of 596 genes, training in aged led to RNA profiles more like young. --> 
@RN2298;<!-- Damas reghulation in rested state pre post 10 w training and pre post ex before and after training n = 9--> @RN2299;<!-- Murton Exercise induced genes with no response in control group-->
@RN1825;<!-- Gordon 2012 Diff expr after acute exercise in trained vs. untrained (identifies ECM as higher in trained)-->
@RN774]<!-- Raue 2012. Differential expression  -->,
as well as described associations between transcriptome characteristics and degrees of muscle growth [@RN774;<!--Raue 2012. 661 genes regulated by exercise and correlated with muscle size and strength--> 
@RN753],<!-- Phillips 2013 association with  -->
and function [@RN2402;<!-- Hangelbroek 2018 gene cluster predicts 1RM, training induces expression of ECM/collagen -->
@RN2398].<!-- Stepto 2009 Difference between trained (strength and endurance) and untrained controls. A gene clsuter correlated with strength -->
Whereas these studies have merited interesting findings, they lack clear coherences in terms of differential expression events, even for classical exercise-inducible genes such as PGC1$\alpha$ [@RN2360].<!-- Pillon et al -->
This lack of clear coherence is potentially due to a combination of issues such as differences in study design and methods for synthesis and analysis of transcriptome data. First, biologically founded variability can be attributed to differences in exercise protocols (e.g., differences in exercise-volume or intensity). This makes it difficult to discern a general transcriptome exercise response, as training variables are not standardized between studies. Biological heterogeneity is also caused by differences between research participants, affecting signal-to-noise ratios and making it difficult to discern the effects of single independent factors such as training variables. Design stage decisions such as the use of within-participant designs [@RN2358;<!-- Hammarstrom 2020 -->
@RN2403]<!-- Tarnopolsky no diff (only 0.6% differentially expressed) between legs in rested state (small n=8) --> are likely to reduce this variation and to provide transcriptome data with increased biological meaningfulness. Second, technical variability can be attributed to decisions made during the bioinformatical treatment of data. As described by Concea *et al.*, there is no optimal pipeline for sequencing technology as new tools keep evolving and emerging, different tools should be explored to an optimum pipeline for the specific type of data [@RN2426].<!-- Conessa 2016 best practices fom RNA seq-->
To exploit the potential of any study design, there is a need for identifying an appropriate pipeline for transcriptome analyses to ensure a biologically valid interpretation of data. This entails identifying potential violations of common assumptions caused by the experimental model at hand, relating to, for example, data normalization [@RN2414;<!-- Robinson 2010 TMM method normalization assumes most genes not affected by treatment-->
@RN2363]<!-- Dillies et al -->. 

For transcriptome data to provide adequate biological information about a given experimental set-up, numerous bioinformatic steps need to be adopted in a customized manner [@RN2426;<!-- Conessa 2016 best practices fom RNA seq-->
@RN2757].<!-- Arora 2020 differences between pipelines in gene abundance estimates --> 
Of these steps, data normalization is particularly decisive [@RN2363]<!-- Dillies et al -->,
as it aims to transform naïve transcript counts into biologically meaningful results. This essentially means expressing them as *per*-cell abundances [@RN2359]<!-- Loven 2012 et al  -->.
For most experimental models, this is equivalent to providing transcript-to-total RNA ratios, given the fulfillment of the assumption that total RNA levels remain stable between conditions on a per-unit-cell or per-unit-tissue basis [@RN2359]<!-- Loven 2012 et al  -->.
In cell models that exhibit high degrees of plasticity, gene expression events result in increased amounts of total RNA and mRNA transcripts per cell [@RN2430]<!-- Lin 2012 c-Myc induced gene amplification through -->, specifically violating the assumption that most genes are not differentially expressed [@RN2359;<!-- Loven 2012 et al  -->
@RN2414].<!-- Robinson 2010 TMM method normalization assumes most genes not affected by treatment-->
We are not aware of any study that has addressed the need to account for such perspectives during transcriptome analyses of skeletal muscle subjected to mechanical stress, such as resistance training. Indeed, this assumption can be expected to be violated, as total RNA content increases markedly on a per-unit-weight basis [@RN2358]<!-- Hammarstrom 2020 -->, with potential global changes also occurring for the mRNA pool, though this remains unknown. The extent to which total RNA, and therefore ribosomal RNA, increases, coincides with the increase in muscle mass [@RN2358<!-- Hammarstrom 2020 -->;
@RN1755]<!-- Stec -->, underlining its importance for cellular growth but also its inevitable presence as a potential confounding factor in RNA sequencing experiments.

## From tr10

Skeletal muscle is a critical target for interventions that promotes health across the lifespan [@RN763]<!-- Wolfe 2006 -->,
with resistance training (RT) being the advocated remedy.  
Prolonged RT leads to changes in the balance between muscle protein breakdown and synthesis, with one bout of resistance exercise acutely increasing protein synthesis for up to 48 hrs after exercise [@RN786],<!-- Phillips 1997 -->
with subsequent repeated bouts leading to accumulation of muscle protein over time
[@RN1272<!-- Phillips review 2014 -->;
@RN1642]<!-- Brook 2015 experiment -->. 




