
# Background
```{r Backgroundsettings, echo = FALSE, warning=FALSE, message=FALSE}
source("./R/libraries.R"); source("./R/themes.R")



```
## Exercise training variables affecting training outcomes

## Exercise volume

### Meta-analysis of exercise volume 

## Molecular determinants of training-induced muscle hypertrophy

### Protein synthesis

A single exercise session promotes protein synthesis <!-- Kumar 2009 10.1152/japplphysiol.91481.2008 -->
<!-- Philips 1997? -->
Positive net protein balance in responce to exercise <!-- Tipton 10.1152/ajpendo.00234.2002.-->

<!-- Coordinated mTOR 10.1038/nature13492. -->

Inhibition of RNA synthesis restrict protein synthesis <!-- Goldspink 1977 -->

Indicated in Goldspink 1977 and 1976 RNA reflects ribosomal availability <!-- Goldspink 1977-->

Protein synthesis is proportional to RNA content <!-- Millward 1973 -->


Increase loading leads to increased RNA <!-- Goldberg 1975 Mechanism of work-induced hypertrophy of skeletal muscle --> 

```{r Millward1973, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2.5, fig.width = 2.5, fig.cap="Data from Millward et al. 1973. Group A were fed a diet containing protein, group B were starved or fed a diet not containing protein."}

millward <- read_excel("./data/litrev.xlsx", sheet = "millward1973") 

millward %>%
        mutate(group = if_else(group == "A", "Protein fed", "Protein starved")) %>%
        ggplot(aes(RNA, proteinSynthesis, fill = group, color = group)) +
        geom_smooth(method = "lm", se = FALSE) + 
        geom_point(color = "black", shape = 21) +
        scale_x_continuous(name = TeX("$RNA\\;(\\mu g \\; mg^{-1} \\, protein)$"), 
                           breaks = c(0, 4, 8, 12, 16),
                           minor_breaks = c(2, 6, 10, 14),
                           limits = c(0, 16), 
                           expand = c(0, 0)) +
        scale_y_continuous(name = TeX("$Protein\\,synthesis\\,rate\\;(days^{-1})"), 
                           breaks = c(0, 0.05, 0.1, 0.15, 0.20, 0.25),
                           minor_breaks = c(0.02, 0.06, 0.10, 0.14, 0.18, 0.22),
                           limits = c(0, 0.25), 
                           expand = c(0, 0)) + 
        

        dissertation_theme() +
                theme(legend.position = c(0.2, 0.8), 
                      legend.title = element_blank()) 

        



```

### The mammalian target of rapamycin (mTOR) and translational efficiency
The mammalian target of rapamycin (mTOR) is a large serine-threonine protein kinase wich in complex with oyher regulatory proteins forms a signaling hub responsible for responses to environmental cues such as nutrients and mechanical stress.

mTOR has several phosphorylation sites 

Phosphorylation of Ser2448 is mediated by S6K1 to reduce mTOR activity in a negative feedback loop <!--[@RN1902]-->. 

Ser2448 is phosphorylated by S6K1, changes in nutrient avaliability modifies S6K1 and Ser2448, Ser2448 phosphorylation is abolished when S6K1 is depleted <!--[@RN1902]-->


When the C-terminal is deleted, mTOR gets constitutively active



<<<<<<< HEAD
## Ribsome biogenesisand muscle growth

<!-- Stec 2016 growth factor induced hypertrophy is abolished when ribo biog is inhibited with the specific CX-5461  -->
<!-- Crossland 2017 in contrasts, IGF-1 induced growth was not impaired when ribo biog was inhibited by CX-5461 -->
Crossland et al. noted that using IGF-1 induces specific growth related effects in muscle cells. These may be different than the effects induced by sreum as in Stec et al. 

<!-- Studies showing correlation between growth and ribo biogenesis -->

Brook 2016 
Stec 2016
Nakada 2016
Figueriedo 2015

<!-- Older studies -->
Millward 1973 correlation between RNA and protein synthesis


=======
>>>>>>> 03261371c0ce23074e5f42ac588b7606632afb28






<<<<<<< HEAD
=======
### Ribsome biogenesis 
>>>>>>> 03261371c0ce23074e5f42ac588b7606632afb28

#### Transcription of ribsomal RNA (rRNA)

## Transcriptional activity related to muscle hypertrophy

### Methods for studying transcriptional regulation