---
output: html_document
bibliography: bib/complete-bib.bib
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Methods

```{r libs, echo=FALSE, warning=FALSE, message=FALSE}
source("./R/libraries.R")
source("./R/themes.R")
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Study participants, protocols and training interventions

Study I was designed to examine effects of low- and moderate-volume on
responses to acute exercise and long-term training within participants.
Forty-one healthy individuals were recruited and 34 of these completed
at least 85% of the prescribed sessions and were thus included in
subsequent data analyses. Reasons for not completing the trial included
injury not related to the study (*n =* 1), pain or discomfort during
exercises (*n =* 5) and non-adherence to the study protocol. There were
no differences in characteristics between participants included in or
excluded from data analysis in Study I.

Study II was designed to study the effects of resistance training *per
se*, and effects of variable volume on selected markers related to
ribosome biogenesis. Participants were therefore recruited to a training
group (*n =* 11) and a non-training control group (*n =* 8). Eligible
for participation in both studies were young (Study I age 18-40; Study
II 18-35), non-smoking men and women. Exclusion criteria included a
training history of more than one weekly session during the last 12
(Study I) or six (Study II) months leading up to the study. Participants
were also screened for intolerance to local anesthetic, current or
previous injuries affecting their ability to perform resistance
training, self-reported symptoms or history of disease, intake of
medication or supplements with known effects on adaptations to training.
Participant characteristics for both studies are shown in Table
\@ref(tab:characteristics-table).

```{r characteristics-table, results = "asis", message=FALSE, warning=FALSE, echo=FALSE}

### Read data on included/excluded, body composition and cybex data
characteristics <- read_excel("./data/study-1/body-composition/bodycomp_DXA.xlsx") 
participants <- read.csv("./data/study-1/body-composition/oneThreeSetLeg.csv", sep=";")
load("./data/study-1/strength/cybex.rda")

library(knitr); library(kableExtra)

# Study 1 characteristics
cybex.temp <- data.frame(cybex) %>%
  filter(timepoint %in% c("pre", "session1"), type=="Isom")%>%
  group_by(subject)%>%
  summarise(trq=max(torque.max))

studyI <- participants %>%
  inner_join(characteristics) %>%
  filter(subject!="FP10") %>% # removes participant excluded prior to comecing the study
  filter(timepoint=="pre") %>%
  inner_join(cybex.temp) %>%
  dplyr::select(subject, sex, include, age, height, weight, fat.whole, lean.whole,BMC.whole, trq)%>%
  mutate(fat=((fat.whole/1000)/weight)*100,
         trq.lean=trq/(lean.whole/1000)) %>%
  inner_join(participants) %>%
        mutate(group = if_else(include == "incl", "Included", "Excluded"), 
               sex = factor(sex, levels = c("female", "male"), 
                            labels = c("Female", "Male")), 
               study = "Study I", 
               bmi = weight/(height/100)^2) %>%
        dplyr::select(participant = subject, 
                      study,
                      sex,group, age, height, weight,bmi,
                      fat.whole, 
                      lean.whole,
                      BMC.whole) 

# 
# age.m1 <- lme(age ~ include + sex, random = list(subject = ~1), data = characteristicsStats)
# weight <- lme(weight ~ include + sex, random = list(subject = ~1), data = characteristicsStats)
# height <- lme(height ~ include + sex, random = list(subject = ~1), data = characteristicsStats)
# fat    <- lme(fat ~ include + sex, random = list(subject = ~1), data = characteristicsStats)
# mvc    <- lme(trq.lean ~ include + sex, random = list(subject = ~1), data = characteristicsStats)
# 
# summary(age.m1)

# summary(weight)

# summary(height)

# summary(fat)

# summary(mvc)
  
### No differences between included not included.

# Study 2 characteristics 



dxa_whole <- read_excel("./data/study-2/body-composition/tr010_dxa_standarddefinitions.xlsx") %>%
  dplyr::select(Etternavn, Kjønn,Målingsdato,Alder, Vekt:`HeleVev%fett`) %>%
  pivot_longer(names_to = "variable", values_to = "value", cols = Vekt:`HeleVev%fett`) %>%
  mutate(value = gsub("kg", "", value), 
         value = gsub("[[:space:]]", "", value, fixed = FALSE), 
         value = as.numeric(value)) %>%

  separate(Etternavn, into = c("Study", "participant")) %>%
  dplyr::select(participant, Date = Målingsdato, variable, value) %>%
  filter(!(participant %in% c("P08", "P20", "P17"))) %>%
  group_by(participant) %>%
  mutate(time = if_else(Date == min(Date), "S0", 
                        if_else(Date == max(Date), "post1w", 
                                "S12")), 
         leg = if_else(variable == "BeinhøyreMagermasse", "R", "L")) %>%
  filter(variable %in% c("HeleBeinmasse", 
                         "HeleFettmasse", 
                         "HeleVevmasse", 
                         "HeleMagermasse" ,
                         "Vekt", 
                         "HeleTotalmasse")) %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  mutate(fat_rel = 100*(HeleFettmasse/HeleTotalmasse)) %>%
          filter(time == "S0") 




studyII <- read_excel("./data/study-2/body-composition/tr010_age_participant.xlsx") %>%
  inner_join(dxa_whole) %>%
  mutate(age = as.numeric(difftime(recruited, dob)/365.2425), 
         bmi = weight/(height/100)^2, 
         study = "Study II") %>%
          dplyr::select(participant, 
                        study,
                        sex,
                        group,
                        age,
                        height, 
                weight = Vekt,
                bmi,

                fat.whole = HeleFettmasse,
                lean.whole = HeleMagermasse,
                BMC.whole = HeleBeinmasse) %>%
        mutate(sex = factor(sex, levels = c("F", "M"), 
                            labels = c("Female", "Male"))) 
        

# Combine both studies and make table
rbind(studyI, studyII) %>%

        mutate(rel_lean = ((lean.whole/1000) / weight) * 100, 
               rel_fat =  ((fat.whole/1000) / weight) * 100) %>%
        dplyr::select(participant:bmi, rel_lean:rel_fat) %>%
        pivot_longer(names_to = "variable", values_to = "values", cols = age:rel_fat ) %>%

        
        group_by(study, group, sex, variable) %>%
        summarise(m = mean(values), 
                  sd = sd(values)) %>% 
        ungroup() %>%
        mutate(m.sd = if_else(variable == "age", paste0(sprintf("%.1f", m), " (", sprintf("%.1f", sd), ")"), 
                              if_else(variable == "bmi", paste0(sprintf("%.1f", m), " (", sprintf("%.1f", sd), ")"), 
                                      if_else(variable == "height", paste0(sprintf("%.0f", m), " (", sprintf("%.0f", sd), ")"), 
                                              if_else(variable == "rel_fat", paste0(sprintf("%.1f", m), " (", sprintf("%.1f", sd), ")"), 
                                                      if_else(variable == "rel_lean", paste0(sprintf("%.1f", m), " (", sprintf("%.1f", sd), ")"), 
                                                              paste0(sprintf("%.1f", m), " (", sprintf("%.1f", sd), ")")))))),
               group = factor(group, levels = c("Included", "Excluded", "experiment", "control"), 
                              labels = c("Included", "Excluded", "Training", "Control"))) %>%
        dplyr::select(study:variable, m.sd) %>%
        pivot_wider(names_from = variable, values_from = m.sd) %>%
        arrange(study, group) %>%
        dplyr::select(study, group, sex, age, height, weight, rel_fat, rel_lean) %>%

  kable(format = "latex", 
        booktabs = TRUE, 
        caption = "Participant characteristics", 
        escape = FALSE, 
        col.names = c(" ", " ", "Sex", "Age (years)", "Stature\n(cm)", "Mass (kg)", "Fat mass (\\%)", "Lean mass (\\%)")) %>%
        
        kable_styling(font_size = 7) %>%
        collapse_rows(1:2) %>%

  footnote(general = "Data are means and (SD)",
           general_title = "", 
           footnote_as_chunk = TRUE)


```

## Resistance training interventions

Each training session started with a light standardized warm-up (5 min
ergometer cycling and 10 repetitions each of push-ups, sit-ups,
back-extensions and squats). Before each exercise in the main program,
one set of 10 repetitions were performed in the specific exercise with
approximately 50% of 1RM.

Studies were fully or partially performed as within-participant studies
as each participant had their legs assigned to different training
conditions (not including the control group in Study II). Allocation was
performed after enrollment where each participant had their legs
randomized to either low- or moderate volume (Study I), or variable or
constant volume (Study II).

In Study I, the low-volume protocol consisted of a single set of each
exercise and the moderate-volume consisted of three sets per exercise.
Three unilateral leg exercises were used (leg press, leg curl and knee
extension). The moderate volume-leg commenced all sessions and the low
volume-leg performed a single set of each exercise in the rest between
second and third set of the moderate volume training protocol.

In Study II, only unilateral knee-extension was performed in an effort
to concentrate the stimulus to the quadriceps
muscles.<!-- Check this --> The constant-volume leg performed six sets
of 10RM throughout the study and variable leg performed six sets in
session one to four, three sets in session five to eight and nine sets
in session nine to twelve with same intensity (10RM).

### Ethical considerations

Both studies were approved by the local ethics committee Lillehammer
University College/Inland Norway University of Applied Sciences and the
Norwegian Centre for Research Data. In accordance with the *Declaration
of Helsinki*[@RN2548] the studies were pre-registered in publicly
accessible databases (Study I, ClinicalTrials.gov Identifier:
NCT02179307; Study II, <https://osf.io/wa96y>). Participants were
informed of the study design, potential risks and sources of discomfort
prior to giving their informed consent.

 ## Measures of muscle mass

In Study I muscle mass was measured by magnetic resonance imaging (MRI)
and dual energy X-ray absorptiometry (DXA) prior to and after the
intervention. Both MRI and DXA measurements were completed during the
same visit to the laboratory. Participants were instructed to refrain
from strenuous physical activity during the last 48 h leading up to the
measurements. The post-training measurements were completed at least 48
h after the last strength testing session. Participants were asked to
refrain from food consumption during 2 h leading up to the measurements.

MRI images were obtained from the mid-thigh and analyzed by the same
investigator blinded for time (pre- and post-training) and condition
(low- and moderate-volume). Multiple images were used to estimate the
cross-sectional area of the extensor muscles at the same distance from
the knee-joint.

See figure <!-- @ref(fig:body-comp-measures)-->

<!--Dallin et al. recently estimated the [@RN2541]
     MRI has shown better agreement with cadaver... see ref in Dallin 

    ## Muscle strength assessments

    Muscle strength was with  -->

In Study II


## Muscle tissue sampling and preparations for downstream analyses

Muscle samples were obtained under local anesthesia (Study I, Xylocaine,
\SI{10}{\mg\per\ml} with adrenalin \SI{5}{\micro\gram\per\ml},
AstraZeneca, Oslo, Norway; Study II, Lidocaine Mylan,
\SI{10}{\mg\per\ml}, Mylan Ireland Ltd, Ireland) with a fine needle
(12-14 gauge; Universal-plus, Medax, Italy) operated with a
spring-loaded instrument (Bard Magnum, Bard Norway AS, Norway). Sampling
was performed as previously described
[@RN824],<!-- Hayot 2005 relible micriboiopsy technique --> with
modifications. Anesthesia was injected in the subcutaneous tissue with
care taken not to inject anesthesia into the muscle itself. Following
pilot experiments we decided not to use an insertion cannula as
described in [@RN824] as the biopsy needle itself could be used to
puncture the skin and muscle fascia. This also resulted in less
discomfort. Several passes through the same skin puncture was made to
obtain sufficient material for downstream analyses. A smaller needle (14
vs. 12 gauge) was used to further minimize discomfort in Study II where
more biopsies were sampled over a shorter time span, with exception from
when material was used for immunohistochemistry. The first biopsy was
sampled at one third of the distance between the patella to the
*anterior superior iliac spinae* with subsequent biopsies sampled
$\sim$\SI{2}{cm} proximal to previous samples. In Study II, samples
obtained more than one week apart were sampled with closer proximity and
distally from previous samples but never at previous sampling sites.

The micro biopsy technique produces smaller samples compared to other
biopsy techniques
[@RN2549]<!--Ekblom describes the Bergström and conchotome -->, and thus
requires several passes to produce sufficient material for multiple
downstream experiments. However, reports confirms that the micro biopsy
technique is comparable to the traditionally used Bergström technique in
several measures of muscle characteristics at the same time as being
well tolerated [@RN824; @RN2553]. Any reported differences in fiber type
distributions between sampling techniques have been suggested relating
to differences in sampling depth [@RN2553; @RN2552].<!-- Hayot -->

For determination of fiber type distributions, a threshold of 200-300
fibers has been suggested as a suitable sample size per specimen as more
fibers does not reduce the variation between dupliacte samples
[@RN874].<!--Blomstrand suggests duplicate samples and at least 200 fibres for determination of fiber distributions -->
In Study I, one or several pieces of muscle (total weight
$\sim$\SI{15}{mg}) were chosen per sampling for analysis of fiber type
distributions (described in detail below). The total number of fibers
were counted from these specimens (Figure ref fig). Using an average of
fibers from the first sampling time point the between leg coefficient of
variation was determined to 14% for Type I fibers and 11.3 for type II
fibers. The between leg variation in Type I fibers is similar to what
has been previously reported...

```{=html}
<!-- Latest paper on variability between samples from the same leg and due to number of fibers counted
Appl Physiol Nutr Metab
. 2020 Apr;45(4):368-375. doi: 10.1139/apnm-2019-0263. Epub 2020 Mar 24.

The within-leg was similar to between leg comparison in our samples. This might highlight sampling depth variation in microbiopsy technique in immunohistochemistry ?

To calculate variation in proportions

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3875499/



```


## Gene expression analysis

### Total RNA extraction
```{r rna-muscle-weight}

# Descriptive data on total RNA extractions from both studies

study1_sample <- read_excel("./data/study-1/sample_weight.xlsx") %>%
  mutate(rna = (conc1.5 * 5) * elution.volume, 
         sample = paste(subject, leg, timepoint, sep = "_")) %>%
  dplyr::select(sample, weight, rna) %>%
    mutate(study = "Study1") 


complete_sample <- readRDS("./data/derivedData/study-2/tot-rna.RDS") %>%
  filter(outlier != "out") %>%
  mutate(sample = paste(participant, leg, time, sep = "_")) %>%
  group_by(sample) %>%
  summarise(weight = mean(tissue_weight, na.rm = TRUE), 
            rna = mean(rna, na.rm = TRUE)) %>%
  mutate(study = "Study2") %>%
  rbind(study1_sample) 

## Calculate mean weights and rna per weight ##

desc_to_rna <- complete_sample %>%
  group_by(study) %>%
  summarise(m.weight = mean(weight, na.rm = TRUE), 
            s.weight = sd(weight, na.rm = TRUE), 
            m.rna_weight = mean(rna/weight, na.rm = TRUE), 
            s.rna_weight = sd(rna/weight, na.rm = TRUE)) 




## Figure on RNA descriptives

rna_fig_a <- complete_sample %>%
  ggplot(aes(weight, rna, fill = study)) + 
  geom_point(shape = 21, alpha = 0.2) +
    scale_x_continuous(limits = c(0, 80), 
                     expand = c(0,0), 
                     breaks = c(0, 20, 40, 60, 80)) +
      scale_y_continuous(limits = c(0, 25000), 
                     expand = c(0,0), 
                     breaks = c(0, 5000, 10000, 15000, 20000, 25000)) +
  dissertation_theme() + 
  theme(axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.line.x = element_blank(), 
        axis.ticks.x = element_blank(), 
        legend.position = "none") + 
  labs(y = "RNA (ng)")


rna_fig_b <- complete_sample %>%
  ggplot(aes(weight, fill = study)) + geom_histogram() +
  facet_grid(study ~ .) +
  dissertation_theme() + 
  scale_x_continuous(limits = c(0, 80), 
                     expand = c(0,0), 
                     breaks = c(0, 20, 40, 60, 80)) +
  theme(legend.position = "none", 
        strip.background = element_blank(),
        strip.text = element_blank()) +

  labs(y = "Counts", 
       x = "Weight (mg)")

# plot_grid(rna_fig_a, rna_fig_b,
#           ncol = 1,
#          rel_heights = c(1, 0.75), 
#          align = "v", 
#          axis = "trbl")
# 


### Calculate RNA per tissue at baseline
# 
# complete_sample %>%
#   separate(sample, into = c("subject", "leg", "timepoint")) %>%
#   filter(timepoint %in% c("S0", "w0")) %>%
#   mutate(rna_corrected = if_else(study == "Study1", rna * 1.125, rna)) %>%
#   lmer(rna ~ study + weight + (1|subject),
#       
#       data = .) %>%
#   summary()
#   
#   ggplot(aes(rna_corrected / weight, fill = study)) + geom_histogram()
#   
#   
#   group_by(study) %>%
#   summarise(rna_weight = mean(rna/weight), 
#             rna_corrected = mean(rna_corrected/weight)) %>%
#   print()
# 
# 
# 

```



Total-RNA was extracted from pre-weighed, frozen muscle samples with a protocol modified from
[@RN2672]<!-- Chomczynski and Sacchi -->
using Trizol reagent (Life Technologies).
Muscle tissue was homogenized in \SI{300}{ul} Trizol with mechanical disruption achieved by Zirconium Oxide Beads (0.5 mm, Next Advance, Inc., New York, USA) and a bead mill (Bullet blender, Next Advance). External, non-mammalian RNA (Lambda PolyA External Standard Kit, Takara Bio Europe, Saint-Germain-en-Laye, France) was added with the initial volume of Trizol to enable per-weight normalization in subsequent analyses. After homogenization, additional Trizol was added to a total volume of \SI{1}{ml}. Phase separation was achieved by centrifugation after addition of chloroform (\SI{200}{ul}). The upper phase (\SI{400}{ul} in Study I; \SI{450}{ul} in Study II) was transferred to a fresh tube and RNA was precipitated using isopropanol (\SI{500}{ul}). After incubation (10 min, room-temperature) and centrifugation (12000 g, 10 min at 4$^{\circ}$C) the resulting RNA pellet was washed three times in chilled 75% ethanol.

As previously mentioned, to minimize discomfort with a larger number of biopsies sampled over a short time, a smaller needle was used for most biopsies in Study II. This generally led to a less tissue used for RNA extractions (Figure XX).  



## Determination of protein abundance

## Statistics and data analysis

TO DO:

-   For methods discussion, compare product length, efficiencies and ct
    values in relation to RQI-values. See Fleige 2006 for reference.


### Normalization

-   An external reference gene was added at a constant amount in Trizol
    preps
-   A normalization factor was used to express relative target gene
    abundance per-weight tissue.
-   In qPCR the linearised expression (effectivety \^cq) was used to
    express the fraction of external reference per total RNA.
-   In RNA-seq the external reference gene was sequenced and counts were
    used to express external RNA as a fraction of total RNA.
-   In both cases the normalization factor was calculated as mw \*
    counts.

A simulation to see that this is equivalent to tissue used in prep when
no measurement errors exists.

```{r}

library(tidyverse)

expand_grid(mg = seq(from = 5, to = 100, by = 5), 
            rna.mg = seq(from = 250, to = 600, by = 25),
            ext = 0.04) %>%
  mutate(tot.rna = mg * rna.mg, 
         ext.frac = ext / (ext + tot.rna), 
         mg.inprep = 1000 / ((ext + tot.rna) / mg), 
         nf = ext.frac * mg) 



expand_grid(mg = seq(from = 5, to = 100, by = 5), 
            rna.mg = seq(from = 50, to = 600, by = 25),
            ext = 0.04) %>%
  mutate(tot.rna = mg * rna.mg, 
         ext.frac = ext / (ext + tot.rna), 
         mg.inprep = 1000 / ((ext + tot.rna) / mg), 
         nf = ext.frac * mg) %>%
         ggplot(aes(rna.mg, mg.inprep,  color = mg)) + geom_point(size = 2)




```

## Meta-analysis of within-session training volume

### Literature search, inclusion criteria and coding of studies

A first set of studies were coded based previously published
meta-analyses [@RN1767; @RN789]. For more recent studies, PubMed, Google
Scholar and SportDiscuss searches were made with search terms being
"training volume", "resistance training", "strength training", "set",
"muscle strength", "muscle hypertrophy" used in different combinations.
Studies examining the effect of within-session training volume on muscle
strength and mass, with all other training variables kept constant
between study groups were considered for inclusion in the meta-analysis.
Studies were further assessed for inclusion based on criteria being; (i)
participants described as healthy without medications affecting muscle
metabolism, (ii) interventions lasting at least 6 weeks and (iii) RT
performed without additional stimuli (e.g. blood flow restriction) at
intensities above 65% of 1RM or 20RM.

All available outcome measures of muscle mass and strength gains in
response to RT were extracted from each study with exception of outcomes
reported both as summaries and individual measures (e.g. muscle
thickness reported as individual muscles and summarized for the whole
muscle group). In such cases the summary was used as outcome. Weekly
training volume was calculated as product of weekly sessions, number of
sets and exercises for each muscle group assessed for muscle hypertrophy
or strength gains. An intervention average of weekly sessions was used
when the number of sessions per week differed over the course of the
intervention. An exercise was assumed to influence an outcome when it
targeted prime movers also assessed for strength or muscle hypertrophy
measures. Participant characteristics were coded based on sex (male,
female or mixed when a study failed to discern between male and
females), age (young, middle-aged, old or mixed), body-mass index (BMI,
calculated from average body mass and height when BMI values were not
available), training status (trained, \> 1 session per week during the
last 6 months leading up to the intervention; untrained \< 1 session).
Study groups were considered independent also in studies utilizing
within-participant models.

### Calculations of effect sizes and statistical analysis

Group-wise effect sizes were calculated for each outcome measure based
on the within-group change score pre- to post-training divided by the
pre-training standard deviation (SD). Pre-training SD's were calculated
as a pooled SD within outcome and study. Variances of the effect size
were calculated using an average effect size across all outcomes within
muscle strength or mass, and correlations specific to each measurement
type (isokinetic-, isometric- or repetition maximum strength tests;
muscle thickness, magnetic resonance imaging, dual energy X‐ray
absorptiometry) estimated from previous studies.\
A correction factor () was applied to both effect sizes and their
variances.

Mixed-effects meta-regression models were used to model the effect of
weekly number of sets on RT-induced muscle mass and strength gains.
Models were fitted in a Bayesian framework using the brms-package
[@RN2562].
