# Results and Discussion

```{r settings, include=FALSE}

library(nlme); library(kableExtra)


```


## Effects of different training volume on changes in muscle size and function

```{r comparing-avg-strength-csa, echo=FALSE, warning=FALSE, message=FALSE}


source("./R/study-1/mr-and-dxa-data.R")


participants <- read.csv("./data/study-1/body-composition/oneThreeSetLeg.csv", sep=";") %>%
        pivot_longer(names_to = "sets", values_to = "leg", cols = multiple:single) 


### Overall % increase in CSA (not corrected for baseline values)
over.all.csa <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(change = ((post/pre)-1)*100) %>%
  group_by(subject) %>%
  summarise(mean.csa = mean(change, na.rm = TRUE)) %>%
  ungroup() %>%
  summarise(m.csa = mean(mean.csa),
            s.csa = sd(mean.csa))

sets.csa <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(change = ((post/pre)-1)*100) %>%
  group_by(sets) %>%
  summarise(m.csa = mean(change),
            s.csa = sd(change))



### Baseline corrected % increase
csa.corrected.data <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(change = ((post/pre)-1)*100) %>% # Calculates % change
  group_by(sex) %>%
  mutate(pre = pre-mean(pre)) %>% # Mean centered baseline 
  # (no need for sex-diff in model)
  group_by(subject, sex) %>% # Keeps sex for test in model
  summarise(mean.csa = mean(change, na.rm = TRUE),
            mean.baseline = mean(pre)) # Average over both conditions 



## Anocova with baseline values as co-variate
csa.corr.m1 <- lm(mean.csa ~  mean.baseline,
                  data = csa.corrected.data)

csa.m1 <- lm(mean.csa ~  1,
                  data = csa.corrected.data)



csa.corr.confint  <- round(confint(csa.corr.m1)[1,],1)
csa.corr.estimate <- round(summary(csa.corr.m1)$coef[1,1],1)




csa <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  group_by(timepoint, sets, sex) %>%
  summarise(m=sprintf("%.1f", round(mean(CSA.avg),1)),
            s=sprintf("%.1f", round(sd(CSA.avg),1)))


csa.r <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(csa.change=((post/pre)-1)*100) %>%
  dplyr::select(subject, sex, sets, csa.change) %>%
  spread(sets, csa.change)

csa.r.model <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject,leg, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(csa.change=((post/pre)-1)*100) %>%
  dplyr::select(subject,leg, sex, sets, pre, csa.change) %>%
  group_by(sex) %>%
  mutate(pre.m = pre-mean(pre, na.rm = TRUE), 
         sets = factor(sets, levels = c("single", "multiple"))) 

#### Change to ancova ###


csa.m0 <-nlme::lme(csa.change ~ sets +  pre.m,
                random = list(subject = ~1),
                na.action = na.omit, 
                data = csa.r.model, method = "ML", 
                control=list(msMaxIter=120,
                             opt = "nloptwrap", msVerbose = FALSE))




csa.estimate <- sprintf("%.2f",round(coef(summary(csa.m0))[2,1], 2))
csa.confint <- intervals(csa.m0)$fixed[2, c(1,3)]

### t-test Does not change the conclusion 
# csa.ttest <- with(csa.r, t.test(multiple,single, paired=T))
# csa.estimate <- sprintf("%.2f",round(csa.ttest$estimate, 2))
# csa.confint <- csa.ttest$conf.int


csa.corr <- cor.test(csa.r$multiple, csa.r$single)


female.t <- with(csa.r[csa.r$sex=="female",], t.test(multiple,single, paired=T))
female.estimate <- sprintf("%.2f",round(female.t$estimate, 2))
female.confint <- female.t$conf.int


male.t <- with(csa.r[csa.r$sex=="male",], t.test(multiple,single, paired=T))
male.estimate <- sprintf("%.2f",round(male.t$estimate, 2))
male.confint <- male.t$conf.int

#### Strength data for overall stats



rm_str <- read_excel("./data/study-1/strength/strengthTests.xlsx") %>%
        filter(!(exercise %in% c("benchpress", "legcurl"))) %>%
        inner_join(participants) %>%
        filter(include == "incl") %>%

  dplyr::select(subject, sex, sets, exercise, timepoint, load)%>%
  mutate(load = as.numeric(load),
         sets = factor(sets, levels = c("single", "multiple"))) %>%
  spread(timepoint, load) %>%
  mutate(baseline = pmax(pre, session1)) %>%
  dplyr::select(subject, sex, sets, exercise, baseline, week2, week5, week9, post) %>%
  gather(week, load, baseline:post) %>%
  mutate(week = factor(week, levels = c("baseline", "week2", "week5", "week9", "week12", "post"))) %>%
  group_by(exercise) %>%
  mutate(str = load / max(load, na.rm = TRUE)) %>%
  dplyr::select(-load) %>%
  filter(week %in% c("baseline", "post")) 

# load cybex data 
load("./data/study-1/strength/cybex.rda")


cybex <- data.frame(cybex) %>%
        mutate(leg = if_else(leg == "Left", "L", "R")) %>%
        inner_join(participants) %>%
        filter(include == "incl", 
               timepoint %in% c("pre","session1", "post")) %>%

        filter(!(subject=="FP6" & timepoint == "post" & speed == 240))  # This removes one record, exclusion due to error in measurement, see protocol FP6


# Clean cybex data set and combine with 1RM 
total_strength <- cybex %>%
  ungroup() %>%
  dplyr::select(subject, timepoint, sets,speed, sex, torque.max) %>%
  spread(timepoint, torque.max) %>%
  rowwise()%>%
  mutate(baseline = max(pre, session1, na.rm=T)) %>%
  ungroup() %>%
  dplyr::select(subject, sex, sets, exercise = speed, baseline,  post) %>%
  gather(week, load, baseline:post) %>%
  group_by(exercise) %>%
  mutate(str = load/max(load, na.rm = TRUE)) %>%
  dplyr::select(-load) %>%
  filter(week %in% c("baseline", "post")) %>%
  rbind(rm_str) %>% # This includes all strength tests
  mutate(ex_fam = if_else(exercise %in% c("legext", "legpress"), 
                          "rm", if_else(exercise == 0, "isom", "isok"))) %>%
  group_by(subject, sex, sets, week, ex_fam) %>%
  summarise(str = mean(str, na.rm = TRUE)) %>%
  group_by(subject, sex, sets, week) %>%
  summarise(str = mean(str, na.rm = TRUE)) 


strength.stats <- total_strength %>%
  spread(week, str) %>%
  mutate(increase = 100 * ((post/baseline)-1)) %>%
  group_by(sex) %>%
  mutate(BASELINE = baseline - mean(baseline)) %>%
    group_by(sex, subject) %>%
    summarise(increase = mean(increase, na.rm = TRUE),
              BASELINE = mean(BASELINE, na.rm = TRUE)) 




## ANCOVA for the strength 1RM
strength.m1 <- lm(increase ~ sex + BASELINE,
                  data = strength.stats)

strength.corr.confint  <- round(confint(strength.m1)[1,],0)
strength.corr.estimate <- round(summary(strength.m1)$coef[1,1],0)



### Correlation in strength gains ###

str.corr.data <-  total_strength %>%
  spread(week, str) %>%
  mutate(increase = 100 * ((post/baseline)-1)) %>%
  dplyr::select(subject, sex, sets, increase) %>%
  spread(sets, increase) 

str.corr <- cor.test(str.corr.data$single,
                          str.corr.data$multiple)

#### Correlation strength and csa increases

# Average increases

csa.str.corr.data <- total_strength %>%
  spread(week, str) %>%
  mutate(str.rel.change = 100 * ((post/baseline)-1)) %>%
  dplyr::select(subject,  sets, sex, sets,baseline, str.rel.change) %>%

  inner_join(mr.results %>%
                mutate(condition=leg) %>%
                inner_join(participants) %>%
                filter(include=="incl") %>%
                dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
                spread(timepoint, CSA.avg) %>%
                mutate(csa.rel.change = 100 * ((post/pre)-1)) %>%
                dplyr::select(subject, sets, csa.rel.change)) %>%
        ungroup() 





# The code below could possibly be used to calculate the relative contribution of 
# csa increase to muscle strength increases


# m <- lme4::lmer(str.rel.change ~ csa.rel.change + sets + (1|subject), 
#                 data = csa.str.corr.data)
# 
# 
# 
# mx.null <- lme4::lmer(str.rel.change ~ 1 + (1|subject), data = csa.str.corr.data)
# mx.csa <- lme4::lmer(str.rel.change ~ csa.rel.change  + (1|subject), data = csa.str.corr.data)
# mx.sets <- lme4::lmer(str.rel.change ~ sets + (1|subject), data = csa.str.corr.data)
# 
# 
# 
# boot.fun <- function(x) {
#         data.frame(nlme::VarCorr(x))[2, 4]
# }
# 
# 
# null.boot <- lme4::bootMer(mx.null, boot.fun, nsim = 500, parallel = "snow", cl = 12)
# csa.boot <- lme4::bootMer(mx.csa, boot.fun, nsim = 500)
# sets.boot <- lme4::bootMer(mx.sets, boot.fun, nsim = 500)
# 
# 
# 
# qt.null <- quantile(null.boot$t, c(0.025, 0.975))
# qt.csa <- quantile(csa.boot$t, c(0.025, 0.975))
# qt.sets <- quantile(sets.boot$t, c(0.025, 0.975))
# 
# 
# results <- data.frame(variable = c("sets", "csa"),
#                            m = c(1-  data.frame(nlme::VarCorr(mx.sets))[2, 4] / # data.frame(nlme::VarCorr(mx.null))[2, 4],
#                                  1-  data.frame(nlme::VarCorr(mx.csa))[2, 4] / # data.frame(nlme::VarCorr(mx.null))[2, 4]),
#                            lwr = c(1- qt.sets[1] / qt.null[1],
#                                    1- qt.csa[1] / qt.null[1]),
#                            upr = c(1- qt.sets[2] / qt.null[2],
#                                    1- qt.csa[2] / qt.null[2]))
# 
# 
# 
# 
# 
# csa <- 1-  data.frame(nlme::VarCorr(mx.csa))[2, 4] / data.frame(nlme::VarCorr(mx.null))[2, 4]
# 
# sets <- 1-  data.frame(nlme::VarCorr(mx.sets))[2, 4] / data.frame(nlme::VarCorr(mx.null))[2, 4]
# 
# exp <- 1 - data.frame(nlme::VarCorr(mx.exp))[2, 4] / data.frame(nlme::VarCorr(mx.null))[2, 4]
# 
# 

# Average csa and strength gains

avg.str.csa <- csa.str.corr.data %>%
        group_by(subject) %>%
        summarise(str = mean(str.rel.change), 
                  csa = mean(csa.rel.change))


csa.str_corr <- cor.test(avg.str.csa$csa, avg.str.csa$str)

## Comparisons with other published studies on per session gains in csa and strength

### Per session ##
per.session <- read_excel("./data/study-1/trainingdata.xlsx", sheet = 1) %>%
  filter(!is.na(date)) %>%
  group_by(subject) %>%
  summarise(sessions = n()) %>%
        
inner_join(csa.str.corr.data) %>%

  mutate(str.change.day = str.rel.change/((sessions/31) * 12*7),
         csa.change.day = csa.rel.change/((sessions/31) * 12*7), 
         str.session = str.rel.change/sessions,
          csa.session = csa.rel.change/sessions ) %>%
   dplyr::select(-baseline) %>%
  ungroup() %>%
        pivot_longer(names_to = "variable", values_to = "gain", cols = str.rel.change:csa.session) 
       

sum.stats <- cbind(per.session %>%
        group_by(sets, variable, sex) %>%
        summarise(mean = mean(gain), 
                  sd = sd(gain)) %>%
        arrange(variable) %>%
        data.frame(), 
        data.frame(reference.mean = c(0.11, 0.11, 0.11, 0.11,
                              NA, NA, NA, NA, 
                              0.0825, 0.1375,0.0825, 0.1375, 
                              NA, NA, NA, NA, 
                               NA, NA, NA, NA,
                              0.6675, 0.475, 0.6675, 0.475), 
           reference.range = c("[0.04-0.26]a", 
                               "[0.04-0.26]a", 
                               "[0.04-0.26]a", 
                               "[0.04-0.26]a", 
                               "", "", "", "",
                               "(0.22)b", "(0.14)b",  "(0.22)b", "(0.14)b", 
                               "", "", "", "",
                              "", "", "", "",
                               "(0.35)b", "(0.22)b","(0.35)b", "(0.22)b" ))) %>%
        mutate(m.sd = if_else(mean > 6, paste0(sprintf("%.1f",mean), " (", sprintf("%.1f",sd), ")"), 
                              paste0(sprintf("%.2f", mean), " (", sprintf("%.2f",sd), ")")), 
               reference = if_else(is.na(reference.mean), "",
                                   paste0(sprintf("%.2f",reference.mean), " ", reference.range))) %>%
        filter(!(variable %in% c("str.change.day") )) %>%
                       dplyr::select(variable, sex, sets, m.sd, reference) %>%
        
        mutate(variable = factor(variable, levels = c("csa.rel.change", 
                                                      "csa.change.day", 
                                                      "csa.session", 
                                                      "str.rel.change", 
                                                      "str.session"), 
                                 labels = c("CSA %-change", 
                                            "CSA %-change per day", 
                                            "CSA %-change per session", 
                                            "Average strength %-change", 
                                            "Average strength %-change per session")), 
               sex = factor(sex, levels = c("female", 
                                            "male"), 
                            labels = c("Female", "Male")), 
               sets = factor(sets, levels = c("single", "multiple"), 
                             labels = c("LOW", "MOD"))) 

## Ahtiaianen on average  20 weeks with 2 sessions per week = 40 sessions

## Muscle hypertrophy
## In the younger cohorts, male average 5.5% (5.6) = 5.5/40 ~ 0.1375 per session
# in females 3.3 (8.6)% = 0.825
# 8.6/3.3 * 3.3/40
# 5.6/5.5 * 5.5/40
# Muscle strength in younger cohorts
#  male    19% (8.6)/40 sessions ~  0.5 % per session
#  female  26.7% (13.9)/40 ~ 
# 8.6/19 * (19/40)
# 13.9/26.7 * 26.7/40

# Wernbom increase per day in quadriceps CSA per day was 0.06%







```


In Study I, the average increases in muscle strength and mass in each volume condition corresponded to what could be expected based on previous research in young healthy participants (Table \@ref(tab:csa-str-tab)) 
[@RN2007;<!-- Wernbom 2007 -->
@RN1741],<!-- Ahtiainen  -->
indicating the general efficiency of the training program.


(ref:wernbom-citation) @RN2007

(ref:ahtiainen-citation) @RN1741

```{r csa-str-tab, warning=FALSE, message=FALSE, echo=FALSE, results = "asis"}

sum.stats %>%
        arrange(variable, sex, sets) %>%
        kable(format = "latex", booktabs = TRUE, linesep = "",
              col.names = c("", "Sex", "Volume condition", "Mean (SD)", "Reference"), 
              caption = "Training induced changes in muscle CSA and average strength in Study I")  %>%
        collapse_rows(c(1, 2, 5))  %>%
        kable_styling(font_size = 7) %>%
        add_footnote(label = c(paste0("Estimates from Wernbom et al. ", "(ref:wernbom-citation)"),
                               paste0("Estimates from Ahtiainen et al. ", "(ref:ahtiainen-citation)")), 
                     notation = "alphabet")

```

```{r comb-fig-s1, warning=FALSE,message=FALSE, echo=FALSE, fig.cap="Differences in training induced relative changes in muscle mass and strength measures. Estimates are derived from ANCOVA models controling for baseline values and sex.", fig.scap = "Differences in training induced changes to muscle mass and strength measures between volume conditions in Study I", fig.height=2.6, fig.width=4.72, fig.align='center'}

library(ggplot2); library(ggtext)

# The figure and estimates are from ./R/study-1/muscle-strength-size.R
muscle_strength_fig <- readRDS("./data/derivedData/study1_muscle-strength-size/muscle-strength-fig.RDS")

muscle_strength_fig
```

The moderate-volume condition consistently showed favorable adaptations when compared to the low-volume condition in measures of muscle hypertrophy and strength gains (Figure \@ref(fig:comb-fig-s1)).
In an attempt to explain differences in training outcomes between volume-conditions, selected molecular markers with known influence on adaptations to resistance training were investigated for volume-dependency.
First, acute activation of signaling along the mechanosensitive mTORC1-pathway has previously been shown to correlate with training-induced muscle growth
[@RN866;<!-- Baar Esser correlation in rats  -->
@RN785;<!--Terzis et al., 2008-->
@RN788].<!-- Mitchell et al., 2013-->
A commonly used readout of mTORC1-signaling is the phosphorylation of S6-kinase (S6K1) at Thr\textsuperscript{389}/Thr\textsuperscript{412} which in turn precedes phosphorylation of ribosomal protein S6 (rpS6).
In the present study, exercise-induced S6K1\textsuperscript{Thr\textsuperscript{389}/Thr\textsuperscript{412}} phosphorylation was indeed shown to be volume dependent along with phosphorylation of rpS6\textsuperscript{Ser\textsuperscript{235/236}} and mTOR\textsuperscript{Ser\textsuperscript{2448}} (Figure \@ref(fig:mtor-fig)).
It is recognized that the phosphorylation status of rpS6 at Ser\textsuperscript{235/236} is not solely due to mTORC1 signalling as both mTORC1 and MEK/ERK converges here
[@RN2311].<!-- Roux S6 phopshorylation at ser 235 236 promoted by ras/erk -->
It is also recognized that phosphorylation of mTOR it self at Ser\textsuperscript{2448} primarily should be regarded as a indicative for negative feedback as this site is phosphorylated due to S6K1 activity
[@RN2309] (Figure \@ref(fig:mtor-fig) a).<!-- Figueriedo mTOR and ser2448 signalling review -->



```{r mtor-fig, warning=FALSE,message=FALSE, echo=FALSE, fig.cap="Measured phosphorylation sites in context (a) and differences between volume conditions in phosphorylation status of S6K1 at Thr\\textsuperscript{389} (p70) Thr\\textsuperscript{412} (p85), rpS6 at Ser\\textsuperscript{235/236} and mTOR at Ser\\textsuperscript{2448} induced by acute exercise expressed as fold-changes(b). Estimates are derived from ANCOVA models controling for baseline values. Non-adjusted estimates of differences in fold-changes between volume conditions are displayed with transparent points and wiskers. Values in b are point estimates with 95% confidence intervals.", fig.scap = "Differences between volume conditions in exercise induced phosphorylation of proteins related to mTORC1 signaling", fig.height=3.5, fig.width=4.72, fig.align='center'}

western_fig <- readRDS("./figures/results/study1-western-blot.RDS")

western_fig


```

These results are in agreement with previous studies showing exercise-volume dependency of mTORC1 related signaling  
[@RN791;<!-- Burd in single session single and multiple sets phosphorylation p70   -->
@RN784;<!-- Terzis volume affect activation -->
@RN1837].<!-- Ahtiainen 5 vs 10 sets p70 and s6 different -->



Given these limitations in using mTORC-signalling as markers of muscle hypertrophy, it is not surprising that previous studies are ambiguous in their associative approach between acute mTORC1-related phosphorylation and hypertrophy in humans. Some studies find a strong correlation

<!-- No correlation with muscle growth -->
[@RN834;<!-- Mitchell et al., 2012-->
@RN2171]<!-- Phillips et al., 2017-->.



 @RN1072;<!-- Goodman 2011 mTOR regulates mechanical induced hypertrophy partly through p70 Thr389 in rodents -->
 @RN785;<!-- Terzis in humans, p70 Thr 389 correlates with training increase in CSA -->

This seems somewhat counterintuitive, as this pathway is a known regulator of translation initiation and elongation, as well as of ribosomal biogenesis 
[@RN1632;<!-- Nader mTOR regulates ribo biogenthrough cell cycle regulators -->
@RN1810;<!-- von Walden 2012 mTOR regulates myotube... -->  
@RN2321;<!-- Chauvin S6K controls RiBi transcriptional program -->
@RN1754]<!-- West rapamycin sensitive and insensitive mechanisms of ribo biog -->


Indeed, in the present study we observed volume-dependence of mTOR phosphorylation at Ser2448, which could be a sign of negative feedback from mTORC1-based activation of S6K1 [@RN1949].<!-- Figueiredo 2017 review


These differences were in agreement to what could be expected based on published meta-analyses
[@RN2492;<!-- Ralston weekly sets volume, medium/higher volume better than low volume for strength gains  -->
@RN789;<!-- Krieger muscle hypertrophy -->
@RN793;<!-- Kriger muscle strength -->
@RN1767].<!-- Schoenfeld  -->
Furthermore, when a  combining data from more recent studies indicates that a higher training volume is generally associated with increased muscle hypertrophy and strength gains (Figure \@ref(fig:comb-fig-s1) and \@ref(fig:comb-fig-s1).   


```{r}

```





In Study II, training efficacy was assessed by comparing outcomes to a non-training control group. The training group displayed increases compared to the control group for both strength muscle thickness measures.


## Acute effects of diffrent training volume on determinants of muscle protein synthesis
<!--
Higher volume of acute resistance exercise led to increased phosphorylation of S6K1, ribosomal protein S6 and mTOR. These phosphorylation sites are indicative of activity along the mTORC1 pathway but mechanistic studies show that multiple sites exists for signal redundancy 
Inhibiting mTOR by Torin1 after electrically stimulated muscle contractions still led to S6K1 phsophorylation<!-- doi: 10.1016/j.cellsig.2013.05.005. -->

<!--
Regardless of upstream paralell signaling patterns the present result confirm volume dependency in a pathway controlling protein synthesis and ribosome biogenesis.

A limitation in the present study is that the limited panel of antibodies and time-points used to capture volume 
-->





