# Results and Discussion

```{r settings, include=FALSE}

library(nlme); library(kableExtra)


```


## Within-session training volume affects 

```{r comparing-avg-strength-csa, echo=FALSE, warning=FALSE, message=FALSE}


source("./R/study-1/mr-and-dxa-data.R")


participants <- read.csv("./data/study-1/body-composition/oneThreeSetLeg.csv", sep=";") %>%
        pivot_longer(names_to = "sets", values_to = "leg", cols = multiple:single) 


### Overall % increase in CSA (not corrected for baseline values)
over.all.csa <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(change = ((post/pre)-1)*100) %>%
  group_by(subject) %>%
  summarise(mean.csa = mean(change, na.rm = TRUE)) %>%
  ungroup() %>%
  summarise(m.csa = mean(mean.csa),
            s.csa = sd(mean.csa))

sets.csa <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(change = ((post/pre)-1)*100) %>%
  group_by(sets) %>%
  summarise(m.csa = mean(change),
            s.csa = sd(change))



### Baseline corrected % increase
csa.corrected.data <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(change = ((post/pre)-1)*100) %>% # Calculates % change
  group_by(sex) %>%
  mutate(pre = pre-mean(pre)) %>% # Mean centered baseline 
  # (no need for sex-diff in model)
  group_by(subject, sex) %>% # Keeps sex for test in model
  summarise(mean.csa = mean(change, na.rm = TRUE),
            mean.baseline = mean(pre)) # Average over both conditions 



## Anocova with baseline values as co-variate
csa.corr.m1 <- lm(mean.csa ~  mean.baseline,
                  data = csa.corrected.data)

csa.m1 <- lm(mean.csa ~  1,
                  data = csa.corrected.data)



csa.corr.confint  <- round(confint(csa.corr.m1)[1,],1)
csa.corr.estimate <- round(summary(csa.corr.m1)$coef[1,1],1)




csa <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  group_by(timepoint, sets, sex) %>%
  summarise(m=sprintf("%.1f", round(mean(CSA.avg),1)),
            s=sprintf("%.1f", round(sd(CSA.avg),1)))


csa.r <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(csa.change=((post/pre)-1)*100) %>%
  dplyr::select(subject, sex, sets, csa.change) %>%
  spread(sets, csa.change)

csa.r.model <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject,leg, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(csa.change=((post/pre)-1)*100) %>%
  dplyr::select(subject,leg, sex, sets, pre, csa.change) %>%
  group_by(sex) %>%
  mutate(pre.m = pre-mean(pre, na.rm = TRUE), 
         sets = factor(sets, levels = c("single", "multiple"))) 

#### Change to ancova ###


csa.m0 <-nlme::lme(csa.change ~ sets +  pre.m,
                random = list(subject = ~1),
                na.action = na.omit, 
                data = csa.r.model, method = "ML", 
                control=list(msMaxIter=120,
                             opt = "nloptwrap", msVerbose = FALSE))




csa.estimate <- sprintf("%.2f",round(coef(summary(csa.m0))[2,1], 2))
csa.confint <- intervals(csa.m0)$fixed[2, c(1,3)]

### t-test Does not change the conclusion 
# csa.ttest <- with(csa.r, t.test(multiple,single, paired=T))
# csa.estimate <- sprintf("%.2f",round(csa.ttest$estimate, 2))
# csa.confint <- csa.ttest$conf.int


csa.corr <- cor.test(csa.r$multiple, csa.r$single)


female.t <- with(csa.r[csa.r$sex=="female",], t.test(multiple,single, paired=T))
female.estimate <- sprintf("%.2f",round(female.t$estimate, 2))
female.confint <- female.t$conf.int


male.t <- with(csa.r[csa.r$sex=="male",], t.test(multiple,single, paired=T))
male.estimate <- sprintf("%.2f",round(male.t$estimate, 2))
male.confint <- male.t$conf.int

#### Strength data for overall stats



rm_str <- read_excel("./data/study-1/strength/strengthTests.xlsx") %>%
        filter(!(exercise %in% c("benchpress", "legcurl"))) %>%
        inner_join(participants) %>%
        filter(include == "incl") %>%

  dplyr::select(subject, sex, sets, exercise, timepoint, load)%>%
  mutate(load = as.numeric(load),
         sets = factor(sets, levels = c("single", "multiple"))) %>%
  spread(timepoint, load) %>%
  mutate(baseline = pmax(pre, session1)) %>%
  dplyr::select(subject, sex, sets, exercise, baseline, week2, week5, week9, post) %>%
  gather(week, load, baseline:post) %>%
  mutate(week = factor(week, levels = c("baseline", "week2", "week5", "week9", "week12", "post"))) %>%
  group_by(exercise) %>%
  mutate(str = load / max(load, na.rm = TRUE)) %>%
  dplyr::select(-load) %>%
  filter(week %in% c("baseline", "post")) 

# load cybex data 
load("./data/study-1/strength/cybex.rda")


cybex <- data.frame(cybex) %>%
        mutate(leg = if_else(leg == "Left", "L", "R")) %>%
        inner_join(participants) %>%
        filter(include == "incl", 
               timepoint %in% c("pre","session1", "post")) %>%

        filter(!(subject=="FP6" & timepoint == "post" & speed == 240))  # This removes one record, exclusion due to error in measurement, see protocol FP6


# Clean cybex data set and combine with 1RM 
total_strength <- cybex %>%
  ungroup() %>%
  dplyr::select(subject, timepoint, sets,speed, sex, torque.max) %>%
  spread(timepoint, torque.max) %>%
  rowwise()%>%
  mutate(baseline = max(pre, session1, na.rm=T)) %>%
  ungroup() %>%
  dplyr::select(subject, sex, sets, exercise = speed, baseline,  post) %>%
  gather(week, load, baseline:post) %>%
  group_by(exercise) %>%
  mutate(str = load/max(load, na.rm = TRUE)) %>%
  dplyr::select(-load) %>%
  filter(week %in% c("baseline", "post")) %>%
  rbind(rm_str) %>% # This includes all strength tests
  mutate(ex_fam = if_else(exercise %in% c("legext", "legpress"), 
                          "rm", if_else(exercise == 0, "isom", "isok"))) %>%
  group_by(subject, sex, sets, week, ex_fam) %>%
  summarise(str = mean(str, na.rm = TRUE)) %>%
  group_by(subject, sex, sets, week) %>%
  summarise(str = mean(str, na.rm = TRUE)) 


strength.stats <- total_strength %>%
  spread(week, str) %>%
  mutate(increase = 100 * ((post/baseline)-1)) %>%
  group_by(sex) %>%
  mutate(BASELINE = baseline - mean(baseline)) %>%
    group_by(sex, subject) %>%
    summarise(increase = mean(increase, na.rm = TRUE),
              BASELINE = mean(BASELINE, na.rm = TRUE)) 




## ANCOVA for the strength 1RM
strength.m1 <- lm(increase ~ sex + BASELINE,
                  data = strength.stats)

strength.corr.confint  <- round(confint(strength.m1)[1,],0)
strength.corr.estimate <- round(summary(strength.m1)$coef[1,1],0)



### Correlation in strength gains ###

str.corr.data <-  total_strength %>%
  spread(week, str) %>%
  mutate(increase = 100 * ((post/baseline)-1)) %>%
  dplyr::select(subject, sex, sets, increase) %>%
  spread(sets, increase) 

str.corr <- cor.test(str.corr.data$single,
                          str.corr.data$multiple)

#### Correlation strength and csa increases

# Average increases

csa.str.corr.data <- total_strength %>%
  spread(week, str) %>%
  mutate(str.rel.change = 100 * ((post/baseline)-1)) %>%
  dplyr::select(subject,  sets, sex, sets,baseline, str.rel.change) %>%

  inner_join(mr.results %>%
                mutate(condition=leg) %>%
                inner_join(participants) %>%
                filter(include=="incl") %>%
                dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
                spread(timepoint, CSA.avg) %>%
                mutate(csa.rel.change = 100 * ((post/pre)-1)) %>%
                dplyr::select(subject, sets, csa.rel.change)) %>%
        ungroup() 





# The code below could possibly be used to calculate the relative contribution of 
# csa increase to muscle strength increases


# m <- lme4::lmer(str.rel.change ~ csa.rel.change + sets + (1|subject), 
#                 data = csa.str.corr.data)
# 
# 
# 
# mx.null <- lme4::lmer(str.rel.change ~ 1 + (1|subject), data = csa.str.corr.data)
# mx.csa <- lme4::lmer(str.rel.change ~ csa.rel.change  + (1|subject), data = csa.str.corr.data)
# mx.sets <- lme4::lmer(str.rel.change ~ sets + (1|subject), data = csa.str.corr.data)
# 
# 
# 
# boot.fun <- function(x) {
#         data.frame(nlme::VarCorr(x))[2, 4]
# }
# 
# 
# null.boot <- lme4::bootMer(mx.null, boot.fun, nsim = 500, parallel = "snow", cl = 12)
# csa.boot <- lme4::bootMer(mx.csa, boot.fun, nsim = 500)
# sets.boot <- lme4::bootMer(mx.sets, boot.fun, nsim = 500)
# 
# 
# 
# qt.null <- quantile(null.boot$t, c(0.025, 0.975))
# qt.csa <- quantile(csa.boot$t, c(0.025, 0.975))
# qt.sets <- quantile(sets.boot$t, c(0.025, 0.975))
# 
# 
# results <- data.frame(variable = c("sets", "csa"),
#                            m = c(1-  data.frame(nlme::VarCorr(mx.sets))[2, 4] / # data.frame(nlme::VarCorr(mx.null))[2, 4],
#                                  1-  data.frame(nlme::VarCorr(mx.csa))[2, 4] / # data.frame(nlme::VarCorr(mx.null))[2, 4]),
#                            lwr = c(1- qt.sets[1] / qt.null[1],
#                                    1- qt.csa[1] / qt.null[1]),
#                            upr = c(1- qt.sets[2] / qt.null[2],
#                                    1- qt.csa[2] / qt.null[2]))
# 
# 
# 
# 
# 
# csa <- 1-  data.frame(nlme::VarCorr(mx.csa))[2, 4] / data.frame(nlme::VarCorr(mx.null))[2, 4]
# 
# sets <- 1-  data.frame(nlme::VarCorr(mx.sets))[2, 4] / data.frame(nlme::VarCorr(mx.null))[2, 4]
# 
# exp <- 1 - data.frame(nlme::VarCorr(mx.exp))[2, 4] / data.frame(nlme::VarCorr(mx.null))[2, 4]
# 
# 

# Average csa and strength gains

avg.str.csa <- csa.str.corr.data %>%
        group_by(subject) %>%
        summarise(str = mean(str.rel.change), 
                  csa = mean(csa.rel.change))


csa.str_corr <- cor.test(avg.str.csa$csa, avg.str.csa$str)

## Comparisons with other published studies on per session gains in csa and strength

### Per session ##
per.session <- read_excel("./data/study-1/trainingdata.xlsx", sheet = 1) %>%
  filter(!is.na(date)) %>%
  group_by(subject) %>%
  summarise(sessions = n()) %>%
        
inner_join(csa.str.corr.data) %>%

  mutate(str.change.day = str.rel.change/((sessions/31) * 12*7),
         csa.change.day = csa.rel.change/((sessions/31) * 12*7), 
         str.session = str.rel.change/sessions,
          csa.session = csa.rel.change/sessions ) %>%
   dplyr::select(-baseline) %>%
  ungroup() %>%
        pivot_longer(names_to = "variable", values_to = "gain", cols = str.rel.change:csa.session) 
       

sum.stats <- cbind(per.session %>%
        group_by(sets, variable, sex) %>%
        summarise(mean = mean(gain), 
                  sd = sd(gain)) %>%
        arrange(variable) %>%
        data.frame(), 
        data.frame(reference.mean = c(0.11, 0.11, 0.11, 0.11,
                              NA, NA, NA, NA, 
                              0.0825, 0.1375,0.0825, 0.1375, 
                              NA, NA, NA, NA, 
                               NA, NA, NA, NA,
                              0.6675, 0.475, 0.6675, 0.475), 
           reference.range = c("[0.04-0.26]a", 
                               "[0.04-0.26]a", 
                               "[0.04-0.26]a", 
                               "[0.04-0.26]a", 
                               "", "", "", "",
                               "(0.22)b", "(0.14)b",  "(0.22)b", "(0.14)b", 
                               "", "", "", "",
                              "", "", "", "",
                               "(0.35)b", "(0.22)b","(0.35)b", "(0.22)b" ))) %>%
        mutate(m.sd = if_else(mean > 6, paste0(sprintf("%.1f",mean), " (", sprintf("%.1f",sd), ")"), 
                              paste0(sprintf("%.2f", mean), " (", sprintf("%.2f",sd), ")")), 
               reference = if_else(is.na(reference.mean), "",
                                   paste0(sprintf("%.2f",reference.mean), " ", reference.range))) %>%
        filter(!(variable %in% c("str.change.day") )) %>%
                       dplyr::select(variable, sex, sets, m.sd, reference) %>%
        
        mutate(variable = factor(variable, levels = c("csa.rel.change", 
                                                      "csa.change.day", 
                                                      "csa.session", 
                                                      "str.rel.change", 
                                                      "str.session"), 
                                 labels = c("CSA %-change", 
                                            "CSA %-change per day", 
                                            "CSA %-change per session", 
                                            "Average strength %-change", 
                                            "Average strength %-change per session")), 
               sex = factor(sex, levels = c("female", 
                                            "male"), 
                            labels = c("Female", "Male")), 
               sets = factor(sets, levels = c("single", "multiple"), 
                             labels = c("LOW", "MOD"))) 

## Ahtiaianen on average  20 weeks with 2 sessions per week = 40 sessions

## Muscle hypertrophy
## In the younger cohorts, male average 5.5% (5.6) = 5.5/40 ~ 0.1375 per session
# in females 3.3 (8.6)% = 0.825
# 8.6/3.3 * 3.3/40
# 5.6/5.5 * 5.5/40
# Muscle strength in younger cohorts
#  male    19% (8.6)/40 sessions ~  0.5 % per session
#  female  26.7% (13.9)/40 ~ 
# 8.6/19 * (19/40)
# 13.9/26.7 * 26.7/40

# Wernbom increase per day in quadriceps CSA per day was 0.06%







```

In Study I, the average increases in muscle strength and mass in each volume condition corresponded to what could be expected based on previous research in young, healthy participants (Table \@ref(tab:csa-str-tab)) 
[@RN2007;<!-- Wernbom 2007 -->
@RN1741],<!-- Ahtiainen  -->
indicating the general efficacy of the training program.


(ref:wernbom-citation) @RN2007

(ref:ahtiainen-citation) @RN1741

```{r csa-str-tab, warning=FALSE, message=FALSE, echo=FALSE, results = "asis"}

sum.stats %>%
        arrange(variable, sex, sets) %>%
  dplyr::select(-variable) %>%
        kable(format = "latex", booktabs = TRUE, linesep = "",
              col.names = c("Sex", "Condition", "Mean (SD)", "Reference"), 
              caption = "Training induced changes in muscle CSA and average strength in Study I")  %>%
        collapse_rows(c(1, 4))  %>%
        kable_styling(font_size = 7) %>%
          
          pack_rows("CSA %-change", 1, 4) %>%
        pack_rows("CSA %-change per day", 5, 8) %>%
          pack_rows("CSA %-change per session", 9, 12) %>%
          pack_rows("Average strength %-change", 13, 16) %>%
          pack_rows("Average strength %-change per session", 17, 20) %>%
  
        add_footnote(label = c(paste0("Estimates from Wernbom et al. ", "(ref:wernbom-citation)"),
                               paste0("Estimates from Ahtiainen et al. ", "(ref:ahtiainen-citation)")), 
                     notation = "alphabet")

```

The multiple-sets (moderate-volume) condition consistently showed favorable adaptations when compared to the single-set (low-volume) condition in measures of muscle hypertrophy and strength gains (Figure \@ref(fig:comb-fig-s1)a).
This was in general agreement with previous meta-analyses indicating that an increased training volume will lead to more favorable outcomes, both in terms of muscle mass
[@RN1767;<!-- Schenfeld  2016--> 
@RN789]<!-- Krieger 2010 -->
and strength
[@RN2492;<!-- Ralston --> 
@RN793].<!-- Krieger 2009 -->


```{r meta-analyses-descriptive, warning=FALSE, message=FALSE, echo=FALSE}

summary_table_meta <- readRDS( "./data/derivedData/meta/naive_models.RDS")


summary_table_meta <- summary_table_meta %>%
  mutate(stat = paste0(sprintf("%.3f", median), 
                       " [",
                       sprintf("%.3f", cri.lwr), 
                       ", ",
                       sprintf("%.3f", cri.upr), 
                       "]"))


```


Since publication of the most recent meta-analyses [@RN1767;@RN2492]<!-- Schenfeld  2016 / Ralston 2107 --> 
additional studies have been published investigating the effect of training volume on muscle mass *and* muscle strength. Based on studies included in the meta-analysis perfomed by Schoenfeld *et al.* [@RN1767] additional studies were coded based on a systematic literature search among newly published studies (from January 2015 to June 2020). Compared to e.g. Schoenfeld *et al.* [@RN1767], an additional 10 studies were included in the analyses.
Subsequently, effect sizes were coded from 25 studies, both for muscle mass and strength and meta-regression models were used to investigate the effect of weekly number of sets on training-induced changes in these outcomes.
In basic models with the weekly number of sets as the only explanatory variable, both muscle mass and strength were shown to be robustly affected by an increased training volume.
For every additional weekly set, the effect size measure increased by
`r summary_table_meta[1, 5]` and `r summary_table_meta[2, 5]`
for muscle mass (Figure \@ref(fig:comb-fig-s1)b)
and strength (Figure \@ref(fig:comb-fig-s1)c),
respectively.


\pagebreak

```{r comb-fig-s1, warning=FALSE,message=FALSE, echo=FALSE, fig.cap="Differences in training induced relative changes in muscle mass and strength measures. Estimates are derived from ANCOVA models controling for baseline values.", fig.scap = "Differences in training induced changes to muscle mass and strength measures between volume conditions in Study I", fig.height=6, fig.width=4.72, fig.align='center'}

library(ggplot2); library(ggtext)

# The figure and estimates are from ./R/study-1/muscle-strength-size.R
muscle_strength_fig <- readRDS("./data/derivedData/study1_muscle-strength-size/muscle-strength-fig.RDS")

muscle_strength_fig
```

\pagebreak

Together these data underlines the dose-dependnace of within-session training volume on both muscle mass and strength. In an attempt to explain differences in training outcomes between volume-conditions, selected molecular markers with known influence on adaptations to resistance training were investigated for volume-dependency.
First, acute activation of signaling along the mechanosensitive mTORC1-pathway was measured in before and after the fifth training session in Study I.
A commonly used readout of mTORC1-signaling is the phosphorylation of S6-kinase (S6K1) at Thr\textsuperscript{389}/Thr\textsuperscript{412} which in turn precedes phosphorylation of ribosomal protein S6 (rpS6, see Figure \@ref(fig:mtor-fig)a).
In the present study, exercise-induced S6K1\textsuperscript{Thr\textsuperscript{389}/Thr\textsuperscript{412}} phosphorylation was indeed shown to be volume dependent along with phosphorylation of rpS6\textsuperscript{Ser\textsuperscript{235/236}} and mTOR\textsuperscript{Ser\textsuperscript{2448}} (Figure \@ref(fig:mtor-fig) b).
Together these observations indicates larger perturbations along the mTORC1 signaling axis which confirms previous observations showing exercise-volume dependency of mTORC1 related signaling in the acute phase (< 5 hours) after exercise with the low volume condition doing at least three sets activationg the muscle under study
[@RN2716;<!-- Kumar 2012 -->
@RN1837;<!-- Ahtiainen 5 vs 10 sets p70 and s6 different -->
@RN784].<!-- Terzis volume affect activation -->
It is recognized that phosphorylation of mTOR itself at Ser\textsuperscript{2448} primarily should be regarded as indicative for negative feedback as this site is phosphorylated due to S6K1 activity
[@RN2309] (Figure \@ref(fig:mtor-fig) a).<!-- Figueriedo mTOR and ser2448 signalling review -->
It is also recognized that the phosphorylation status of rpS6 at Ser\textsuperscript{235/236} is not solely due to mTORC1 signalling as both mTORC1 and extracellular signal-regulated kinases (ERK)-signalling converges here
[@RN2311].<!-- Roux S6 phopshorylation at ser 235 236 promoted by ras/erk -->

Interestingly, albeit with limited resolution, Terzis *et al.* and Ahtiainen *et al.* did not report any clear volume dependency in exercise-induced activation of ERK 1/2
[@RN785;<!--Terzis et al., 2008-->
@RN1837], 
suggesting that any volume-dependent regulation of this pathway outside the time-frame of these studies. However, it might also suggest that volume selectively modulates specific pathways in the acute phase after resistance exercise.
It should however be noted that the present, and previous studies are limited in their temopral resolution and different patterns over time in different volume conditions cannot be ruled out.

The importance of mTORC1 signalling in protein synthesis in the acute phase after resistance exercise is well established.
In humans, administration of rapamycin, a selective inhibitor of mTORC1, prior to resistance exercise leads to delayed or blunted activation of mTORC1 effectors such as
S6K1 and rpS6 as well as unchanged levels of phopshorylation of 4E-BP1 and eEF2
concomitantly with abolished exercise-induced increase in protein synthesis
[@RN780].<!-- Drummond rapamycin treatment and acute phase recovery (2 hours) -->
Volume dependent increase in mTORC1 signalling also coincide with larger protein synthesis rates
[@RN791;<!-- Burd in single session single and multiple sets phosphorylation p70   -->
@RN2716]<!-- Kumar volume affects protein synthesis -->
As such, volume dependence of mTORC1-related signalling suggests that higher within-session volume can be regarded as leading to an increased potential for protein synthesis in the acute-phase after exercise.

Previous studies has indirectly linked mTORC1-related signaling to muscle growth as acute pertubation along this pathway has been correlated with muscle growth
[@RN785;<!--Terzis et al., 2008-->
@RN788].<!-- Mitchell et al., 2013-->
However, this is not a consistent other findings in the literature 
[@RN834;<!-- Mitchell et al., 2012-->
@RN2171],<!-- Phillips et al., 2017-->
```{r p70-csa-correlation-data, warning=FALSE, message=FALSE, echo=FALSE}

### Check association with 

p70_csa <- readRDS("./data/derivedData/study1-western-blot/csa_p70.RDS")

### Calculate raw correlations
cor_est_p70 <- p70_csa %>%
  group_by(sets) %>%
   summarise(p70.fc.cor = cor.test(fc, change, method = "spearman")$estimate, 
            p85.fc.cor = cor.test(fc.p85, change, method = "spearman")$estimate, 
            p70.fc.cor.p = cor.test(fc, change, method = "spearman")$p.value, 
            p85.fc.cor.p = cor.test(fc.p85, change, method = "spearman")$p.value)  %>%
  data.frame()

p70.m <- lme(change ~ w2pre + scale(fc) + sets, 
             random = list(subject = ~ 1), 
             weights = varExp(),
             data = p70_csa)

p85.m <- lme(change ~  w2pre.p85 + scale(fc.p85)  + sets , 
             random = list(subject = ~ 1),
             weights = varExp(),
             data = p70_csa)

# plot(p70.m)
# plot(p85.m)


p70.mod <- paste0(sprintf("%.2f", data.frame(intervals(p70.m)$fixed)[3, 2]), 
                  " [", 
                  sprintf("%.2f", data.frame(intervals(p70.m)$fixed)[3, 1]),
                  ", ",
                  sprintf("%.2f", data.frame(intervals(p70.m)$fixed)[3, 3]), 
                  "]")

p85.mod <- paste0(sprintf("%.2f", data.frame(intervals(p85.m)$fixed)[3, 2]), 
                  " [", 
                  sprintf("%.2f", data.frame(intervals(p85.m)$fixed)[3, 1]),
                  ", ",
                  sprintf("%.2f", data.frame(intervals(p85.m)$fixed)[3, 3]), 
                  "]")

corr_est_p70 <- paste0(sprintf("%.3f",min(c(cor_est_p70[,2],cor_est_p70[,3]))), 
       " - ", 
       sprintf("%.3f",max(c(cor_est_p70[,2],cor_est_p70[,3])))) 

```
nor is it consistent with results in the present study as acute activation of S6K1 in response to the fifth session was not associated with training-induced changes in muscle mass.^[Raw Spearman's correlations between exercise induced change in S6K1 phosphorylation and change in muscle mass (measured with MRI) indicated weak (insignificant) associations ($\rho=$ `r corr_est_p70`) depedning on the isoform measured (p70 or p85) and number of sets. More elaborate modeling confirmed weak relationships between degree of phsophorylation and muscle mass change. When controlling for baseline values in degree of phsophorylation (as this influence the calculation of fold-change) and number of sets, standardized estimates for the effect of phosphorylation were (increase in %-muscle mass change for every 1SD change in fold-change phsophorylation, with [95% CI])  $\beta=$ `r p70.mod` for p70 S6K1-p70\textsuperscript{Thr389} and $\beta=$ `r p85.mod` for S6K1-p85\textsuperscript{Thr412}.].
Although mTORC1 signaling most certainly contributes to protein synthesis in the acute recovery phase after resistance exercise [@RN780],<!-- Drummond rapamycin treatment and acute phase recovery (2 hours) -->
its use as a predictor of training induced muscle hypertrophy is complicated by a number of reasons.
First, measuring its activity is not straight forward for biological reasons including pathway cross-talk and negative feedback as well as technical aspects
[@RN1949;<!-- Figueriedo, mTOR feedback -->
@RN2311;<!-- Cross talk with erk -->
@RN1942].<!--  Bass technical considerations in western blotting -->
Secondly, mTORC1 signaling is not a stable phenomenon as repeated bouts of resistance exercise leads to diminished activity along its pathway
[@RN1809;<!-- mTOR signal lost after three weeks -->
@RN1871].
Thirdly, mTORC1 readily integrate multiple signals but also convey these to multiple downstream processes
[@RN2320].
As such, mTORC1 activity represent an early, transient response to mechanical stress with its signaling contributing to accumulative responses.
One such downstream process is ribosomal biogenesis
[@RN1632;<!-- Nader mTOR regulates ribo biogenthrough cell cycle regulators -->
@RN1810;<!-- von Walden 2012 mTOR regulates myotube... -->  
@RN1828;<!-- von Walden 2016 -->
@RN2321],<!-- Chauvin S6K controls RiBi transcriptional program -->
ultimately leading to accumulation of ribosomal RNA, a response typically seen in connection with to resistance training



```{r mtor-fig, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Measured phosphorylation sites in context (a) and differences between volume conditions in phosphorylation status of S6K1 at Thr \\textsuperscript{389} (p70) Thr \\textsuperscript{412} (p85), rpS6 at Ser \\textsuperscript{235 236} and mTOR at Ser \\textsuperscript{2448} induced by acute exercise and expressed as fold-changes (b). Estimates are derived from ANCOVA models controling for baseline values. Values in b are point estimates with 95\\% confidence intervals.", fig.scap = "Differences between volume conditions in exercise induced phosphorylation of proteins related to mTORC1 signaling", fig.height=3.5, fig.width=4.72, fig.align='center'}

western_fig <- readRDS("./figures/results/study1-western-blot.RDS")
western_fig

```

Indeed, mTORC1 affects ribosomal biogenesis through multiple mechanisms including activation of transcriptional programs related to ribosomal biogenesis (mediated through S6K1)
[@RN2321],<!-- Chauvin S6K1/S6K2 null mice does not respond with ribosomal biogenesis transcription after feeding -->
selective translation of specific mRNA related to ribosomal biogenesis
[@RN2320]<!-- Goodman mRNA with "weak" complicated 5'-UTR translation is facilitated by mTORC1/eIF4E  -->
which in turn stimulates to increased ribosomal RNA transcription, mediated by cyclin D1 abundance and UBF availability
[@RN1632].<!-- Nader 2005 -->

```{r ribo-markers}

descriptive_ribo_markers <- readRDS("./data/derivedData/study1-ribo-biogenesis/descriptive_ribo_markers.RDS")

rrna <- descriptive_ribo_markers %>%
  filter(target != "totalrna") %>%
  summarise(min.w2pre = round(min(w2pre),0), 
            max.w2pre = round(max(w2pre),0), 
            min.w12 = round(min(w12),0), 
            max.w12 = round(max(w12),0)) 
```

```{r rrna-fig, warning=FALSE,message=FALSE, echo=FALSE, fig.cap="Differences between volume conditions in total RNA and ribosomal RNA-species (pre-rRNA 45S, rRNA 18S, rRNA 5.8S and rRNA 28S) measured at rest during the course of Study I (a). Acute changes in abundance of c-Myc mRNA and pre-rRNA 45S in response to acute exercise in Week 2 and differences between volume conditions (b). Errorbars represents 95\\% confidence intervals, transparent points and errorbars signifies that the confidence interval contain 1.", fig.scap = "Differences between volume conditions total RNA and ribosomal RNA", fig.height=2.5, fig.width=4.72, fig.align='center'}

ribo_markers_fig <- readRDS("./figures/results/study1-ribo-markers.RDS")

ribo_markers_fig

```

As ribosomal RNA is the most abundant constituent of muscle RNA it provides an estimate of muscle ribosomal abundance when expressed per unit tissue weight
[@RN2054;<!-- Zak 1967 in heart muscle 85% of RNA is ribosomal RNA -->
@RN2145].
From baseline to prior to the fifth training session (Week 2), total RNA per mg tissue increased by $\sim$ `r round(descriptive_ribo_markers[9,2], 0)` and `r round(descriptive_ribo_markers[10,2], 0)`% in the single-set and multiple-sets condition, respectively. Total RNA levels were still elevated above baseline after the intervention (Week 12, single-set $\sim$ `r round(descriptive_ribo_markers[9,3], 0)`; multiple-sets $\sim$ `r round(descriptive_ribo_markers[10,3], 0)`%). Similar patterns were seen in target analysis of ribosomal RNA using qPCR (single-set increase from baseline to Week 2 $\sim$ `r rrna[1,2]`-`r rrna[1,3]`% and Week 12 $\sim$ `r rrna[1,4]`-`r rrna[1,5]`%; multiple-sets $\sim$ `r rrna[2,2]`-`r rrna[2,3]`% and Week 12 $\sim$ `r rrna[2,4]`-`r rrna[2,5]`%). Comparing volume conditions revealed higher levels of total RNA and mature ribosomal RNA species in the multiple-sets condition at Week 2 (Figure \@ref(fig:rrna-fig)a). At Week 12 differences between volume conditions were in total RNA less clear and ribosomal RNA 28S showing higher levels in the single-set leg (Figure \@ref(fig:rrna-fig)a).

Together with indications of greater mTORC1 activation, analysis of c-Myc mRNA abundance in response to the fifth training session (Week 2) also showed volume-dependent regulation with exercise-induced increases being ~ 1.5-fold in response to the multiple-sets compared to the single-set condition (Figure \@ref(fig:rrna-fig)b). c-Myc represents a rapamycin-insesitive signalling pathway known to also stimulating ribosomal biogenesis 
[]<!-- West -->
[]<!-- cMyc driver of ribosomal biogenesis -->




```{r rna-csa-tab, warning=FALSE, message=FALSE, echo=FALSE, results = "asis"}

rna_csa_change_models <- readRDS("./data/derivedData/study1-rna-csa-cor/regression.RDS")

rna_csa_change_models %>%
  mutate(Estimate = round(Estimate, 2), 
         SE = round(SE, 2), 
         df = df, 
         t.value = round(t.value, 2), 
         ci = paste0("[", round(lower, 3), ", ", round(upper, 2), "]"), 
         beta = paste0(sprintf('%#.2f',scaled.est), " [", sprintf('%#.2f', scaled.lower), ", ", round(scaled.upper, 2), "]")) %>%
  dplyr::select(term, Estimate, SE, df, t.value, ci, beta) %>%
  mutate(term = factor(term, 
                       levels = c("(Intercept)", 
                                        "w0", 
                                        "w2pre", 
                                        "w12"), 
                       labels = c("(Intercept)", 
                                  "Week 0", 
                                  "Week 2", 
                                  "Week 12"))) %>%
  kable(format = "latex", 
        booktabs = TRUE, 
        escape = FALSE,
        caption = "Influence of RNA abundance on training-induced muscle growth measured with MRI", 
        col.names = c("", "Estimate", "SE", "df", "\\textit{t}-value", "95\\% CI", "Estimate [95\\% CI]")) %>%
  add_header_above(c(" " = 6, "Standardized coefficients" = 1), 
                   escape = FALSE) %>%
  pack_rows("RNA abundance (ng mg\\textsuperscript{-1})", 2, 4, escape = FALSE) %>%
  kable_styling(font_size = 8) %>%
  add_footnote("Standardized coefficients scaled by its SD")


```



```{r rrna-csa-fig, warning=FALSE,message=FALSE, echo=FALSE, fig.cap="Relationship between total RNA and training induced change in thigh cross sectional area (CSA) as measured by MRI. Dashed line represent 'naive' relationship, solid line represent adjusted reliationship as seen in \\@ref(table:rna-csa-tab).", fig.scap = "Relationship between total RNA and training induced muscle growth", fig.height=4.72/3, fig.width=4.72, fig.align='center'}

rna_csa_fig <- readRDS("./figures/results/study1-rna-csa.RDS")


rna_csa_fig

```

```{r}

p70_csa <- readRDS("./data/derivedData/study1-western-blot/csa_p70.RDS")

p70_csa %>%
  summarise(cor.r = cor.test(fc, change)$estimate, 
            cor.r.p = cor.test(fc, change)$p.value, 
            cor.p = cor.test(fc, change, method = "spearman")$estimate, 
            cor.p.p = cor.test(fc, change, method = "spearman")$p.value)


m1 <- lme(change ~ scale(fc) + sets, 
          random = list(subject = ~ 1),
          weights = varExp(),
          data = p70_csa)


summary(m1)



```





Additionally, results from genetically modified mice lacking the mTORC1 effector S6K1 confirms its specific importance for maintenance of skeletal muscle size as these mice 
[@RN2828].<!-- Ohanna deleting S6K1 leads to reduced cell size and myofiber csa similarly to rapamycin treatment  -->
Particularly S6K1 


mTORC controls protein synthesis trough multiple mechanisms
4E-BP1 and S6K1-rpS6

mTORC1 enhance translation of specific mRNA, including ribosoma biogenesis genes




[@RN2827]<!-- Pende S6K1/S6K2 null mice still have ser235/235 phosphorylation,  -->


[@RN2829]<!-- rpS6 is an effector of mTOR, determines cell size --> 






Given these limitations in using mTORC-signalling as markers of muscle hypertrophy, it is not surprising that previous studies are ambiguous in their associative approach between acute mTORC1-related phosphorylation and hypertrophy in humans. Some studies find a strong correlation

<!-- No correlation with muscle growth -->
[@RN834;<!-- Mitchell et al., 2012-->
@RN2171]<!-- Phillips et al., 2017-->.



 @RN1072;<!-- Goodman 2011 mTOR regulates mechanical induced hypertrophy partly through p70 Thr389 in rodents -->
 @RN785;<!-- Terzis in humans, p70 Thr 389 correlates with training increase in CSA -->

This seems somewhat counterintuitive, as this pathway is a known regulator of translation initiation and elongation, as well as of ribosomal biogenesis 
[@RN1632;<!-- Nader mTOR regulates ribo biogenthrough cell cycle regulators -->
@RN1810;<!-- von Walden 2012 mTOR regulates myotube... -->  
@RN2321;<!-- Chauvin S6K controls RiBi transcriptional program -->
@RN1754]<!-- West rapamycin sensitive and insensitive mechanisms of ribo biog -->


Indeed, in the present study we observed volume-dependence of mTOR phosphorylation at Ser2448, which could be a sign of negative feedback from mTORC1-based activation of S6K1 [@RN1949].<!-- Figueiredo 2017 review


These differences were in agreement to what could be expected based on published meta-analyses
[@RN2492;<!-- Ralston weekly sets volume, medium/higher volume better than low volume for strength gains  -->
@RN789;<!-- Krieger muscle hypertrophy -->
@RN793;<!-- Kriger muscle strength -->
@RN1767].<!-- Schoenfeld  -->
Furthermore, when a  combining data from more recent studies indicates that a higher training volume is generally associated with increased muscle hypertrophy and strength gains (Figure \@ref(fig:comb-fig-s1) and \@ref(fig:comb-fig-s1).   


```{r}

```





In Study II, training efficacy was assessed by comparing outcomes to a non-training control group. The training group displayed increases compared to the control group for both strength muscle thickness measures.


## Acute effects of diffrent training volume on determinants of muscle protein synthesis
<!--
Higher volume of acute resistance exercise led to increased phosphorylation of S6K1, ribosomal protein S6 and mTOR. These phosphorylation sites are indicative of activity along the mTORC1 pathway but mechanistic studies show that multiple sites exists for signal redundancy 
Inhibiting mTOR by Torin1 after electrically stimulated muscle contractions still led to S6K1 phsophorylation<!-- doi: 10.1016/j.cellsig.2013.05.005. -->

<!--
Regardless of upstream paralell signaling patterns the present result confirm volume dependency in a pathway controlling protein synthesis and ribosome biogenesis.

A limitation in the present study is that the limited panel of antibodies and time-points used to capture volume 
-->





