# Results and Discussion

```{r settings, include=FALSE}

library(nlme); library(kableExtra)


```


## Effects of different training volume on changes in muscle size and function

```{r comparing-avg-strength-csa, echo=FALSE, warning=FALSE, message=FALSE}


source("./R/study-1/mr-and-dxa-data.R")


participants <- read.csv("./data/study-1/body-composition/oneThreeSetLeg.csv", sep=";") %>%
        pivot_longer(names_to = "sets", values_to = "leg", cols = multiple:single) 


### Overall % increase in CSA (not corrected for baseline values)
over.all.csa <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(change = ((post/pre)-1)*100) %>%
  group_by(subject) %>%
  summarise(mean.csa = mean(change, na.rm = TRUE)) %>%
  ungroup() %>%
  summarise(m.csa = mean(mean.csa),
            s.csa = sd(mean.csa))

sets.csa <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(change = ((post/pre)-1)*100) %>%
  group_by(sets) %>%
  summarise(m.csa = mean(change),
            s.csa = sd(change))



### Baseline corrected % increase
csa.corrected.data <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(change = ((post/pre)-1)*100) %>% # Calculates % change
  group_by(sex) %>%
  mutate(pre = pre-mean(pre)) %>% # Mean centered baseline 
  # (no need for sex-diff in model)
  group_by(subject, sex) %>% # Keeps sex for test in model
  summarise(mean.csa = mean(change, na.rm = TRUE),
            mean.baseline = mean(pre)) # Average over both conditions 



## Anocova with baseline values as co-variate
csa.corr.m1 <- lm(mean.csa ~  mean.baseline,
                  data = csa.corrected.data)

csa.m1 <- lm(mean.csa ~  1,
                  data = csa.corrected.data)



csa.corr.confint  <- round(confint(csa.corr.m1)[1,],1)
csa.corr.estimate <- round(summary(csa.corr.m1)$coef[1,1],1)




csa <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  group_by(timepoint, sets, sex) %>%
  summarise(m=sprintf("%.1f", round(mean(CSA.avg),1)),
            s=sprintf("%.1f", round(sd(CSA.avg),1)))


csa.r <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(csa.change=((post/pre)-1)*100) %>%
  dplyr::select(subject, sex, sets, csa.change) %>%
  spread(sets, csa.change)

csa.r.model <- mr.results %>%

  inner_join(participants) %>%
  filter(include=="incl") %>%
  dplyr::select(subject,leg, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(csa.change=((post/pre)-1)*100) %>%
  dplyr::select(subject,leg, sex, sets, pre, csa.change) %>%
  group_by(sex) %>%
  mutate(pre.m = pre-mean(pre, na.rm = TRUE), 
         sets = factor(sets, levels = c("single", "multiple"))) 

#### Change to ancova ###


csa.m0 <-nlme::lme(csa.change ~ sets +  pre.m,
                random = list(subject = ~1),
                na.action = na.omit, 
                data = csa.r.model, method = "ML", 
                control=list(msMaxIter=120,
                             opt = "nloptwrap", msVerbose = FALSE))




csa.estimate <- sprintf("%.2f",round(coef(summary(csa.m0))[2,1], 2))
csa.confint <- intervals(csa.m0)$fixed[2, c(1,3)]

### t-test Does not change the conclusion 
# csa.ttest <- with(csa.r, t.test(multiple,single, paired=T))
# csa.estimate <- sprintf("%.2f",round(csa.ttest$estimate, 2))
# csa.confint <- csa.ttest$conf.int


csa.corr <- cor.test(csa.r$multiple, csa.r$single)


female.t <- with(csa.r[csa.r$sex=="female",], t.test(multiple,single, paired=T))
female.estimate <- sprintf("%.2f",round(female.t$estimate, 2))
female.confint <- female.t$conf.int


male.t <- with(csa.r[csa.r$sex=="male",], t.test(multiple,single, paired=T))
male.estimate <- sprintf("%.2f",round(male.t$estimate, 2))
male.confint <- male.t$conf.int

#### Strength data for overall stats



rm_str <- read_excel("./data/study-1/strength/strengthTests.xlsx") %>%
        filter(!(exercise %in% c("benchpress", "legcurl"))) %>%
        inner_join(participants) %>%
        filter(include == "incl") %>%

  dplyr::select(subject, sex, sets, exercise, timepoint, load)%>%
  mutate(load = as.numeric(load),
         sets = factor(sets, levels = c("single", "multiple"))) %>%
  spread(timepoint, load) %>%
  mutate(baseline = pmax(pre, session1)) %>%
  dplyr::select(subject, sex, sets, exercise, baseline, week2, week5, week9, post) %>%
  gather(week, load, baseline:post) %>%
  mutate(week = factor(week, levels = c("baseline", "week2", "week5", "week9", "week12", "post"))) %>%
  group_by(exercise) %>%
  mutate(str = load / max(load, na.rm = TRUE)) %>%
  dplyr::select(-load) %>%
  filter(week %in% c("baseline", "post")) 

# load cybex data 
load("./data/study-1/strength/cybex.rda")


cybex <- data.frame(cybex) %>%
        mutate(leg = if_else(leg == "Left", "L", "R")) %>%
        inner_join(participants) %>%
        filter(include == "incl", 
               timepoint %in% c("pre","session1", "post")) %>%

        filter(!(subject=="FP6" & timepoint == "post" & speed == 240))  # This removes one record, exclusion due to error in measurement, see protocol FP6


# Clean cybex data set and combine with 1RM 
total_strength <- cybex %>%
  ungroup() %>%
  dplyr::select(subject, timepoint, sets,speed, sex, torque.max) %>%
  spread(timepoint, torque.max) %>%
  rowwise()%>%
  mutate(baseline = max(pre, session1, na.rm=T)) %>%
  ungroup() %>%
  dplyr::select(subject, sex, sets, exercise = speed, baseline,  post) %>%
  gather(week, load, baseline:post) %>%
  group_by(exercise) %>%
  mutate(str = load/max(load, na.rm = TRUE)) %>%
  dplyr::select(-load) %>%
  filter(week %in% c("baseline", "post")) %>%
  rbind(rm_str) %>% # This includes all strength tests
  mutate(ex_fam = if_else(exercise %in% c("legext", "legpress"), 
                          "rm", if_else(exercise == 0, "isom", "isok"))) %>%
  group_by(subject, sex, sets, week, ex_fam) %>%
  summarise(str = mean(str, na.rm = TRUE)) %>%
  group_by(subject, sex, sets, week) %>%
  summarise(str = mean(str, na.rm = TRUE)) 


strength.stats <- total_strength %>%
  spread(week, str) %>%
  mutate(increase = 100 * ((post/baseline)-1)) %>%
  group_by(sex) %>%
  mutate(BASELINE = baseline - mean(baseline)) %>%
    group_by(sex, subject) %>%
    summarise(increase = mean(increase, na.rm = TRUE),
              BASELINE = mean(BASELINE, na.rm = TRUE)) 




## ANCOVA for the strength 1RM
strength.m1 <- lm(increase ~ sex + BASELINE,
                  data = strength.stats)

strength.corr.confint  <- round(confint(strength.m1)[1,],0)
strength.corr.estimate <- round(summary(strength.m1)$coef[1,1],0)



### Correlation in strength gains ###

str.corr.data <-  total_strength %>%
  spread(week, str) %>%
  mutate(increase = 100 * ((post/baseline)-1)) %>%
  dplyr::select(subject, sex, sets, increase) %>%
  spread(sets, increase) 

str.corr <- cor.test(str.corr.data$single,
                          str.corr.data$multiple)

#### Correlation strength and csa increases

# Average increases

csa.str.corr.data <- total_strength %>%
  spread(week, str) %>%
  mutate(str.rel.change = 100 * ((post/baseline)-1)) %>%
  dplyr::select(subject,  sets, sex, sets,baseline, str.rel.change) %>%

  inner_join(mr.results %>%
                mutate(condition=leg) %>%
                inner_join(participants) %>%
                filter(include=="incl") %>%
                dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
                spread(timepoint, CSA.avg) %>%
                mutate(csa.rel.change = 100 * ((post/pre)-1)) %>%
                dplyr::select(subject, sets, csa.rel.change)) %>%
        ungroup() 





# The code below could possibly be used to calculate the relative contribution of 
# csa increase to muscle strength increases


# m <- lme4::lmer(str.rel.change ~ csa.rel.change + sets + (1|subject), 
#                 data = csa.str.corr.data)
# 
# 
# 
# mx.null <- lme4::lmer(str.rel.change ~ 1 + (1|subject), data = csa.str.corr.data)
# mx.csa <- lme4::lmer(str.rel.change ~ csa.rel.change  + (1|subject), data = csa.str.corr.data)
# mx.sets <- lme4::lmer(str.rel.change ~ sets + (1|subject), data = csa.str.corr.data)
# 
# 
# 
# boot.fun <- function(x) {
#         data.frame(nlme::VarCorr(x))[2, 4]
# }
# 
# 
# null.boot <- lme4::bootMer(mx.null, boot.fun, nsim = 500, parallel = "snow", cl = 12)
# csa.boot <- lme4::bootMer(mx.csa, boot.fun, nsim = 500)
# sets.boot <- lme4::bootMer(mx.sets, boot.fun, nsim = 500)
# 
# 
# 
# qt.null <- quantile(null.boot$t, c(0.025, 0.975))
# qt.csa <- quantile(csa.boot$t, c(0.025, 0.975))
# qt.sets <- quantile(sets.boot$t, c(0.025, 0.975))
# 
# 
# results <- data.frame(variable = c("sets", "csa"),
#                            m = c(1-  data.frame(nlme::VarCorr(mx.sets))[2, 4] / # data.frame(nlme::VarCorr(mx.null))[2, 4],
#                                  1-  data.frame(nlme::VarCorr(mx.csa))[2, 4] / # data.frame(nlme::VarCorr(mx.null))[2, 4]),
#                            lwr = c(1- qt.sets[1] / qt.null[1],
#                                    1- qt.csa[1] / qt.null[1]),
#                            upr = c(1- qt.sets[2] / qt.null[2],
#                                    1- qt.csa[2] / qt.null[2]))
# 
# 
# 
# 
# 
# csa <- 1-  data.frame(nlme::VarCorr(mx.csa))[2, 4] / data.frame(nlme::VarCorr(mx.null))[2, 4]
# 
# sets <- 1-  data.frame(nlme::VarCorr(mx.sets))[2, 4] / data.frame(nlme::VarCorr(mx.null))[2, 4]
# 
# exp <- 1 - data.frame(nlme::VarCorr(mx.exp))[2, 4] / data.frame(nlme::VarCorr(mx.null))[2, 4]
# 
# 

# Average csa and strength gains

avg.str.csa <- csa.str.corr.data %>%
        group_by(subject) %>%
        summarise(str = mean(str.rel.change), 
                  csa = mean(csa.rel.change))


csa.str_corr <- cor.test(avg.str.csa$csa, avg.str.csa$str)

## Comparisons with other published studies on per session gains in csa and strength

### Per session ##
per.session <- read_excel("./data/study-1/trainingdata.xlsx", sheet = 1) %>%
  filter(!is.na(date)) %>%
  group_by(subject) %>%
  summarise(sessions = n()) %>%
        
inner_join(csa.str.corr.data) %>%

  mutate(str.change.day = str.rel.change/((sessions/31) * 12*7),
         csa.change.day = csa.rel.change/((sessions/31) * 12*7), 
         str.session = str.rel.change/sessions,
          csa.session = csa.rel.change/sessions ) %>%
   dplyr::select(-baseline) %>%
  ungroup() %>%
        pivot_longer(names_to = "variable", values_to = "gain", cols = str.rel.change:csa.session) 
       

sum.stats <- cbind(per.session %>%
        group_by(sets, variable, sex) %>%
        summarise(mean = mean(gain), 
                  sd = sd(gain)) %>%
        arrange(variable) %>%
        data.frame(), 
        data.frame(reference.mean = c(0.11, 0.11, 0.11, 0.11,
                              NA, NA, NA, NA, 
                              0.0825, 0.1375,0.0825, 0.1375, 
                              NA, NA, NA, NA, 
                               NA, NA, NA, NA,
                              0.6675, 0.475, 0.6675, 0.475), 
           reference.range = c("[0.04-0.26]a", 
                               "[0.04-0.26]a", 
                               "[0.04-0.26]a", 
                               "[0.04-0.26]a", 
                               "", "", "", "",
                               "(0.22)b", "(0.14)b",  "(0.22)b", "(0.14)b", 
                               "", "", "", "",
                              "", "", "", "",
                               "(0.35)b", "(0.22)b","(0.35)b", "(0.22)b" ))) %>%
        mutate(m.sd = if_else(mean > 6, paste0(sprintf("%.1f",mean), " (", sprintf("%.1f",sd), ")"), 
                              paste0(sprintf("%.2f", mean), " (", sprintf("%.2f",sd), ")")), 
               reference = if_else(is.na(reference.mean), "",
                                   paste0(sprintf("%.2f",reference.mean), " ", reference.range))) %>%
        filter(!(variable %in% c("str.change.day") )) %>%
                       dplyr::select(variable, sex, sets, m.sd, reference) %>%
        
        mutate(variable = factor(variable, levels = c("csa.rel.change", 
                                                      "csa.change.day", 
                                                      "csa.session", 
                                                      "str.rel.change", 
                                                      "str.session"), 
                                 labels = c("CSA %-change", 
                                            "CSA %-change per day", 
                                            "CSA %-change per session", 
                                            "Average strength %-change", 
                                            "Average strength %-change per session")), 
               sex = factor(sex, levels = c("female", 
                                            "male"), 
                            labels = c("Female", "Male")), 
               sets = factor(sets, levels = c("single", "multiple"), 
                             labels = c("LOW", "MOD"))) 

## Ahtiaianen on average  20 weeks with 2 sessions per week = 40 sessions

## Muscle hypertrophy
## In the younger cohorts, male average 5.5% (5.6) = 5.5/40 ~ 0.1375 per session
# in females 3.3 (8.6)% = 0.825
# 8.6/3.3 * 3.3/40
# 5.6/5.5 * 5.5/40
# Muscle strength in younger cohorts
#  male    19% (8.6)/40 sessions ~  0.5 % per session
#  female  26.7% (13.9)/40 ~ 
# 8.6/19 * (19/40)
# 13.9/26.7 * 26.7/40

# Wernbom increase per day in quadriceps CSA per day was 0.06%







```


In Study I, the average increases in muscle strength and mass in each volume condition corresponded to what could be expected based on previous research in young, healthy participants (Table \@ref(tab:csa-str-tab)) 
[@RN2007;<!-- Wernbom 2007 -->
@RN1741],<!-- Ahtiainen  -->
indicating the general efficacy of the training program.


(ref:wernbom-citation) @RN2007

(ref:ahtiainen-citation) @RN1741

```{r csa-str-tab, warning=FALSE, message=FALSE, echo=FALSE, results = "asis"}

sum.stats %>%
        arrange(variable, sex, sets) %>%
        kable(format = "latex", booktabs = TRUE, linesep = "",
              col.names = c("", "Sex", "Volume condition", "Mean (SD)", "Reference"), 
              caption = "Training induced changes in muscle CSA and average strength in Study I")  %>%
        collapse_rows(c(1, 2, 5))  %>%
        kable_styling(font_size = 7) %>%
        add_footnote(label = c(paste0("Estimates from Wernbom et al. ", "(ref:wernbom-citation)"),
                               paste0("Estimates from Ahtiainen et al. ", "(ref:ahtiainen-citation)")), 
                     notation = "alphabet")

```

```{r comb-fig-s1, warning=FALSE,message=FALSE, echo=FALSE, fig.cap="Differences in training induced relative changes in muscle mass and strength measures. Estimates are derived from ANCOVA models controling for baseline values.", fig.scap = "Differences in training induced changes to muscle mass and strength measures between volume conditions in Study I", fig.height=2.6, fig.width=4.72, fig.align='center'}

library(ggplot2); library(ggtext)

# The figure and estimates are from ./R/study-1/muscle-strength-size.R
muscle_strength_fig <- readRDS("./data/derivedData/study1_muscle-strength-size/muscle-strength-fig.RDS")

muscle_strength_fig
```

The multiple-sets (moderate-volume) condition consistently showed favorable adaptations when compared to the single-set (low-volume) condition in measures of muscle hypertrophy and strength gains (Figure \@ref(fig:comb-fig-s1)).
In an attempt to explain differences in training outcomes between volume-conditions, selected molecular markers with known influence on adaptations to resistance training were investigated for volume-dependency.
First, acute activation of signaling along the mechanosensitive mTORC1-pathway has previously been shown to correlate with training-induced muscle growth
[@RN866;<!-- Baar Esser correlation in rats  -->
@RN785;<!--Terzis et al., 2008-->
@RN788],<!-- Mitchell et al., 2013-->
presumably trough increased translation indicated by greater ribosomal association with mRNA
[@RN866].<!-- Baar Esser correlation in rats larger polysomal pool  -->
A commonly used readout of mTORC1-signaling is the phosphorylation of S6-kinase (S6K1) at Thr\textsuperscript{389}/Thr\textsuperscript{412} which in turn precedes phosphorylation of ribosomal protein S6 (rpS6, see Figure \@ref(fig:mtor-fig) a).
In the present study, exercise-induced S6K1\textsuperscript{Thr\textsuperscript{389}/Thr\textsuperscript{412}} phosphorylation was indeed shown to be volume dependent along with phosphorylation of rpS6\textsuperscript{Ser\textsuperscript{235/236}} and mTOR\textsuperscript{Ser\textsuperscript{2448}} (Figure \@ref(fig:mtor-fig) b).
It is recognized that phosphorylation of mTOR itself at Ser\textsuperscript{2448} primarily should be regarded as indicative for negative feedback as this site is phosphorylated due to S6K1 activity
[@RN2309] (Figure \@ref(fig:mtor-fig) a).<!-- Figueriedo mTOR and ser2448 signalling review -->
It is also recognized that the phosphorylation status of rpS6 at Ser\textsuperscript{235/236} is not solely due to mTORC1 signalling as both mTORC1 and extracellular signal-regulated kinases (ERK)-signalling converges here
[@RN2311].<!-- Roux S6 phopshorylation at ser 235 236 promoted by ras/erk -->
Interestingly, albeit circumstantial, Terzis et al., did not find any clear volume dependency in exercise induced activation of either ERK 1/2 nor p38
[@RN785]<!--Terzis et al., 2008-->
suggesting that mTORC1, partly through S6K1 drives volume-dependent rpS6\textsuperscript{Ser\textsuperscript{235/236}} phosphorylation.

The importance of mTORC1 signalling in protein synthesis in the acute phase after resistance exercise is well established.
In humans, administration of rapamycin, a selective inhibitor of mTORC1, prior to resistance exercise leads to delayed or blunted activation of mTORC1 effectors such as
S6K1\textsuperscript{Thr\textsuperscript{389}},
S6K1\textsuperscript{Thr\textsuperscript{421}/Ser\textsuperscript{424}}, 
rpS6\textsuperscript{Ser\textsuperscript{235/236}} and
rpS6\textsuperscript{Ser\textsuperscript{240/244}} as well as
unchanged levels of phopshorylation of 4E-BP1 and eEF2
concomitantly with abolished exercise-induced increase in protein synthesis
[@RN780].<!-- Drummond rapamycin treatment and acute phase recovery (2 hours) -->
As such, volume dependence of mTORC1-related signalling suggests that higher within-session volume is associated with increased potential for protein synthesis in the acute-phase after exercise.
However, an increased signalling through this pathway could also be interpreted as indicative for increased ribosomal biogenesis. 

```{r mtor-fig, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Measured phosphorylation sites in context (a) and differences between volume conditions in phosphorylation status of S6K1 at Thr \\textsuperscript{389} (p70) Thr \\textsuperscript{412} (p85), rpS6 at Ser \\textsuperscript{235 236} and mTOR at Ser \\textsuperscript{2448} induced by acute exercise and expressed as fold-changes (b). Estimates are derived from ANCOVA models controling for baseline values. Values in b are point estimates with 95\\% confidence intervals.", fig.scap = "Differences between volume conditions in exercise induced phosphorylation of proteins related to mTORC1 signaling", fig.height=3.5, fig.width=4.72, fig.align='center'}


western_fig <- readRDS("./figures/results/study1-western-blot.RDS")

western_fig


```

Indeed, mTORC1 affects ribosomal biogenesis through multiple mechanisms including activation of transcriptional programs related to ribosomal biogenesis (mediated through S6K1)
[@RN2321],<!-- Chauvin S6K1/S6K2 null mice does not respond with ribosomal biogenesis transcription after feeding -->
selective translation of specific mRNA related to ribosomal biogenesis
[@RN2320]<!-- Goodman mRNA with "weak" complicated 5'-UTR translation is facilitated by mTORC1/eIF4E  -->
which in turn stimulates to increased ribosomal RNA transcription, mediated by cyclin D1 abundance and UBF availability
[@RN1632].<!-- Nader 2005 -->

```{r ribo-markers}

descriptive_ribo_markers <- readRDS("./data/derivedData/study1-ribo-biogenesis/descriptive_ribo_markers.RDS")

rrna <- descriptive_ribo_markers %>%
  filter(target != "totalrna") %>%
  summarise(min.w2pre = round(min(w2pre),0), 
            max.w2pre = round(max(w2pre),0), 
            min.w12 = round(min(w12),0), 
            max.w12 = round(max(w12),0)) 

```


```{r rrna-fig, warning=FALSE,message=FALSE, echo=FALSE, fig.cap="Differences between volume conditions in total RNA and ribosomal RNA-species (pre-rRNA 45S, rRNA 18S, rRNA 5.8S and rRNA 28S) measured at rest during the course of Study I (a). Acute changes in abundance of c-Myc mRNA and pre-rRNA 45S in response to acute exercise in Week 2 and differences between volume conditions (b). Errorbars represents 95\\% confidence intervals, transparent points and errorbars signifies that the confidence interval contain 1.", fig.scap = "Differences between volume conditions total RNA and ribosomal RNA", fig.height=2.5, fig.width=4.72, fig.align='center'}

ribo_markers_fig <- readRDS("./figures/results/study1-ribo-markers.RDS")

ribo_markers_fig

```

As ribosomal RNA is the most abundant constituent of muscle RNA it provides an estimate of muscle ribosomal abundance when expressed per unit tissue weight
[@RN2054;<!-- Zak 1967 in heart muscle 85% of RNA is ribosomal RNA -->
@RN2145].
From baseline to prior to the fifth training session (Week 2), total RNA per mg tissue increased by $\sim$ `r round(descriptive_ribo_markers[9,2], 0)` and `r round(descriptive_ribo_markers[10,2], 0)`% in the single-set and multiple-sets condition, respectively. Total RNA levels were still elevated above baseline after the intervention (Week 12, single-set $\sim$ `r round(descriptive_ribo_markers[9,3], 0)`; multiple-sets $\sim$ `r round(descriptive_ribo_markers[10,3], 0)`%). Similar patterns were seen in target analysis of ribosomal RNA using qPCR (single-set increase from baseline to Week 2 $\sim$ `r rrna[1,2]`-`r rrna[1,3]`% and Week 12 $\sim$ `r rrna[1,4]`-`r rrna[1,5]`%; multiple-sets $\sim$ `r rrna[2,2]`-`r rrna[2,3]`% and Week 12 $\sim$ `r rrna[2,4]`-`r rrna[2,5]`%). Comparing volume conditions revealed higher levels of total RNA and mature ribosomal RNA species in the multiple-sets condition at Week 2 (Figure \@ref(fig:rrna-fig)a). At Week 12 differences between volume conditions were in total RNA less clear and ribosomal RNA 28S showing higher levels in the single-set leg (Figure \@ref(fig:rrna-fig)a).

```{r rna-csa-tab, warning=FALSE, message=FALSE, echo=FALSE, results = "asis"}

rna_csa_change_models <- readRDS("./data/derivedData/study1-rna-csa-cor/regression.RDS")

rna_csa_change_models %>%
  mutate(Estimate = round(Estimate, 2), 
         SE = round(SE, 2), 
         df = df, 
         t.value = round(t.value, 2), 
         ci = paste0("[", round(lower, 3), ", ", round(upper, 2), "]"), 
         beta = paste0(sprintf('%#.2f',scaled.est), " [", sprintf('%#.2f', scaled.lower), ", ", round(scaled.upper, 2), "]"), 
         beta = if_else(term %in% c("(Intercept)", "setssingle"), "", beta)) %>%
  dplyr::select(term, Estimate, SE, df, t.value, ci, beta) %>%
  mutate(term = factor(term, 
                       levels = c("(Intercept)", 
                                        "setssingle", 
                                        "w0", 
                                        "w2pre", 
                                        "w12"), 
                       labels = c("(Intercept)", 
                                  "Single-set", 
                                  "Week 0", 
                                  "Week 2", 
                                  "Week 12"))) %>%
  kable(format = "latex", booktabs = TRUE) %>%
  kable_styling()


```



```{r rrna-csa-fig, warning=FALSE,message=FALSE, echo=FALSE, fig.cap="Relationship between total RNA and training induced change in thigh cross sectional area (CSA) as measured by MRI. Dashed line represent 'naive' relationship, solid line represent adjusted reliationship as seen in \\@ref(table:rna-csa-tab).", fig.scap = "Relationship between total RNA and training induced muscle growth", fig.height=4.72/3, fig.width=4.72, fig.align='center'}

rna_csa_fig <- readRDS("./figures/study1-rna-csa.RDS")


rna_csa_fig

```



Additionally, results from genetically modified mice lacking the mTORC1 effector S6K1 confirms its specific importance for maintenance of skeletal muscle size as these mice 
[@RN2828].<!-- Ohanna deleting S6K1 leads to reduced cell size and myofiber csa similarly to rapamycin treatment  -->
Particularly S6K1 


mTORC controls protein synthesis trough multiple mechanisms
4E-BP1 and S6K1-rpS6

mTORC1 enhance translation of specific mRNA, including ribosoma biogenesis genes




[@RN2827]<!-- Pende S6K1/S6K2 null mice still have ser235/235 phosphorylation,  -->


[@RN2829]<!-- rpS6 is an effector of mTOR, determines cell size --> 


These results are in agreement with previous studies showing exercise-volume dependency of mTORC1 related signaling  
[@RN791;<!-- Burd in single session single and multiple sets phosphorylation p70   -->
@RN784;<!-- Terzis volume affect activation -->
@RN1837],<!-- Ahtiainen 5 vs 10 sets p70 and s6 different -->
coinciding with volume dependent protein synthesis
[@RN791].<!-- Burd in single session single and multiple sets phosphorylation p70   -->
Together



Given these limitations in using mTORC-signalling as markers of muscle hypertrophy, it is not surprising that previous studies are ambiguous in their associative approach between acute mTORC1-related phosphorylation and hypertrophy in humans. Some studies find a strong correlation

<!-- No correlation with muscle growth -->
[@RN834;<!-- Mitchell et al., 2012-->
@RN2171]<!-- Phillips et al., 2017-->.



 @RN1072;<!-- Goodman 2011 mTOR regulates mechanical induced hypertrophy partly through p70 Thr389 in rodents -->
 @RN785;<!-- Terzis in humans, p70 Thr 389 correlates with training increase in CSA -->

This seems somewhat counterintuitive, as this pathway is a known regulator of translation initiation and elongation, as well as of ribosomal biogenesis 
[@RN1632;<!-- Nader mTOR regulates ribo biogenthrough cell cycle regulators -->
@RN1810;<!-- von Walden 2012 mTOR regulates myotube... -->  
@RN2321;<!-- Chauvin S6K controls RiBi transcriptional program -->
@RN1754]<!-- West rapamycin sensitive and insensitive mechanisms of ribo biog -->


Indeed, in the present study we observed volume-dependence of mTOR phosphorylation at Ser2448, which could be a sign of negative feedback from mTORC1-based activation of S6K1 [@RN1949].<!-- Figueiredo 2017 review


These differences were in agreement to what could be expected based on published meta-analyses
[@RN2492;<!-- Ralston weekly sets volume, medium/higher volume better than low volume for strength gains  -->
@RN789;<!-- Krieger muscle hypertrophy -->
@RN793;<!-- Kriger muscle strength -->
@RN1767].<!-- Schoenfeld  -->
Furthermore, when a  combining data from more recent studies indicates that a higher training volume is generally associated with increased muscle hypertrophy and strength gains (Figure \@ref(fig:comb-fig-s1) and \@ref(fig:comb-fig-s1).   


```{r}

```





In Study II, training efficacy was assessed by comparing outcomes to a non-training control group. The training group displayed increases compared to the control group for both strength muscle thickness measures.


## Acute effects of diffrent training volume on determinants of muscle protein synthesis
<!--
Higher volume of acute resistance exercise led to increased phosphorylation of S6K1, ribosomal protein S6 and mTOR. These phosphorylation sites are indicative of activity along the mTORC1 pathway but mechanistic studies show that multiple sites exists for signal redundancy 
Inhibiting mTOR by Torin1 after electrically stimulated muscle contractions still led to S6K1 phsophorylation<!-- doi: 10.1016/j.cellsig.2013.05.005. -->

<!--
Regardless of upstream paralell signaling patterns the present result confirm volume dependency in a pathway controlling protein synthesis and ribosome biogenesis.

A limitation in the present study is that the limited panel of antibodies and time-points used to capture volume 
-->





