---
author: 'Daniel Hammarström'
date: 'May 20xx'
year: '2019'
isbn: 'Provided by the library'
place: 'Stockholm'
printedby: 'Printer service, Stockholm, 2019'
coverinfo: ' '
sernr: '999' # Add number in "Avhandlingsserie" 
title: 'Determinants of intra-individual variation in adaptability to resistance training of different volumes with special reference to skeletal muscle phenotypes'
subtitle: 'Subtitle of my thesis' # currently not in use ##
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  thesisdowngih::thesis_pdf: default
#  thesisdown::thesis_gitbook: default
#  thesisdown::thesis_word: default # The word and e-pub versions are not tested.
#  thesisdown::thesis_epub: default
# If you are creating a PDF you'll need to write your preliminary content here or
# use code similar to line 20 for the files.  If you are producing in a different
# format than PDF, you can delete or ignore lines 20-31 in this YAML header.
abstract: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-abstract.Rmd"), collapse = '\n  ')`
# If you'd rather include the preliminary content in files instead of inline
# like below, use a command like that for the abstract above.  Note that a tab is 
# needed on the line after the |.
listofpapers: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-listofpapers.Rmd"), collapse = '\n  ')`
dedication: |
  You can have a dedication here if you wish. 
# preface: | # If you wish to add a preface, remove comments on these lines
#   This is an example of a thesis setup to use the reed thesis document class
#   (for LaTeX) and the R bookdown package, in general.
defence: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-defence.Rmd"), collapse = '\n  ')`
bibliography: 
        - bib/thesis.bib
        - bib/packages.bib
# Download your specific bibliography database file and refer to it in the line above.
csl: csl/apa.csl
# Download your specific csl file and refer to it in the line above.
lot: true
lof: true
#space_between_paragraphs: true
# Delete the # at the beginning of the previous line if you'd like
# to have a blank new line between each paragraph
header-includes:
#- \usepackage[utf8]{inputenc}
#- \usepackage{erewhon}
#- \usepackage{lipsum}
#  \usepackage[utf8]{inputenc}
#  \usepackage[T1]{fontenc}
#  \usepackage{libertine}
#  \usepackage[svgnames]
#- \usepackage{GoudyIn}
#- \usepackage{xcolor}
---


<!--
Above is the YAML (YAML Ain't Markup Language) header that includes a lot of metadata used to produce the document.  Be careful with spacing in this header!

If you'd prefer to not include a Dedication, for example, simply delete lines 27 and 28 above or add a # before them to comment them out.  If you have other LaTeX packages you would like to include, delete the # before header-includes and list the packages after hyphens on new lines.

If you'd like to include a comment that won't be produced in your resulting file enclose it in a block like this.
-->


```{r include_packages, include = FALSE}
# This chunk ensures that the thesisdown package is
# installed and loaded. This thesisdown package includes
# the template files for the thesis.
library(devtools)
library(thesisdowngih)
```


```{r create_package_bib, include = FALSE}
knitr::write_bib(c("base", "qpcR"), "./bib/packages.bib") ### This creates a bibfile for R packages
```

<!-- \setcounter{DefaultLines}{3}-->


# Introduction
Proper skeletal muscle functioning is essential in everyday life by enabling movement and thus complex interaction with our environment. Nutrition and pharmacological agents can have substantial effects on skeletal muscle mass and function. However, resistance exercise  of sufficient volume, intensity and frequency is the most potent stimuli to promote morphological and functional changes in the human neuromuscular system. Exercise training can be modulated indefinitely by combining different variations of training variables and in addition, adaptation to exercise training is a phenomenon characterized by great inter-individual variability. The purpose of the present project is therefore to explore potential determinants of variation in adaptability to resistance-exercise modulated by selected exercise-training variables.    

Functional independence




<!-- Why study muscle functioning and health? -->
[@RN2512]<!-- Ran 2018 Low muscle strength was independently associated with all cause mortality -->

[@RN2184] <!-- Excess health care costs associated with muscle weakness was estimated to 1.3 billion punds in UK in population above 70 yrs   --> 

[@RN2513] <!-- Fukasawa 2016 81 hemodialysis patients, low thigh muscle mass associated with all cause and cv mortality-->

[@RN2514] <!-- Miyake 2019 Low muscle mass associated with ACM in type 2 diabetes, SMI = appendicular skeletal muscle mass kg / height m^2 -->

[@RN2515] <!-- Szulc 2010 715 men Loss of appendicular muscle mass predicts all cause mortality in elderly -->

[@RN2516]<!-- Abramowitz 2018 Higher ASMI was independently associated with lower mortality -->

[@RN2517]<!-- Janssen 2002 Low muscle mass associated with disabilities -->

<!-- Muscle mass and strength, multiple components contribute to muscle health -->



<!-- Pharmacological/nutritional/exercise interventions to maintain muscle mass --->

<!-- Problem with prescribing the right dose ---> 

<!-- Manipulation of exercise variables may overcome non-responders -->

<!-- Aim loosely defined -->


