---
author: 'Daniel Hammarström'
date: 'May 20xx'
year: '2019'
isbn: 'Provided by the library'
place: 'Stockholm'
printedby: 'Printer service, Stockholm, 2019'
coverinfo: ' '
sernr: '999' # Add number in "Avhandlingsserie" 
title: 'Determinants of intra-individual variation in adaptability to resistance training of different volumes with special reference to skeletal muscle phenotypes'
subtitle: 'Subtitle of my thesis' # currently not in use ##
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  thesisdowngih::thesis_pdf: default
#  thesisdown::thesis_gitbook: default
#  thesisdown::thesis_word: default # The word and e-pub versions are not tested.
#  thesisdown::thesis_epub: default
# If you are creating a PDF you'll need to write your preliminary content here or
# use code similar to line 20 for the files.  If you are producing in a different
# format than PDF, you can delete or ignore lines 20-31 in this YAML header.
abstract: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-abstract.Rmd"), collapse = '\n  ')`
# If you'd rather include the preliminary content in files instead of inline
# like below, use a command like that for the abstract above.  Note that a tab is 
# needed on the line after the |.
listofpapers: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-listofpapers.Rmd"), collapse = '\n  ')`
dedication: |
  You can have a dedication here if you wish. 
# preface: | # If you wish to add a preface, remove comments on these lines
#   This is an example of a thesis setup to use the reed thesis document class
#   (for LaTeX) and the R bookdown package, in general.
defence: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-defence.Rmd"), collapse = '\n  ')`
bibliography: 
        - bib/thesis.bib
        - bib/packages.bib
# Download your specific bibliography database file and refer to it in the line above.
csl: csl/vancouver.csl
# Download your specific csl file and refer to it in the line above.
lot: true
lof: true
#space_between_paragraphs: true
# Delete the # at the beginning of the previous line if you'd like
# to have a blank new line between each paragraph
header-includes:
#- \usepackage[utf8]{inputenc}
#- \usepackage{erewhon}
#- \usepackage{lipsum}
#  \usepackage[utf8]{inputenc}
#  \usepackage[T1]{fontenc}
#  \usepackage{libertine}
#  \usepackage[svgnames]
#- \usepackage{GoudyIn}
#- \usepackage{xcolor}
---


<!--
Above is the YAML (YAML Ain't Markup Language) header that includes a lot of metadata used to produce the document.  Be careful with spacing in this header!

If you'd prefer to not include a Dedication, for example, simply delete lines 27 and 28 above or add a # before them to comment them out.  If you have other LaTeX packages you would like to include, delete the # before header-includes and list the packages after hyphens on new lines.

If you'd like to include a comment that won't be produced in your resulting file enclose it in a block like this.
-->


```{r include_packages, include = FALSE}
# This chunk ensures that the thesisdown package is
# installed and loaded. This thesisdown package includes
# the template files for the thesis.
library(devtools)
library(thesisdowngih)
```


```{r create_package_bib, include = FALSE}
knitr::write_bib(c("base", "qpcR"), "./bib/packages.bib") ### This creates a bibfile for R packages
```

<!-- \setcounter{DefaultLines}{3}-->


# Introduction
Proper skeletal muscle functioning is essential in everyday life by enabling movement and thus independent interaction with the environment. In a lifespan perspective, measures of muscle mass and/or strength are inversely associated with mortality
[@RN2512;<!-- Ran 2018 Low muscle strength was independently associated with all cause mortality -->
@RN2513; <!-- Fukasawa 2016 81 hemodialysis patients, low thigh muscle mass associated with all cause and cv mortality-->
@RN2514; <!-- Miyake 2019 Low muscle mass associated with ACM in type 2 diabetes, SMI = appendicular skeletal muscle mass kg / height m^2 -->
@RN2376;<!-- Ruiz n ~ 8000, strength inversely related to all cause and cancer  mortality in men -->
@RN2515; <!-- Szulc 2010 715 men Loss of appendicular muscle mass predicts all cause mortality in elderly -->
@RN2516]<!-- Abramowitz 2018 Higher ASMI was independently associated with lower mortality -->
and disability
[@RN2517].<!-- Janssen 2002 Low muscle mass associated with disabilities -->
Besides adverse individual consequences of low muscle mass and strength and associated clinical conditions, muscle weakness accounts for increased health care costs in patient populations
[@RN2532;<!-- Sousa 2016 --> 
@RN2184].<!-- Pinedo-Villanuev Excess health care costs associated with muscle weakness was estimated to 1.3 billion punds in UK in population above 70 yrs   -->
Causal links between muscle mass and functioning on the one hand and adverse health outcomes on the other hand are diverse and related to e.g., age and primary  illness or injury
[@RN763].<!-- Wolfe 2006 amino acid availability in acute illness and injury, link between metabolic functioning of mm.-->
This highlights that interventions designed to increase muscle mass and strength are likely to prevent adverse health outcomes across the lifespan. A higher level of muscle mass and functional capacity would e.g., counteract the effects of muscle loss due to illness (cachexia), age (sarcopenia) or inactivity (athrophy).

Although a large degree of the observed variations in lean mass and strength are attributed to genetic components 
[@RN2526; <!-- Arden genetic factors explain leg extensor strength (60%), lean body mass (52%) and grip strength (36%) -->
@RN2527],<!-- Roth review, > 50% genetic component in muscle mass -->
environmental factors also contribute leaving a window of opportunity to increase muscle mass and functional capacity. Among factors possibly affecting muscle mass and functioning are nutrition and pharmacological agents. However, physical activity and specifically resistance training of sufficient volume, intensity and frequency provides a stimulus that promote morphological and functional changes to the human neuromuscular system without adverse side-effects. Irrespective of age, resistance training generally leads to increased muscle mass and strength
[@RN1741;<!-- Ahtiainen 2016 pooled analysis including controls   -->
@RN2534]<!-- Grgic 2020 > 75 of age increase strength and muscle mass relative to control and training is safe -->
and is considered safe when performed in a well organized manner
[@RN2534;<!-- Grgic safety in elderly -->
@RN2536].<!-- Faigenbaum 2010 resistance training in young safety -->

Resistance training can be modulated indefinitely by combining different variations of training variables
[@RN1;<!-- ACSM guidelines -->
@RN798].<!-- Bird 2005 acute programming variables -->
Well designed training prescriptions incorporates information on current state and goals of the trainee to maximize the potential of the training program
[@RN1;<!-- ACSM guidelines -->
@RN2538;<!-- Feigenbaum consider personal status in programming-->
@RN798].<!-- Bird 2005 acute programming variables -->
It has been argued that such considerations should include information such as age, training background,  

and in addition, adaptation to exercise training is a phenomenon characterized by great inter-individual variability. The purpose of the present project is therefore to explore potential determinants of variation in adaptability to resistance-exercise modulated by selected exercise-training variables.    

[@RN2063]<!-- Time constraints limits participation in physical activity -->


[@RN794]<!-- Carpinelli argues that single set programs are as effective as multiple set programs -->
[@RN2538]<!-- Feigenbaum, argues that more volume does not outweigh the time cost -->


<!-- Why study muscle functioning and health? -->


<!-- Muscle mass and strength, multiple components contribute to muscle health -->



<!-- Pharmacological/nutritional/exercise interventions to maintain muscle mass --->





<!-- Problem with prescribing the right dose ---> 

<!-- Manipulation of exercise variables may overcome non-responders -->

<!-- Aim loosely defined -->


