---
author: 'Daniel Hammarström'
date: 'May 20xx'
year: '2019'
isbn: 'Provided by the library'
place: 'Stockholm'
printedby: 'Printer service, Stockholm, 2019'
coverinfo: ' '
sernr: '999' # Add number in "Avhandlingsserie" 
title: 'Determinants of intra-individual variation in adaptability to resistance training of different volumes with special reference to skeletal muscle phenotypes'
subtitle: 'Subtitle of my thesis' # currently not in use ##
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  thesisdowngih::thesis_pdf: default
  latex_engine: xelatex
#  thesisdown::thesis_gitbook: default
#  thesisdown::thesis_word: default # The word and e-pub versions are not tested.
#  thesisdown::thesis_epub: default
# If you are creating a PDF you'll need to write your preliminary content here or
# use code similar to line 20 for the files.  If you are producing in a different
# format than PDF, you can delete or ignore lines 20-31 in this YAML header.
abstract: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-abstract.Rmd"), collapse = '\n  ')`
# If you'd rather include the preliminary content in files instead of inline
# like below, use a command like that for the abstract above.  Note that a tab is 
# needed on the line after the |.
listofpapers: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-listofpapers.Rmd"), collapse = '\n  ')`
dedication: |
  You can have a dedication here if you wish. 
# preface: | # If you wish to add a preface, remove comments on these lines
#   This is an example of a thesis setup to use the reed thesis document class
#   (for LaTeX) and the R bookdown package, in general.
defence: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-defence.Rmd"), collapse = '\n  ')`
bibliography: 
        - bib/thesis.bib
        - bib/complete-bib.bib
# Download your specific bibliography database file and refer to it in the line above.
csl: csl/vancouver.csl
# Download your specific csl file and refer to it in the line above.
lot: true
lof: true
#space_between_paragraphs: true
# Delete the # at the beginning of the previous line if you'd like
# to have a blank new line between each paragraph
header-includes:
#- \usepackage[utf8]{inputenc}
#- \usepackage{erewhon}
#- \usepackage{lipsum}
#  \usepackage[utf8]{inputenc}
#  \usepackage[T1]{fontenc}
#  \usepackage{libertine}
#  \usepackage[svgnames]
#- \usepackage{GoudyIn}
#- \usepackage{xcolor}
  \usepackage{lettrine}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{multirow}
  \usepackage{wrapfig}
  \usepackage{float}
  \usepackage{colortbl}
  \usepackage{pdflscape}
  \usepackage{tabu}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage[normalem]{ulem}
  \usepackage{makecell}
  \usepackage{siunitx}

---


<!--
Above is the YAML (YAML Ain't Markup Language) header that includes a lot of metadata used to produce the document.  Be careful with spacing in this header!

If you'd prefer to not include a Dedication, for example, simply delete lines 27 and 28 above or add a # before them to comment them out.  If you have other LaTeX packages you would like to include, delete the # before header-includes and list the packages after hyphens on new lines.

If you'd like to include a comment that won't be produced in your resulting file enclose it in a block like this.
-->


```{r include_packages, include = FALSE}
# This chunk ensures that the thesisdown package is
# installed and loaded. This thesisdown package includes
# the template files for the thesis.
library(devtools)
library(thesisdowngih)
```


```{r create_package_bib, include = FALSE}
knitr::write_bib(c("base", "qpcR"), "./bib/packages.bib") ### This creates a bibfile for R packages
```

\setcounter{DefaultLines}{3}


# Introduction
\lettrine{S}keletal muscle health is essential for physical independence. In a lifespan perspective, measures of muscle mass and/or strength are inversely associated with mortality
[@RN2512;<!-- Ran 2018 Low muscle strength was independently associated with all cause mortality -->
@RN2808; <!-- Muscular strength and mortality meta analysis higher levels of strength lower risk of death -->
@RN2513; <!-- Fukasawa 2016 81 hemodialysis patients, low thigh muscle mass associated with all cause and cv mortality-->
@RN2514; <!-- Miyake 2019 Low muscle mass associated with ACM in type 2 diabetes, SMI = appendicular skeletal muscle mass kg / height m^2 -->
@RN2809;<!-- Ruiz n ~ 8000, strength inversely related to all cause and cancer  mortality in men -->
@RN2515; <!-- Szulc 2010 715 men Loss of appendicular muscle mass predicts all cause mortality in elderly -->
@RN2516]<!-- Abramowitz 2018 Higher ASMI was independently associated with lower mortality -->
and disability
[@RN2517].<!-- Janssen 2002 Low muscle mass associated with disabilities -->
Besides adverse associations between of low muscle mass and strength and clinical conditions, muscle weakness also accounts for increased health care costs in patient populations
[@RN2532;<!-- Sousa 2016 --> 
@RN2184].<!-- Pinedo-Villanuev Excess health care costs associated with muscle weakness was estimated to 1.3 billion punds in UK in population above 70 yrs   -->
The intercept between muscle mass, muscle function and health status is interrelated with variables such as age and primary illness or injury
[@RN763].<!-- Wolfe 2006 amino acid availability in acute illness and injury, link between metabolic functioning of mm.-->
This highlights that interventions designed to increase muscle mass and strength are likely to prevent adverse health outcomes across the lifespan. A higher level of muscle mass and functional capacity would counteract the effects of muscle loss due to illness, age or inactivity.

Although a large degree of the observed variations in lean mass and strength are attributed to genetic components 
[@RN2526; <!-- Arden genetic factors explain leg extensor strength (60%), lean body mass (52%) and grip strength (36%) -->
@RN2527],<!-- Roth review, > 50% genetic component in muscle mass -->
environmental factors also contribute, leaving a window of opportunity to increase muscle mass and functional capacity. Among factors affecting muscle mass and functioning are nutrition and pharmacological agents. However, physical activity and specifically systematic resistance training of sufficient volume, intensity and frequency provides a stimulus that promote morphological and functional changes to the human neuromuscular system without adverse side-effects. Irrespective of age, resistance training generally leads to increased muscle mass and strength
[@RN1741;<!-- Ahtiainen 2016 pooled analysis including controls   -->
@RN2534]<!-- Grgic 2020 > 75 of age increase strength and muscle mass relative to control and training is safe -->
and is considered safe when performed in a well organized manner
[@RN2534;<!-- Grgic safety in elderly -->
@RN2536].<!-- Faigenbaum 2010 resistance training in young safety -->

Resistance training can be modulated indefinitely through combined variations of training variables such as frequency, intensity and volume 
[@RN1;<!-- ACSM guidelines -->
@RN798].<!-- Bird 2005 acute programming variables -->
Well designed training prescriptions should incorporate information about the current state and goals of the trainee to maximize the potential outcome of the training program
[@RN1;<!-- ACSM guidelines -->
@RN2538;<!-- Feigenbaum consider personal status in programming-->
@RN798].<!-- Bird 2005 acute programming variables -->
Training volume has received particular attention in the scientific community for many reasons. Evidence suggests that exercise volume affects selected molecular determinants of muscle hypertrophy in a dose-dependent manner
[@RN791; <!-- Burd  -->
@RN784; <!-- Terzis 2010 training volume p70 and s6 volume dependent -->
@RN1837].<!-- Ahtiainen 2015 10 X 10 > 5 X 10 in p70, S6 phospho --> 
Such effects are believed to facilitate long-term training effects as training programs with higher volume generally result in higher gains in muscle mass and strength with little evidence of differences between age groups or participants with different training backgrounds
[@RN793; <!-- Krieger 2009 meta on strength, 2-3 set more efficient than 1 set but not 4-6 sets -->
@RN789; <!-- Krieger 2010 meta analysis on training volume for hypertrophy, multiple sets more effective than signle set -->
@RN1767]. <!-- Shoenfeld 2016 weekly number of sets related to gains in muscle mass  -->  
A consequence of a more extensive training program is the increased time required to complete such a program. As time constraints has been reported as a limiting factor for engaging in physical activity
[@RN2063]<!-- Time constraints limits participation in physical activity -->
some merit can be given to arguments against guidlines suggesting higher volume in resistance training prescription 
[@RN794;<!-- Carpinelli argues that single set programs are as effective as multiple set programs -->
@RN2538].<!-- Feigenbaum, argues that more volume does not outweigh the time cost -->
From an individual perspective, training prescription that balances time-requirement with efficacy presumably increases the likelihood of participation in physical activity [@RN2063]<!-- Choi time-constrains and expectations of outcomes are related to participation -->.
From a more general perspective, increased knowledge about mechanisms governing responses to physical training could improve training prescription also for individuals and populations that experience attenuated benefit of resistance training
[@RN2547]. <!-- Pickering 2019, non-responders might not exists, changing training prescription could increase efficacy -->
<!-- Add "Individualized prescrioption is needed -->
The overreaching goal of the present thesis is to contribute to understanding individualized training loads. To this end, training volume was used to study the effects of variable training<!-- Change this? --> stimulus in within-participant models of exercise-training.   




