---
author: 'Daniel Hammarström'
date: 'May 20xx'
year: '2019'
isbn: 'Provided by the library'
place: 'Stockholm'
printedby: 'Printer service, Stockholm, 2019'
coverinfo: ' '
sernr: '999' # Add number in "Avhandlingsserie" 
title: 'Determinants of intra-individual variation in adaptability to resistance training of different volumes with special reference to skeletal muscle phenotypes'
subtitle: 'Subtitle of my thesis' # currently not in use ##
knit: "bookdown::render_book"
site: bookdown::bookdown_site
output: 
  thesisdowngih::thesis_pdf: default
#  thesisdown::thesis_gitbook: default
#  thesisdown::thesis_word: default # The word and e-pub versions are not tested.
#  thesisdown::thesis_epub: default
# If you are creating a PDF you'll need to write your preliminary content here or
# use code similar to line 20 for the files.  If you are producing in a different
# format than PDF, you can delete or ignore lines 20-31 in this YAML header.
abstract: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-abstract.Rmd"), collapse = '\n  ')`
# If you'd rather include the preliminary content in files instead of inline
# like below, use a command like that for the abstract above.  Note that a tab is 
# needed on the line after the |.
dedication: |
  You can have a dedication here if you wish. 
#preface: | # If you wish to add a preface, remove comments on these lines
#  This is an example of a thesis setup to use the reed thesis document class
#  (for LaTeX) and the R bookdown package, in general.
defence: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-defence.Rmd"), collapse = '\n  ')`
bibliography: 
        - bib/thesis.bib
        - bib/packages.bib
# Download your specific bibliography database file and refer to it in the line above.
csl: csl/apa.csl
# Download your specific csl file and refer to it in the line above.
lot: true
lof: true
#space_between_paragraphs: true
# Delete the # at the beginning of the previous line if you'd like
# to have a blank new line between each paragraph
header-includes:
#- \usepackage[utf8]{inputenc}
#- \usepackage{erewhon}
#- \usepackage{lipsum}
- \usepackage[utf8]{inputenc}
- \usepackage[T1]{fontenc}

- \usepackage{libertine}
- \usepackage{lettrine}

- \usepackage[svgnames]
#- \usepackage{GoudyIn}
#- \usepackage{xcolor}
---


<!--
Above is the YAML (YAML Ain't Markup Language) header that includes a lot of metadata used to produce the document.  Be careful with spacing in this header!

If you'd prefer to not include a Dedication, for example, simply delete lines 27 and 28 above or add a # before them to comment them out.  If you have other LaTeX packages you would like to include, delete the # before header-includes and list the packages after hyphens on new lines.

If you'd like to include a comment that won't be produced in your resulting file enclose it in a block like this.
-->


```{r include_packages, include = FALSE}
# This chunk ensures that the thesisdown package is
# installed and loaded. This thesisdown package includes
# the template files for the thesis.
library(devtools)
library(thesisdowngih)
```


```{r create_package_bib, include = FALSE}
knitr::write_bib(c("base", "qpcR"), "./bib/packages.bib") ### This creates a bibfile for R packages
```

\setcounter{DefaultLines}{3}


# Introduction

\lettrine[findent=3pt, nindent=0pt, ante=\rlap{\color{LightSalmon!50}\rule[-0.55\LettrineHeight]{\LettrineWidth}{0.65\LettrineHeight}} ]{P}{roper} 
 skeletal muscle functioning is essential in everyday life by enabling movement and thus complex interaction with our environment. Nutrition and pharmacological agents can have substantial effects on skeletal muscle mass and function. However, resistance exercise  of sufficient volume, intensity and frequency is the most potent stimuli to promote morphological and functional changes in the human neuromuscular system. Exercise training can be modulated indefinitely by combining different variations of training variables and in addition, adaptation to exercise training is a phenomenon characterized by great inter-individual variability. The purpose of the present project is therefore to explore potential determinants of variation in adaptability to resistance-exercise modulated by selected exercise-training variables.    

<!-- Why study muscle functioning and health? -->

[@RN2184] <!-- Excess health care costs associated with muscle weakness was estimated to 1.3 billion punds in UK in population above 70 yrs   --> 


<!-- Muscle mass and strength, multiple components contribute to muscle health -->



<!-- Pharmacological/nutritional/exercise interventions to maintain muscle mass --->

<!-- Problem with prescribing the right dose ---> 

<!-- Manipulation of exercise variables may overcome non-responders -->

<!-- Aim loosely defined -->



<!--chapter:end:index.Rmd-->


# Background
```{r Backgroundsettings, echo = FALSE, warning=FALSE, message=FALSE}
source("./R/libraries.R"); source("./R/themes.R")



```
## Exercise training variables affecting training outcomes

## Exercise volume

### Meta-analysis of exercise volume 

## Molecular determinants of training-induced muscle hypertrophy

### Protein synthesis



```{r Millward1973, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2.5, fig.width = 2.5, fig.cap="Data from Millward et al. 1973. Group A were fed a diet containing protein, group B were starved or fed a diet not containing protein."}

millward <- read_excel("./data/litrev.xlsx", sheet = "millward1973") 

millward %>%
        mutate(group = if_else(group == "A", "Protein fed", "Protein starved")) %>%
        ggplot(aes(RNA, proteinSynthesis, fill = group, color = group)) +
        geom_smooth(method = "lm", se = FALSE) + 
        geom_point(color = "black", shape = 21) +
        scale_x_continuous(name = TeX("$RNA\\;(\\mu g \\; mg^{-1} \\, protein)$"), 
                           breaks = c(0, 4, 8, 12, 16),
                           minor_breaks = c(2, 6, 10, 14),
                           limits = c(0, 16), 
                           expand = c(0, 0)) +
        scale_y_continuous(name = TeX("$Protein\\,synthesis\\,rate\\;(days^{-1})"), 
                           breaks = c(0, 0.05, 0.1, 0.15, 0.20, 0.25),
                           minor_breaks = c(0.02, 0.06, 0.10, 0.14, 0.18, 0.22),
                           limits = c(0, 0.25), 
                           expand = c(0, 0)) + 
        

        dissertation_theme() +
                theme(legend.position = c(0.2, 0.8), 
                      legend.title = element_blank()) 

        



```

### The mammalian target of rapamycin (mTOR) and translational efficiency
The mammalian target of rapamycin (mTOR) is a large serine-threonine protein kinase wich in complex with oyher regulatory proteins forms a signaling hub responsible for responses to environmental cues such as nutrients and mechanical stress.

mTOR has several phosphorylation sites 

Phosphorylation of Ser2448 is mediated by S6K1 to reduce mTOR activity in a negative feedback loop <!--[@RN1902]-->. 

Ser2448 is phosphorylated by S6K1, changes in nutrient avaliability modifies S6K1 and Ser2448, Ser2448 phosphorylation is abolished when S6K1 is depleted <!--[@RN1902]-->


When the C-terminal is deleted, mTOR gets constitutively active



### Ribsome biogenesis 

#### Transcription of ribsomal RNA (rRNA)

## Transcriptional activity related to muscle hypertrophy

### Methods for studying transcriptional regulation

<!--chapter:end:01-Background.Rmd-->

# Aims and hypotheses


<!--chapter:end:02-Aims.Rmd-->

# Methods


TO DO:

- For methods discussion, compare product length, efficiencies and ct values in relation to RQI-values. See Fleige 2006 for reference.



## Training protocols

A full body protocol was used in study I including 

<!--chapter:end:03-Methods.Rmd-->


# Results

<!--chapter:end:04-results.Rmd-->


# Discussion

<!--chapter:end:05-Discussion.Rmd-->

# Conclusion {-}

If we don't want Conclusion to have a chapter number next to it, we can add the `{-}` attribute.

**More info**

And here's some other random info: the first paragraph after a chapter title or section head _shouldn't be_ indented, because indents are to tell the reader that you're starting a new paragraph. Since that's obvious after a chapter or section title, proper typesetting doesn't add an indent there.


<!--chapter:end:06-conclusion.Rmd-->

<!--
The bib chunk below must go last in this document according to how R Markdown renders.  More info is at http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
-->

\backmatter

<!-- 
If you'd like to change the name of the bibliography to something else,
delete "References" and replace it.
-->

# References {-}
<!--
This manually sets the header for this unnumbered chapter.
-->
\markboth{References}{References}
<!--
To remove the indentation of the first entry.
-->
\noindent

<!--
To create a hanging indent and spacing between entries.  These three lines may need to be removed for styles that don't require the hanging indent.
-->

\setlength{\parindent}{-0.20in}
\setlength{\leftskip}{0.20in}
\setlength{\parskip}{8pt}



<!--chapter:end:99-references.Rmd-->

